---
title: 'Annotation of BAL cells'
author:
- name: Anson Wong
date: "`r Sys.Date()`"
description: null
output: html_document
editor_options:
  chunk_output_type: inline
---

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
xaringanExtra::use_panelset()

```

# 1 Load packages

```{r}
suppressPackageStartupMessages({
library(Seurat)
library(here)
library(qs)
library(forcats)
library(pals)
library(ggplot2)
library(speckle)
library(scCustomize)
library(patchwork)
library(dplyr)
library(Cepo)
library(rstatix)
library(ggpubr)
})
set.seed(1990)

```

# 2 Reference-based annotation, integration, clustering

After filtering low-quality cells and uninformative genes, reference-based annotation (Human Lung Cell Atlas v2) was performed on the clean dataset using RunAzimuth(). Due to internal in-house server compatibility issues, R v4.2.2 instead of R v4.1.2 was used for running this step. For details, refer to code/annotation_HLCAv2.R

The dataset was normalized across batches using SCTransform v2 regularization and subjected to PCA. Harmony was used to adjust PCs by capture and sample ID for 500 iteration. The first 50 PCs were used to compute the shared nearest neighbor graph for Leiden clustering with a resolution of 3. Cells were manually annotated with reference to Azimuth annotation results and the RNA expression of known marker genes provided in previous studies (Madissoon et al. 2022, Sikkema et al. 2023). For details, refer to code/integration_clustering.R

Manual annoatation was performed on all BAL cells to define major cell type.

# 3 Manual annotation - Combined

## 3.1 Load clustered object

```{r}
seu <- qread(here("data",
                  "SCEs",
                  "clustering",
                  "bal.annotated.merged.integrated.clustered.SEU.qs"), nthreads=16)

seu$predicted.ann_level_3 <- factor(seu$predicted.ann_level_3)
seu$predicted.ann_finest_level <- factor(seu$predicted.ann_finest_level)
seu$sampleID <- factor(seu$sampleID)
seu$Capture <- factor(seu$Capture)
seu$Condition <- factor(seu$Condition)
seu$SCT_snn_res.3 <- factor(seu$SCT_snn_res.3,
                            levels=as.character(seq(1,64)))

seu@reductions$umap@cell.embeddings[,1] <- -seu@reductions$umap@cell.embeddings[,1]
seu@reductions$umap@cell.embeddings[,2] <- -seu@reductions$umap@cell.embeddings[,2]

```

## 3.2 Visualize UMAP embeddings

```{r}
# prepare colours
colours <- c(brewer.paired(12), brewer.pastel1(9), rev(brewer.accent(8)), rev(brewer.dark2(8)), 
             brewer.set1(9), brewer.pastel2(8), brewer.spectral(11), 
             brewer.brbg(11)[1:5],rev(brewer.set3(12)),brewer.brbg(11)[6:11],
             brewer.piyg(11)[1:5],brewer.piyg(11)[7:11], brewer.piyg(11)[1:5],brewer.piyg(11)[7:11])


p1 <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        label = FALSE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2.5, 
        cols=c(brewer.paired(12),brewer.pastel1(9),rev(brewer.dark2(8)),brewer.accent(8)),
        # cols=c("#054a91", "#fb6f92", "#81a4cd", "#b79ced","7d451b","#a3b18a", 
        #        "#0e9594", "#f77f00", "#b23a48"),
        split.by = "Condition",
        group.by = "predicted.ann_finest_level") +
  #scale_x_continuous(breaks = c(-5, 0, 5)) +
  #scale_color_manual(values=Annotation.colours) +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))

p2 <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        #label.box = TRUE,
        label = FALSE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2.5, 
        cols=colours,
        # cols=c("#054a91", "#fb6f92", "#81a4cd", "#b79ced","7d451b","#a3b18a", 
        #        "#0e9594", "#f77f00", "#b23a48"),
        split.by = "Condition",
        group.by = "SCT_snn_res.3") + #NoLegend() +
  #scale_x_continuous(breaks = c(-5, 0, 5)) +
  #scale_color_manual(values=Annotation.colours) +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))

p3 <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        label = FALSE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2.5, 
        cols=c(brewer.paired(12),brewer.pastel1(9),rev(brewer.dark2(8)),brewer.accent(8)),
        # cols=c("#054a91", "#fb6f92", "#81a4cd", "#b79ced","7d451b","#a3b18a", 
        #        "#0e9594", "#f77f00", "#b23a48"),
        # split.by = "Condition",
        group.by = "predicted.ann_finest_level") +
  #scale_x_continuous(breaks = c(-5, 0, 5)) +
  #scale_color_manual(values=Annotation.colours) +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))

p4 <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        label.box = TRUE,
        label = TRUE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2, 
        cols=colours,
        # cols=c("#054a91", "#fb6f92", "#81a4cd", "#b79ced","7d451b","#a3b18a", 
        #        "#0e9594", "#f77f00", "#b23a48"),
        # split.by = "Condition",
        group.by = "SCT_snn_res.3") + NoLegend() +
  #scale_x_continuous(breaks = c(-5, 0, 5)) +
  #scale_color_manual(values=Annotation.colours) +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))
p1
p2
p3
p4

```

## 3.3 Proportion of cells per cluster

```{r}

labels <- c("predicted.ann_finest_level","SCT_snn_res.3")
p <- vector("list",length(labels))

for(label in labels){
  seu@meta.data %>%
    ggplot(aes(x = !!sym(label), fill = !!sym(label))) +
    geom_bar() +
    geom_text(aes(label = ..count..), stat = "count",
              angle=90, colour = "black", size = 2) +
    scale_y_log10() +
    theme(axis.text.x = element_blank(),
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank()) +
    NoLegend() + scale_fill_manual(values = rev(colours)) +
    labs(y = "No. Cells (log scale)") -> p1
  
  seu@meta.data %>%
    dplyr::select(!!sym(label), Condition) %>%
    group_by(!!sym(label), Condition) %>%
    summarise(num = n()) %>%
    mutate(prop = num / sum(num)) %>%
  ggplot(aes(x = !!sym(label), y = prop * 100, 
             fill = Condition)) + 
    geom_bar(stat = "identity") +
    theme(axis.text.x = element_text(angle = 45, size=6,
                                     vjust = 1, 
                                     hjust = 1)) +
    labs(y = "% Cells", fill = "Condition") -> p2
  
  p1 / p2 -> p[[label]]
}
p

```

## 3.4 Visualize feature expression

```{r}
# # proliferating
# FeaturePlot(seu, c("MKI67","PCLAF"), ncol = 2, raster=TRUE, cols=inferno(256))
# #FeaturePlot(seu, c("G2M.Score","S.Score"), ncol = 2, raster=TRUE, cols=inferno(256))
# 
# # Secretory
# FeaturePlot(seu, c("SCGB3A2","SCGB1A1","CCNO","VSIG2","CEACAM5","CYP2F1","TSPAN8","MUC5AC"
#                    ), ncol = 3, raster=FALSE, cols=inferno(256))
# 
# # Multiciliated
# FeaturePlot(seu, c("CCDC78","FOXJ1","PIFO","DCDC2B","DTHD1","GSTA1"
#                    ), ncol = 2, raster=FALSE, cols=inferno(256))
# 
# # pDC
# FeaturePlot(seu, c("JCHAIN","CLEC4C","PLD4","LIRA4","CXCR3","IL3RA"
#                    ), ncol = 2, raster=FALSE, cols=inferno(256))
# 
# # TRM
# FeaturePlot(seu, c("MARCO","APOE","C1QA","CD163"), ncol = 2, raster=TRUE, cols=inferno(256))
# 
# # monocyte-derived macrophage
# FeaturePlot(seu,c("MARCO","C1QA","VCAN","CCL2","FCN1","EREG","MS4A6A","CD14","CD300E"),
#             ncol=3,raster=TRUE, cols=inferno(256))
# 
# # cDC
# FeaturePlot(seu, c("CLEC9A","XCR1","IDO1","CD1C","CLEC10A","FCGR2B","FCER1A","MEF2C",
#                    "CCR7","BIRC3","RASSF4","TRAF1","EBI3"
#                    ), ncol = 4, raster=FALSE, cols=inferno(256))
# 
# # mast
# FeaturePlot(seu, c("BIRC3","NFKB1","NAMPT","BATF","TNFAIP3","HDC","RGS13",
#                    "TPSD1","SLC18A2","MS4A2","HPGDS","CPA3","TPSB2","TPSAB1"
#                    ), ncol = 5, raster=FALSE, cols=inferno(256))
# 
# # RecM intravascular
# FeaturePlot(seu, c("LILRB5","F13A1","STAB1","RNASE1","MAF","FOLR2",
#                    "MARCO","C1QA","CD14","SLC40A1","SLC48A1"), ncol = 4, raster=TRUE, cols=inferno(256))
# 
# # interstitial
# FeaturePlot(seu, c("CXCL10","CXCL9","CXCL11","GBP1","ISG15","MARCO",
#                    "GBP2","PSTPIP2","GCH1","APOL3","IL1B","MARCKS"), ncol = 3, raster=TRUE, cols=inferno(256))
# 
# # CD14+ monocytes
# FeaturePlot(seu, c("CD300E","FCN1","S100A12","EREG","CD14","VCAN",
#                    "MARCO","LST1","LYZ","CCR2",
#                    "IL1B","C1QA"), ncol = 4, raster=TRUE, cols=inferno(256))
# 
# # CD16+ monocytes
# FeaturePlot(seu, c("FCGR3A","LILRB2","LILRB1","MTSS1","FAM110A","LRRC25",
#                    "CDKN1C","ABI3","C1QA"), ncol = 3, raster=TRUE, cols=inferno(256))
# 
# # T cell
# FeaturePlot(seu, c("CD96","MARCO","CCL5","CD3D"), ncol = 2, raster=TRUE, cols=inferno(256))
# 
# # NK & NKT
# FeaturePlot(seu, c("CCL4","CCL5","PFN1","GZMA","GZMB",
#                    "KLF2","GNLY","FGFBP2","NKG7","CD3D","CD3E","CD3G","ITGAM","ITGA1",
#                    "CD8A","CD8B"),
#                    ncol=4, raster=FALSE, cols=inferno(256))

```

## 3.5 Manual annotation

```{r}
# Use Azimuth result to annotate mast cells
draft1 <- factor(
      case_when(
        seu$predicted.ann_finest_level == "Mast cells" ~ "Mast"))

seu <- AddMetaData(seu, draft1, col.name = "draft1")

# integrated 50PCs
draft2 <- setNames(c("AM","AM","AM","T","AM", # 1-5
                      "AM","AM","AM","T","AM", # 6-10
                      "AM","AM","AM","AM","AM", # 11-15
                      "RecM","RecM","RecM","AM","AM", # 16-20
                      "AM","B","cDC","AM","AM", # 21-25
                      "AM","AM","RecM","AM","AM", # 26-30
                      "AM","AM","AM","Prolif M","Epithelial", # 31-35
                      "AM","Prolif M","AM","AM","AM", # 36-40
                      "AM","AM","AM","AM","T", # 41-45
                      "T","pDC","NK","RecM","RecM", # 46-50
                      "RecM","RecM","DC migratory","AM","RecM", # 51-55
                      "RecM","Epithelial","Prolif M","AM","AM", # 56-60
                      "T","B","cDC","T" # 61-64
                       ), levels(seu$SCT_snn_res.3))

draft2 <- factor(setNames(draft2[seu$SCT_snn_res.3],colnames(seu)),
                     levels=c("AM",
                              "RecM",
                              "Prolif M",
                              "cDC",
                              "pDC",
                              "DC migratory",
                              "T",
                              "NK",
                              "B",
                              "Epithelial"))

seu <- AddMetaData(seu, draft2, col.name = "draft2")

# coalesce two draft columns
seu@meta.data %>%
  mutate(celltype=coalesce(draft1,draft2)) -> seu@meta.data

# reorder levels
seu$celltype <- factor(seu$celltype,
                       levels=c("AM","RecM","Prolif M","cDC","pDC","DC migratory","Mast",
                                "T", "NK","B","Epithelial"))

# remove draft columns
seu$draft1 <- NULL
seu$draft2 <- NULL

Annotation.colours <- setNames(c("#054a91", # AM
                                 "#dda15e", # RecM
                                 "#81a4cd", # Prolif M
                                 "#a3b18a", # cDC
                                 "#81b29a", # pDC
                                 "#3a5a40", # DC migratory
                                 "#000000", # Mast
                                 "#c3a995", # T
                                 "#7f5539", # NK
                                 "#5a189a", # B
                                 "#b23a48" # Epithelial
                                 ), # Secretory
               levels(seu$celltype))
```

## 3.6 Check batch effect

### 3.6.1 UMAP

```{r fig.width=10, fig.height=3}
# rename batch ID
seu$batchID <- factor(seu$batchID,
                       levels=c("A","B","C","D"))

batchID2 <- setNames(c("batch 3","batch 4","batch 2", "batch 1"),
                     levels(seu$batchID))

batchID2 <- setNames(factor(batchID2[seu$batchID],levels=paste0("batch ",1:4)),
                     colnames(seu))

seu <- AddMetaData(seu, batchID2, col.name = "batchID2")

batch.colours <- setNames(c("#e07a5f", 
                           "#3d405b", 
                           "#81b29a", 
                           "#f2cc8f"),
                  levels(seu$batchID2))

# # umap
# seu2 <- seu
# seu <- RunUMAP(seu, reduction="pca", dims=1:50)
# seu@reductions$umap@cell.embeddings[,2] <- -seu@reductions$umap@cell.embeddings[,2]
# 
# # before integration
# DimPlot(seu, 
#         reduction = 'umap',
#         pt.size = 0.00001,
#         label.box = FALSE,
#         label = FALSE, 
#         shuffle = TRUE,
#         raster=FALSE,
#         repel = TRUE,
#         label.size = 2, 
#         group.by = "batchID2") + #NoLegend() +
#   scale_color_manual(values=batch.colours) +
#   labs(title="Pre-integration",x = NULL, y = NULL) + 
#   guides(x = "none", y = "none") +
#   theme(plot.title = element_text(size=10),
#         text=element_text(family="Arial"),
#         axis.text = element_text(size=8),
#         axis.title.y=element_text(vjust=-1,
#                                   size=10),
#         axis.title.x=element_text(#vjust=10,
#                                   size=10),
#         legend.text = element_text(size = 10),
#         strip.text.x = element_text(face = "bold", vjust=2, size=12)) -> p1a
# 
# # after integration
# DimPlot(seu2, 
#         reduction = 'umap',
#         pt.size = 0.00001,
#         label.box = FALSE,
#         label = FALSE, 
#         shuffle = TRUE,
#         raster=FALSE,
#         repel = TRUE,
#         label.size = 2, 
#         group.by = "batchID2") + #NoLegend() +
#   scale_color_manual(values=batch.colours) +
#   labs(title="Post-integration", x = NULL, y = NULL) + 
#   guides(x = "none", y = "none") +
#   theme(plot.title = element_text(size=10),
#         text=element_text(family="Arial"),
#         axis.text = element_text(size=8),
#         axis.title.y=element_text(vjust=-1,
#                                   size=10),
#         axis.title.x=element_text(#vjust=10,
#                                   size=10),
#         legend.text = element_text(size = 10),
#         strip.text.x = element_text(face = "bold", vjust=2, size=12)) -> p1b
# 
# p1a|p1b
# 
# # split by batchID
# DimPlot(seu2, 
#         reduction = 'umap',
#         pt.size = 0.00001,
#         label.box = FALSE,
#         label = FALSE, 
#         shuffle = FALSE,
#         raster=FALSE,
#         repel = TRUE,
#         label.size = 2, 
#         split.by = "batchID2",
#         group.by = "celltype") + #NoLegend() +
#   scale_color_manual(values=Annotation.colours) +
#   labs(x = NULL, y = NULL) + 
#   guides(x = "none", y = "none") +
#   theme(plot.title = element_blank(),
#         text=element_text(family="Arial"),
#         axis.text = element_text(size=8),
#         axis.title.y=element_text(vjust=-1,
#                                   size=10),
#         axis.title.x=element_text(#vjust=10,
#                                   size=10),
#         legend.text = element_text(size = 10),
#         strip.text.x = element_text(face = "bold", vjust=2, size=12)) -> p1c
#         
# suppressWarnings(p1c)

```

### 3.6.2 Proportion

```{r fig.width=2, fig.height=2}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=seu$celltype,
                             sample=seu$batchID2)
                             
props$Proportions %>%
  data.frame() -> df

# stacked bar plot of cell proportion per sampleID
pe2 <- ggplot(df, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", position="stack") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=8,
                                   colour = "black"),
        axis.text.y = element_text(size=8,
                                   colour = "black"),
        legend.text = element_text(size = 8),
        legend.title = element_text(face="bold", size=12)) +
  labs( y = "Cell proportion", fill="batchID2", x=NULL) + NoLegend() +
  guides(fill = guide_legend(keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = Annotation.colours) +
  theme(legend.text.align = 0, legend.title = element_blank())
pe2

```

## 3.7 Save object

```{r}
out <- here("data","SCEs","manual.annot","bal.SEU.qs")
if(!file.exists(out)) {
  qsave(seu, file=out,nthreads =24)
}
```

# 4 Visualize results after annotation

## 4.1 Load object

```{r}
# 10.07.2025
out <- here("data","SCEs","manual.annot","bal.SEU.qs")
if(file.exists(out)) {
  seu <- qread(out,nthreads =24)
}

# set color
Annotation.colours <- setNames(c("#054a91", # AM
                                 "#dda15e", # RecM
                                 "#81a4cd", # Prolif M
                                 "#a3b18a", # cDC
                                 "#81b29a", # pDC
                                 "#3a5a40", # DC migratory
                                 "#000000", # Mast
                                 "#c3a995", # T
                                 "#7f5539", # NK
                                 "#5a189a", # B
                                 "#b23a48" # Epithelial
                                 ), # Secretory
               levels(seu$celltype))
```

## 4.2 Proportion per major cell type

```{r fig.height=2.5, fig.width=5}
# Subset object
seu <- subset(seu, Condition != "CF.ETI" & 
              sampleID != "M1C170C" & sampleID != "M1C176" & sampleID != "M1C180")
gc()
seu$Condition <- fct_drop(seu$Condition)
seu$sampleID <- fct_drop(seu$sampleID)

# Add Condition2
draft1 <- factor(
      case_when(
        seu$Condition == "Control" ~ "C"))
draft2 <- factor(
  case_when(
    seu$Bronchiectasis == "No" ~ "CFN",
    seu$Bronchiectasis == "Yes" ~ "CFB"
  )
)

Condition2 <- coalesce(draft1,draft2)
names(Condition2) <- colnames(seu)
seu <- AddMetaData(seu, Condition2, col.name = "Condition2")
seu$Condition2 <- factor(seu$Condition2,levels=c("C","CFB","CFN"))

# Rename Condition
seu$Condition <- factor(
  case_when(
    seu$Condition == "Control" ~ "C",
    seu$Condition == "CF" ~ "CF"
  ),
  levels=c("C","CF")
)
# Assign ID for analysis
seu$sampleID <- factor(seu$sampleID,
                       levels=c("M1N066","M1N075","M1N078","M1N080","M1N087","M1N092", # C
                                "M1C160(1)","M1C188","M1C166(1)","M1C190C","M1C196B","M1C198B","M1C208A","M1C209A", # CFB
                                "M1C170D","M1C176C","M1C180D","M1C177","M1C191B","M1C199B","M1C201A","M1C207B" # CFN
                                ))

ID <- setNames(c(paste0("C-",1:6),paste0("CFB-",1:8),paste0("CFN-",1:8)),
               levels(seu$sampleID))

ID <- setNames(factor(ID[seu$sampleID],levels=c(paste0("C-",1:6),paste0("CFB-",1:8),paste0("CFN-",1:8)),),
               colnames(seu))

seu <- AddMetaData(seu, ID, col.name = "ID")

# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=seu$celltype,
                             sample=seu$ID)
                             
props$Proportions %>%
  data.frame %>%
  dplyr::mutate(Condition = c(rep("C",66),rep("CF",176)),
                Condition2 = c(rep("C",66),rep("CFB",88),rep("CFN",88))) -> df

df$Condition <- factor(df$Condition, levels=c("C","CF"))
df$Condition2 <- factor(df$Condition2, levels=c("C","CFB","CFN"))
df$sample <- factor(df$sample, levels=c(paste0("C-",1:6),paste0("CFB-",1:8),paste0("CFN-",1:8)))

# stacked bar plot of cell proportion per sampleID
p2 <- ggplot(df, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", position="stack") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=8,
                                   colour = "black"),
        axis.text.y = element_text(size=8,
                                   colour = "black"),
        legend.text = element_text(size = 8),
        legend.title = element_text(face="bold", size=12)) +
  labs( y = "Cell proportion", fill="Subtype", x=NULL) + NoLegend() +
  guides(fill = guide_legend(keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = Annotation.colours) +
  theme(legend.text.align = 0, legend.title = element_blank())
p2
```

## 4.3 UMAP

### 4.3.1 Split by Condition2

```{r fig.height=3.5, fig.width=10}
p1.nosplit <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.00001,
        label.box = FALSE,
        label = FALSE, 
        shuffle = FALSE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2, 
        #split.by = "Condition2",
        group.by = "celltype") + #NoLegend() +
  scale_color_manual(values=Annotation.colours) +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 12),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))

DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.00001,
        label.box = FALSE,
        label = FALSE, 
        shuffle = FALSE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2, 
        split.by = "Condition2",
        group.by = "celltype") + #NoLegend() +
  scale_color_manual(values=Annotation.colours) +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        #text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 12),
        strip.text.x = element_text(face = "bold", vjust=2, size=14)) -> p1
suppressWarnings(p1)

```

### 4.3.2 For Epithelial/RecM/AM/ProlifM

```{r fig.height=3, fig.width=3}
# Annotation.colours <- setNames(c("#054a91", # AM
#                                  "#dda15e", # RecM
#                                  "#81a4cd", # Prolif M
#                                  "#a3b18a", # cDC
#                                  "#81b29a", # pDC
#                                  "#3a5a40", # DC migratory
#                                  "#000000", # Mast
#                                  "#c3a995", # T
#                                  "#7f5539", # NK
#                                  "#5a189a", # B
#                                  "#b23a48" # Epithelial
#                                  ), # Secretory
#                levels(seu$celltype))
# 
# Annotation.colours <- setNames(c("#ced4da", # AM
#                                  "#dda15e", # RecM
#                                  "#ced4da", # Prolif M
#                                  "#ced4da", # cDC
#                                  "#ced4da", # pDC
#                                  "#ced4da", # DC migratory
#                                  "#ced4da", # Mast
#                                  "#ced4da", # T
#                                  "#ced4da", # NK
#                                  "#ced4da", # B
#                                  "#ced4da" # Epithelial
#                                  ), # Secretory
#                levels(seu$celltype))
# 
# p4 <- DimPlot(seu, 
#         reduction = 'umap',
#         pt.size = 0.00001,
#         label.box = FALSE,
#         label = FALSE, 
#         shuffle = FALSE,
#         raster=FALSE,
#         repel = TRUE,
#         label.size = 2, 
#         #split.by = "Condition2",
#         group.by = "celltype") + NoLegend() +
#   scale_color_manual(values=Annotation.colours) +
#   labs(x = NULL, y = NULL) + 
#   guides(x = "none", y = "none") +
#   theme(plot.title = element_blank(),
#         text=element_text(family="Arial"),
#         axis.text = element_text(size=8),
#         axis.title.y=element_text(vjust=-1,
#                                   size=10),
#         axis.title.x=element_text(#vjust=10,
#                                   size=10),
#         legend.text = element_text(size = 12),
#         strip.text.x = element_text(face = "bold", vjust=2, size=12))
# 
# 
# p4
# 
# 
# # Prolif M UMAP
# celltype <- "RecM"
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 1600)
# 
# p4 +
#   theme(plot.tag = element_text(size = 14, face = "bold"))
# 
# ggsave(plot=last_plot(),
#        device="jpeg",
#        dpi=1600,
#        filename = here("output","plots","combined",paste0(celltype,".smallumap.jpeg")),
#        height=3,
#        width=3,
#        bg = "transparent")
# 
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 100)
```

## 4.4 Dotplot of marker genes

```{r fig.height=4.5, fig.width=10}
# Run Cepo
cepoMarkers <- Cepo(seu[["SCT"]]@data,
                    seu$celltype,
                    exprsPct = 0.2)
cepoMarkers$stats
cepoMarkers$stats <- cepoMarkers$stats[, c("AM","RecM","Prolif.M","cDC","pDC",
                                           "DC.migratory","Mast","T","NK","B",
                                           "Epithelial")]

sapply(1:ncol(cepoMarkers$stats), function(i){
  names(sort(cepoMarkers$stats[,i], decreasing = TRUE))[1:300]
}) %>% data.frame() -> dat

colnames(dat) <- colnames(cepoMarkers$stats)
dat %>% data.frame() %>% write.csv(file = here("data","Major.markers.csv"),row.names = FALSE)

# Cepo marker gene -------------------------------------------------------------
maxGenes <- 3
sigGenes <- lapply(1:ncol(dat), function(i){
  dat[,i][1:maxGenes] #%>% sort()
})

sig <- unlist(sigGenes)
cepo <- sig #sig[!duplicated(sig)]
# T
cepo[22:24] <- c("CD3D","CD3G","CD3E")

# TRM & RecM
cepo2 <- c("FABP4","GPD1","PCOLCE2","MME","SCD","VCAN","FCN1","EMP1","CD14","CCL2",cepo[7:length(cepo)]) #


cepo2 <- cepo2[!duplicated(cepo2)]

# plot
DefaultAssay(seu) <- "SCT"
seu$celltype <- factor(seu$celltype, levels=rev(names(Annotation.colours)))

# horizontal
p3 <- DotPlot(seu,
        features = cepo2,
        group.by = "celltype",
        col.max = 2,
        col.min = -2,
        dot.scale = 6
        ) + #coord_flip() +
  geom_point(aes(size=pct.exp), shape = 21, colour="black", stroke=0.5) +
  #FontSize(y.text = 10, x.text = 8) + 
  labs(y = element_blank(), x = element_blank()) + 
  theme(text=element_text(family="Ariel"),
        axis.text.x = element_text(size=10,
                                   face="italic",
                                   angle = 45,
                                   hjust = 1,
                                   vjust = 1),
        axis.text.y = element_text(size=11),
        legend.direction="vertical",
        legend.position="right",
        legend.justification = "right",
        legend.box="vertical",
        legend.text = element_text(size = 8),
        #legend.title = element_text(size = 8,family = "Arial"),
        ) +
  scale_color_gradient(low = brewer.ylorrd(9)[1], high=brewer.ylorrd(9)[8]) +
  #scale_y_discrete(labels = subtype.labels) +
  scale_size_continuous(limits = c(0, 100), 
                        breaks = c(0, 25, 50, 75, 100),
                        labels=c(0,"",50,"",100)) +
  guides(size=guide_legend(override.aes=list(shape=21, 
                                             colour="black", 
                                             fill="#adb5bd"),
                                             title="% of exp. cells",
                                             label.position="right",
                                             title.position="left",
                                             title.hjust=0.5,
                           title.theme= element_text(angle = 90, vjust = 0.5, size=10)
                                             ))

p3$guides$colour$title <- "Mean Expression"
p3$guides$colour$title.position <- "left"
p3$guides$colour$title.hjust <- 0.5
p3$guides$colour$title.theme <- element_text(angle = 90, vjust = 0.5, size=10)
seu$celltype <- factor(seu$celltype, levels=names(Annotation.colours))

suppressWarnings(print(p3))


```

## 4.5 CFTR expression proportion test

```{r}
Condition2.colours <- setNames(c("#26547c","#ef476f","#ffd166"),
                               c("C","CFB","CFN"))

DefaultAssay(seu) <- "RNA"
cftr.list <- lapply(levels(seu$celltype), FUN = function(x) {
  pt <- DotPlot(subset(seu, celltype==x), features = "CFTR", group.by = "ID")
  dat <- pt$data %>% mutate(celltype=x)
  return(dat)
})

df <- do.call("rbind", cftr.list) %>% dplyr::select(pct.exp,id,celltype) %>% dplyr::filter(celltype=="Epithelial")
colnames(df) <- c("pct.exp","ID","celltype")
rownames(df) <- NULL

n.celltype <- length(levels(seu$celltype))

df <- df %>% arrange(ID) %>% dplyr::mutate(Condition=c(rep("CF",16),
                                                       rep("C",6)),
                                           Condition2=c(rep("CFN",8),
                                                        rep("CFB",8),
                                                        rep("C",6)))

df$Condition <- factor(df$Condition, levels=c("C","CF"))
df$Condition2 <- factor(df$Condition2, levels=c("C","CFN","CFB"))

# stacked bar plot of cell proportion per sampleID
p5 <- ggplot(df, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", position="stack") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=8,
                                   colour = "black"),
        axis.text.y = element_text(size=8,
                                   colour = "black"),
        legend.text = element_text(size = 8),
        legend.title = element_text(face="bold", size=12)) +
  labs( y = "Proportion", fill="Subtype", x="Participant") +
  guides(fill = guide_legend(keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = Annotation.colours) +
  theme(legend.text.align = 0, legend.title = element_blank())

# Wilcox test on Healthy control vs CF -----------------------------------------
stat.test <- df %>%
  group_by(celltype) %>%
  wilcox_test(pct.exp ~ Condition, ref.group = "C") %>%
  adjust_pvalue(method="BH") %>%
  add_significance()

stat.test <- stat.test %>%
  add_xy_position(x = "celltype", dodge = 0.8)

# box plot
p6 <- ggboxplot(df, x="celltype",
          y="pct.exp",
          fill="Condition",
          palette = c("#26547c","#fb8b24"), outlier.shape=NA
          ) +
  stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01,
                     bracket.nudge.y = 0, vjust = 0, hide.ns = TRUE
                     ) +
  labs(y="Proportion of CFTR-expressing cells") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=10,
                                   colour = "black"),
        axis.title.x=element_blank(),
        axis.text.y = element_text(size=8,
                                   colour = "black"),
        legend.text = element_text(size = 10),
        legend.position = "right",
        legend.title = element_blank())

# Wilcox test on pairwise comparisons of control, CFN and CFB --------
stat.test <- df %>%
  group_by(celltype) %>%
  wilcox_test(pct.exp ~ Condition2, ref.group = "C") %>%
  adjust_pvalue(method="BH") %>%
  add_significance()

stat.test
stat.test <- stat.test %>%
  add_xy_position(x = "celltype", dodge = 0.8)

# box plot
p7 <- ggboxplot(df, x="celltype",
          y="pct.exp",
          fill="Condition2",
          palette = Condition2.colours, outlier.shape=NA
          ) +
  stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01,
                     bracket.nudge.y = 0, vjust = 0, hide.ns = TRUE
                     ) +
  labs(y="Proportion of CFTR-expressing cells") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=10,
                                   colour = "black"),
        axis.title.x=element_blank(),
        axis.text.y = element_text(size=8,
                                   colour = "black"),
        legend.text = element_text(size = 10),
        legend.position = "right",
        legend.title = element_blank())

p6
p7

```

# 5 Save figure

```{r}
# celltype <- "combined"
# if(!dir.exists(here("output","plots",celltype))) {
#   dir.create(here("output","plots",celltype),recursive=TRUE)
# }
# 
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 1600)
# 
# # batch effect
# suppressWarnings((p1a|p1b)) +
#   theme(plot.tag = element_text(size = 14, face = "bold", family="Ariel"))
# ggsave(plot=last_plot(),
#        device="jpeg",
#        dpi=1600,
#        filename = here("output","plots",celltype,"bal_combined.umap.groupByBatch.jpeg"),
#        height=3,
#        width=8)
# 
# suppressWarnings(p1c) +
#   theme(plot.tag = element_text(size = 14, face = "bold", family="Ariel"))
# ggsave(plot=last_plot(),
#        device="jpeg",
#        dpi=1600,
#        filename = here("output","plots",celltype,"bal_combined.umap.splitByBatch.jpeg"),
#        height=3,
#        width=13)
# 
# # umap
# p1 +
#   theme(plot.tag = element_text(size = 14, face = "bold"))
# 
# ggsave(plot=last_plot(),
#        device="jpeg",
#        dpi=1600,
#        filename = here("output","plots",celltype,"bal_combined.umap.jpeg"),
#        height=3.5,
#        width=10.5)
# 
# # no split
# # umap
# p1.nosplit +
#   theme(plot.tag = element_text(size = 14, face = "bold", family="Ariel"))
# 
# ggsave(plot=last_plot(),
#        device="jpeg",
#        dpi=1600,
#        filename = here("output","plots",celltype,"bal_combined_nosplit.umap.jpeg"),
#        height=4,
#        width=5)
# 
# # proportion
# p2 +
#   theme(plot.tag = element_text(size = 14, face = "bold", family="Ariel"))
# 
# ggsave(plot=last_plot(),
#        device="jpeg",
#        dpi=1600,
#        filename = here("output","plots",celltype,"bal_combined.prop.jpeg"),
#        height=3,
#        width=6)
# 
# # dotplot of marker genes
# suppressWarnings(print(p3 +
#   theme(plot.tag = element_text(size = 14, face = "bold", family="Ariel"))))
# 
# ggsave(plot=last_plot(),
#        device="jpeg",
#        dpi=1600,
#        filename = here("output","plots",celltype,"bal_combined.dotplot.jpeg"),
#        height=4.5,
#        width=10)
# 
# # Prolif M UMAP
# p4 +
#   theme(plot.tag = element_text(size = 14, face = "bold"))
# 
# ggsave(plot=last_plot(),
#        device="jpeg",
#        dpi=1600,
#        filename = here("output","plots",celltype,"ProlifM.smallumap.jpeg"),
#        height=3,
#        width=3,
#        bg = "transparent")
# 
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 100)
```

# 6 Save object

```{r}
if(!dir.exists(here("data","SCEs","manual.annot"))) {
  dir.create(here("data","SCEs","manual.annot"), recursive = TRUE)
}

out <- here("data","SCEs","manual.annot","bal_RecM.SEU.qs")
if(!file.exists(out)) {
  s <- subset(seu, celltype == "RecM")
  qsave(s,file=out,nthreads =16)
}

out <- here("data","SCEs","manual.annot","bal_AM.SEU.qs")
if(!file.exists(out)) {
  s <- subset(seu, celltype %in% c("AM","Prolif M","RecM"))
  qsave(s,file=out,nthreads =24)
}

out <- here("data","SCEs","manual.annot","bal_Epithelial.SEU.qs")
if(!file.exists(out)) {
  s <- subset(seu, celltype == "Epithelial")
  qsave(s,file=out,nthreads =16)
}

# 18.07.2025
out <- here("data","SCEs","analysis","bal.SEU.qs")
if(!file.exists(out)) {
  qsave(seu,file=out,nthreads =24)
}
```

# 7 Save annotation

```{r}
# 1) Write out annotations from the T cell subclustering seu
out_dir  <- here("data", "annotation")
out_file <- file.path(out_dir, "bal_annotations.csv")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

ann <- tibble(
  cell      = rownames(seu@meta.data),
  sampleID  = seu$sampleID,
  subtype   = as.character(seu$celltype)
)

readr::write_csv(ann, out_file)
out_file

```
