---
title: 'Analysis of Recruited monocytes and macrophages (RecM)'
author:
- name: Anson Wong
date: "`r Sys.Date()`"
description: null
output: html_document
editor_options:
  chunk_output_type: inline
---

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
xaringanExtra::use_panelset()

```

# 1 Load packages
```{r}
suppressPackageStartupMessages({
library(Seurat)
library(here)
library(qs)
library(forcats)
library(pals)
library(ggplot2)
library(ggpubr)
library(speckle)
library(rstatix)
library(scCustomize)
library(patchwork)
library(dplyr)
library(Cepo)
library(phateR)
library(slingshot)
library(readr)
library(cowplot)
library(scattermore)
library(ggrastr)
library(viridis)
library(scater)
library(circlize)
library(ggplotify)
library(tibble)
library(ComplexHeatmap)
library(RColorBrewer)
library(reshape2)
library(tradeSeq)
library(sceasy)
})
set.seed(1990)

```

# 2 Annotation, integration, clustering, finding marker genes
Cells annotated as RecM in the combined dataset were extracted and re-clustered.
Methods for normalization, integration, clustering and annotation were same as before, except that a resolution of 2 was selected.

For details, refer to code/bal_RecM.integration_clustering.R

# 3 Manual annotation - RecM
## 3.1 Load clustered object
```{r}
seu <- qread(here("data",
                  "SCEs",
                  "clustering",
                  "bal_RecM.integrated.clustered.SEU.qs"),nthreads=16)

seu$predicted.ann_finest_level <- factor(seu$predicted.ann_finest_level)
seu$sampleID <- factor(seu$sampleID)
seu$Capture <- factor(seu$Capture)
seu$Condition <- factor(seu$Condition)
seu$Bronchiectasis <- factor(seu$Bronchiectasis)

# SCT_snn_res.2 was selected for subclustering
levels(seu$SCT_snn_res.2)
seu$SCT_snn_res.3 <- factor(seu$SCT_snn_res.2,
                            levels=as.character(seq(1,27)))

# flip the UMAP x axis for easier visualization
seu@reductions$umap@cell.embeddings[,1] <- -seu@reductions$umap@cell.embeddings[,1]

```

## 3.2 Visualize UMAP embeddings
```{r}
# prepare colours
colours <- c(brewer.paired(12), brewer.pastel1(9), brewer.accent(8), brewer.set1(9),rev(brewer.dark2(8)), 
             brewer.set3(12))

p1 <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        label = FALSE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2.5, 
        cols=colours,
        # cols=c("#054a91", "#fb6f92", "#81a4cd", "#b79ced","7d451b","#a3b18a", 
        #        "#0e9594", "#f77f00", "#b23a48"),
        split.by = "Condition",
        group.by = "predicted.ann_finest_level") +
  #scale_x_continuous(breaks = c(-5, 0, 5)) +
  #scale_color_manual(values=Annotation.colours) +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))

p2 <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        #label.box = TRUE,
        label = FALSE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2.5, 
        cols=colours,
        # cols=c("#054a91", "#fb6f92", "#81a4cd", "#b79ced","7d451b","#a3b18a", 
        #        "#0e9594", "#f77f00", "#b23a48"),
        split.by = "Condition",
        group.by = "SCT_snn_res.2") + #NoLegend() +
  #scale_x_continuous(breaks = c(-5, 0, 5)) +
  #scale_color_manual(values=Annotation.colours) +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))

p3 <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        label = FALSE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2.5, 
        cols=colours,
        # cols=c("#054a91", "#fb6f92", "#81a4cd", "#b79ced","7d451b","#a3b18a", 
        #        "#0e9594", "#f77f00", "#b23a48"),
        # split.by = "Condition",
        group.by = "celltype") +
  #scale_x_continuous(breaks = c(-5, 0, 5)) +
  #scale_color_manual(values=Annotation.colours) +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))

p4 <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        label.box = TRUE,
        label = TRUE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 4, 
        cols=colours,
        # cols=c("#054a91", "#fb6f92", "#81a4cd", "#b79ced","7d451b","#a3b18a", 
        #        "#0e9594", "#f77f00", "#b23a48"),
        # split.by = "Condition",
        group.by = "SCT_snn_res.2") + NoLegend() +
  #scale_x_continuous(breaks = c(-5, 0, 5)) +
  #scale_color_manual(values=Annotation.colours) +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))

p1
p2
p3
p4

```

```{r fig.height=3, fig.width=9}
VlnPlot(seu,c("CCR2","VCAN"), group.by = "SCT_snn_res.2")

```

## 3.3 Visualize proportion of cells per cluster
```{r}

labels <- c("predicted.ann_finest_level","SCT_snn_res.2")
p <- vector("list",length(labels))

for(label in labels){
  seu@meta.data %>%
    ggplot(aes(x = !!sym(label), fill = !!sym(label))) +
    geom_bar() +
    geom_text(aes(label = ..count..), stat = "count",
              angle=90, colour = "black", size = 2) +
    scale_y_log10() +
    theme(axis.text.x = element_blank(),
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank()) +
    NoLegend() + scale_fill_manual(values = rev(colours)) +
    labs(y = "No. Cells (log scale)") -> p1
  
  seu@meta.data %>%
    dplyr::select(!!sym(label), Condition) %>%
    group_by(!!sym(label), Condition) %>%
    summarise(num = n()) %>%
    mutate(prop = num / sum(num)) %>%
  ggplot(aes(x = !!sym(label), y = prop * 100, 
             fill = Condition)) + 
    geom_bar(stat = "identity") +
    theme(axis.text.x = element_text(angle = 45, size=6,
                                     vjust = 1, 
                                     hjust = 1)) +
    labs(y = "% Cells", fill = "Condition") -> p2
  
  p1 / p2 -> p[[label]]
}
p

```

## 3.4 Manual annotation
```{r}
subtype <- setNames(c("IM","MDM-PLA2G7hi","MDM-FABP4hi","MDM-PLA2G7hi","MDM-FABP4hi", # 1-5
                      "Mono","Mono","MDM-FABP4lo","MDM-FABP4hi","MDM-PLA2G7hi", # 6-10
                      "MDM-FABP4lo","MDM-FABP4lo","MDM-PLA2G7hi","MDM-FABP4hi","MDM-IV", # 11-15
                      "MDM-FABP4lo","MDM-FABP4hi","MDM-FABP4hi","IM","MDM-FABP4lo", # 16-20
                      "MDM-inflam","MDM-FABP4lo","MDM-inflam","MDM-FABP4lo","MDM-FABP4hi", # 21-25
                      "MDM-inflam","IM" # 26-27
                       ), levels(seu$SCT_snn_res.2))

subtype <- factor(setNames(subtype[seu$SCT_snn_res.2],colnames(seu)),
                     levels=c("Mono","MDM-IV","MDM-FABP4lo","MDM-FABP4hi","MDM-inflam","MDM-PLA2G7hi","IM"
                              ))

seu <- AddMetaData(seu, metadata = subtype, col.name = "subtype")

```

## 3.5 Save object
```{r}
celltype <- "RecM"
date <- Sys.Date()
if(!dir.exists(here("data","SCEs","final.annot"))) {
  dir.create(here("data","SCEs","final.annot"), recursive = TRUE)
}

out <- here("data","SCEs","final.annot",paste0("bal_",celltype,".final_annot.SEU.qs"))
if(!file.exists(out)) {
  qsave(seu,file=out,nthreads=24)
}

```


# 4 Analysis
## 4.1 Read processed object
```{r}
# load analysis object
celltype <- "RecM"
seu <- qread(here("data","SCEs","final.annot",paste0("bal_",celltype,".final_annot.SEU.qs")),
             nthreads=24)
```

## 4.2 Prepare object for downstream analysis
```{r}
# Subset object
seu <- subset(seu, subset = Condition != "CF.ETI" & 
              sampleID != "M1C170C" & sampleID != "M1C176" & sampleID != "M1C180")

seu$Condition <- fct_drop(seu$Condition)
seu$sampleID <- fct_drop(seu$sampleID)

# Add Condition2
draft1 <- factor(
      case_when(
        seu$Condition == "Control" ~ "C"))
draft2 <- factor(
  case_when(
    seu$Bronchiectasis == "No" ~ "CFN",
    seu$Bronchiectasis == "Yes" ~ "CFB"
  )
)

Condition2 <- coalesce(draft1,draft2)
names(Condition2) <- colnames(seu)
seu <- AddMetaData(seu, Condition2, col.name = "Condition2")
seu$Condition2 <- factor(seu$Condition2,levels=c("C","CFB","CFN"))
gc()

# Rename Condition
seu$Condition <- factor(
  case_when(
    seu$Condition == "Control" ~ "C",
    seu$Condition == "CF" ~ "CF"
  ),
  levels=c("C","CF")
)
# Assign ID for analysis
seu$sampleID <- factor(seu$sampleID,
                       levels=c("M1N066","M1N075","M1N078","M1N080","M1N087","M1N092", # healthy
                                "M1C160(1)","M1C188","M1C166(1)","M1C190C","M1C196B","M1C198B","M1C208A","M1C209A", # CFB
                                "M1C170D","M1C176C","M1C180D","M1C177","M1C191B","M1C199B","M1C201A","M1C207B" # CFN
                                ))

ID <- setNames(c(paste0("C-",1:6),paste0("CFB-",1:8),paste0("CFN-",1:8)),
               levels(seu$sampleID))

ID <- setNames(factor(ID[seu$sampleID],levels=c(paste0("C-",1:6),paste0("CFB-",1:8),paste0("CFN-",1:8)),),
               colnames(seu))

seu <- AddMetaData(seu, ID, col.name = "ID")

# Set colours
subtype.colours <- setNames(c("#f2cc8f","#7209b7","#83c5be","#3a6ea5","#9e2a2b","#9c6644","#9a8c98"),
                            c("Mono","MDM-IV","MDM-FABP4lo","MDM-FABP4hi","MDM-inflam","MDM-PLA2G7hi","IM"))
         
   
Condition2.colours <- setNames(c("#26547c","#ef476f","#ffd166"),
                               levels(seu$Condition2))
Condition.colours <- setNames(c("#26547c","#fb8b24"),
                              levels(seu$Condition))

subtype.labels <- c("Mono","MDM-IV",
                    expression('MDM-'*italic('FABP4')^'lo'),expression('MDM-'*italic('FABP4')^'hi'),
                    "MDM-inflam",expression('MDM-'*italic('PLA2G7')^'hi'),
                    "IM")
```

## 4.3 UMAP embeddings
### 4.3.1 UMAP split by Condition2
```{r fig.height=3.5, fig.width=10}
# split by severity
p1 <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.2,
        label.box = FALSE,
        label = FALSE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 4, 
        cols=subtype.colours,
        split.by = "Condition2",
        group.by = "subtype") + #NoLegend() +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text.align = 0,
        legend.text = element_text(size = 12),
        strip.text.x = element_text(face = "bold", vjust=2, size=16)) +
  scale_color_manual(values = subtype.colours,
                    labels = subtype.labels) +
  guides(
    color = guide_legend(
      override.aes = list(
        size  = 5,    # make the points in the legend larger
        shape = 15    # change the legend symbol shape if you like
      )
    )
  )

suppressWarnings(print(p1))

```

### 4.3.2 UMAP colored by MDM-PLA2G7hi / MDM-inflam
```{r fig.height=3, fig.width=3}
Annotation.colours <- setNames(c("#ced4da", 
                                 "#ced4da",
                                 "#ced4da",
                                 "#ced4da", 
                                 "#9e2a2b", 
                                 "#ced4da",
                                 "#ced4da"
                                 ), # Secretory
               levels(seu$subtype))

DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.2,
        label.box = FALSE,
        label = FALSE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 4, 
        cols=subtype.colours,
        #split.by = "Condition2",
        group.by = "subtype") + #NoLegend() +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text.align = 0,
        legend.text = element_text(size = 12),
        strip.text.x = element_text(face = "bold", vjust=2, size=16)) +
  scale_color_manual(values = Annotation.colours,
                    labels = subtype.labels) +
  NoLegend() +
  guides(
    color = guide_legend(
      override.aes = list(
        size  = 5,    # make the points in the legend larger
        shape = 15    # change the legend symbol shape if you like
      )
    )
  ) -> p2

suppressWarnings(print(p2))

# umap
showtext::showtext_auto()
showtext::showtext_opts(dpi = 1600)

p2 +
  theme(plot.tag = element_text(size = 14, face = "bold", family="Ariel"))

ggsave(plot=last_plot(),
       device="jpeg",
       dpi=1600,
       filename = here("data","plots",celltype,"MDM-inflam_umap.jpeg"),
       height=3,
       width=3)
showtext::showtext_auto()
showtext::showtext_opts(dpi = 100)

```

## 4.4 Proportion of cells
```{r fig.width=5, fig.height=4}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=seu$subtype,
                             sample=seu$ID)
                             
props$Proportions %>%
  data.frame %>%
  dplyr::mutate(Condition = c(rep("C",42),rep("CF",112)),
                Condition2 = c(rep("C",42),rep("CFB",56),rep("CFN",56))) -> df

df$Condition <- factor(df$Condition, levels=c("C","CF"))
df$Condition2 <- factor(df$Condition2, levels=c("C","CFB","CFN"))

# stacked bar plot of cell proportion per sampleID
ggplot(df, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", position="stack") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=8,
                                   colour = "black"),
        axis.text.y = element_text(size=8,
                                   colour = "black"),
        legend.text = element_text(size = 8),
        legend.title = element_text(face="bold", size=12)) +
  labs( y = "Cell proportion in RecM", fill="Subtype", x="Participant") +
  guides(fill = guide_legend(keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = subtype.colours, labels = subtype.labels) +
  theme(legend.text.align = 0, legend.title = element_blank())

# Wilcox test on pairwise comparisons of control, CF mild and CF severe --------
stat.test <- df %>%
  group_by(clusters) %>%
  wilcox_test(
    ref.group = "C",
    Freq ~ Condition2,
    p.adjust.method = "BH",
  ) %>%
  filter(p.adj < 0.1)

stat.test <- stat.test %>%
  add_xy_position(x = "clusters", dodge = 0.8)

# box plot
dodge_width <- 0.8  # adjust this value to widen spacing between groups within clusters

p5 <- ggplot(df, aes(x = clusters, y = Freq, color = Condition2)) +
  geom_boxplot(
    fill = NA,
    outlier.shape = NA,
    position = position_dodge(width = dodge_width)
  ) +
  geom_point(
    position = position_dodge(width = dodge_width),
    size = 2, alpha = 0.8
  ) +
  stat_pvalue_manual(
    stat.test,
    label = "p.adj.signif",
    tip.length = 0.01,
    bracket.size = 0.4,
    hide.ns = TRUE
  ) +
  scale_color_manual(values = Condition2.colours) +
  labs(
    y = "Cell proportion in RecM"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_blank(),
    axis.text.x = element_text(angle = 40, hjust = 1, size = 10, colour = "black"),
    axis.text.y = element_text(size = 12, colour = "black"),
    axis.title.x = element_blank(),
    legend.title = element_blank(),
    legend.position = c(0.01, 0.99),
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.size = unit(1, "lines"),
    legend.text = element_text(size = 12)
  ) +
  scale_x_discrete(labels = subtype.labels)

p5
```

## 4.5 Marker gene analysis
### 4.5.1 Call marker genes with Cepo
```{r}
# Cepo
cepoMarkers <- Cepo(seu[["SCT"]]@data,
                    seu$subtype,
                    exprsPct = 0.2)
cepoMarkers$stats
cepoMarkers$stats <- cepoMarkers$stats[, c("Mono","MDM.IV","MDM.FABP4lo","MDM.FABP4hi","MDM.inflam","MDM.PLA2G7hi","IM")]
cepoMarkers$stats <- cepoMarkers$stats[!grepl('^A[C|F|P|L][0|1]', row.names(cepoMarkers$stats)),]

sapply(1:ncol(cepoMarkers$stats), function(i){
  names(sort(cepoMarkers$stats[,i], decreasing = TRUE))[1:300]
}) %>% data.frame() -> dat2

colnames(dat2) <- colnames(cepoMarkers$stats)
dat2

dat2 %>% data.frame() %>% write.csv(file=here("output","RecM.markers.csv"),row.names = FALSE)
```

### 4.5.2 DotPlot of selected features
```{r fig.width=4,fig.height=15}
# Selected ---------------------------------------------------------------------
selected <- c("CHIT1","CHI3L1","CD14","FCGR3A","FCN1","IL1B","S100A12",
              "S100A8","EREG","NAMPT","CD93","SERPINB9",
              "MARCKS","F13A1","MAF","FOLR2","RNASE1",
              "STAB1","PDGFC","MEF2C","MAFB","ITGA9",
              "SLC40A1","CCR2","CLEC5A","ITGAM","NLRP1",
              "NLRP12","VCAN","C1QA","C1QB","FABP4",
              "MARCO","CD163","MRC1","APOC1","PLA2G7",
              "LGMN","CD109","SPP1","MMP9","CD36",
              "MMP19","TGFBI","SDC2","GBP4","GBP5",
              "IDO1","CCL2","CCL8","CXCL9","CXCL10",
              "CXCL11","ISG15","ISG20","GCH1")

# function colors
alarmin <- "#E85D04"
cytokine <- "#D00000"
iron <- "#972D07"
inflammasome <- "#5A189A"
lipid <- "#003F88"
scavenger <- "#197278"
ecm <- "#FF5D8F"
others <- "#000000"

gene.colours <- setNames(c(ecm,ecm,others, others, others, cytokine,alarmin,
                           alarmin,rep(others,4),
                           others,rep(iron,3),others,
                           rep(iron,4),ecm,
                           iron, cytokine,others, others, inflammasome,
                           inflammasome,ecm,others,others,lipid,
                           rep(scavenger,3),lipid,ecm,
                           rep(ecm,5),
                           rep(ecm,3),rep(inflammasome,2),
                           others, rep(cytokine,4),
                           cytokine,rep(others,3)),
                         selected)

selected.order <- c("CD14","FCGR3A",
                    "CCR2","CLEC5A","ITGAM","NLRP1","NLRP12",
                    "F13A1","MAF","FOLR2","RNASE1",
                    "STAB1","PDGFC","MEF2C","MAFB","TGFBI","ITGA9",
                    "SLC40A1","VCAN","C1QA","C1QB","FABP4",
                    "MARCO","CD163","MRC1","APOC1",
                    "FCN1","IL1B","S100A12",
                    "S100A8","EREG","NAMPT","CD93","SERPINB9","MARCKS",
                    "PLA2G7","SPP1","CHIT1","CHI3L1","MMP9","MMP19","SDC2","LGMN","CD109","CD36",
                    "GBP4","GBP5",
                    "IDO1","CCL2","CCL8","CXCL9","CXCL10",
                    "CXCL11","ISG15","ISG20","GCH1")

gene.colours <- gene.colours[selected.order]

# plot
# # version 1: horizontal
# p6 <- DotPlot(seu,
#         features = selected.order,
#         group.by = "subtype",
#         col.max = 2,
#         col.min = -2,
#         dot.scale = 6
#         ) + #coord_flip() +
#   geom_point(aes(size=pct.exp), shape = 21, colour="black", stroke=0.5) +
#   FontSize(y.text = 10, x.text = 8) + 
#   labs(y = element_blank(), x = element_blank()) + 
#   theme(text=element_text(family="Arial"),
#         axis.text.x = element_text(size=10,
#                                    face="italic",
#                                    angle = 45,
#                                    hjust = 1,
#                                    color=gene.colours,
#                                    vjust = 1
#                                    ),
#         axis.text.y = element_text(size=10),
#         legend.direction="horizontal",
#         legend.position="bottom",
#         legend.justification = "right",
#         legend.box="vertical",
#         legend.text = element_text(size = 8),
#         legend.title = element_text(size = 8,family = "Arial"),
#         ) +
#   scale_color_gradient(low = brewer.ylorrd(9)[1], high=brewer.ylorrd(9)[8]) +
#   scale_y_discrete(labels = subtype.labels) +
#   scale_size_continuous(limits = c(0, 100), 
#                         breaks = c(0, 25, 50, 75, 100),
#                         labels=c(0,"",50,"",100)) +
#   guides(size=guide_legend(override.aes=list(shape=21, 
#                                              colour="black", 
#                                              fill="#adb5bd"),
#                                              title="% of exp. cells",
#                                              label.position="bottom",
#                                              title.position="top",
#                                              title.hjust=0.5
#                                              ))
# 
# p6$guides$colour$title <- "Mean Expression"
# p6$guides$colour$title.position <- "top"
# p6$guides$colour$title.hjust <- 0.5
# #suppressWarnings(print(p6))

# version 2: vertical
p6 <- DotPlot(seu,
        assay="SCT",
        features = rev(selected.order),
        group.by = "subtype",
        col.max = 2,
        col.min = -2,
        dot.scale = 6
        ) + coord_flip() +
  geom_point(aes(size=pct.exp), shape = 21, colour="black", stroke=0.5) +
  FontSize(y.text = 10, x.text = 8) + 
  labs(y = element_blank(), x = element_blank()) + 
  theme(#text=element_text(family="Arial"),
        axis.text.y = element_text(size=10,
                                   face="italic",
                                   #angle = 45,
                                   #hjust = 1,
                                   #vjust = 1,
                                   color=rev(gene.colours)),
        axis.text.x = element_text(size=10,angle=45,vjust=1,hjust=1),
        legend.direction="horizontal",
        legend.position="bottom",
        legend.justification = "right",
        legend.box="vertical",
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 8,family = "Arial"),
        ) +
  scale_color_gradient(low = brewer.ylorrd(9)[1], high=brewer.ylorrd(9)[8]) +
  scale_y_discrete(labels = subtype.labels) +
  scale_size_continuous(limits = c(0, 100), 
                        breaks = c(0, 25, 50, 75, 100),
                        labels=c(0,"",50,"",100)) +
  guides(size=guide_legend(override.aes=list(shape=21, 
                                             colour="black", 
                                             fill="#adb5bd"),
                                             title="% of exp. cells",
                                             label.position="bottom",
                                             title.position="top",
                                             title.hjust=0.5
                                             ))

p6$guides$colour$title <- "Mean Expression"
p6$guides$colour$title.position <- "top"
p6$guides$colour$title.hjust <- 0.5
suppressWarnings(print(p6))

```


### 4.5.3 Feature plot of representative markers
```{r fig.width=15, fig.height=12}
features <- c("CD14","FCGR3A","FCN1","S100A12",
              "CCR2","FOLR2","F13A1","VCAN",
              "IL1B","CCL2","CCL3","CCL4",
              "FABP4","C1QA","MARCO","APOC1",
              "PLA2G7","SPP1","MMP9","CHI3L1"
              )
p7 <- suppressMessages(FeaturePlot(seu,
            pt.size = 0.001,
            ncol = 5,
            order = FALSE,
            features = features) &
  scale_color_gradientn(colours = brewer.reds(9)),
                        #colours = c("#d6ccc2",brewer.brbg(9)[6:9])
                        ) &
    theme_minimal() &
    theme(panel.border = element_blank(), 
        panel.grid = element_blank(),
        axis.text=element_blank(),
        axis.title = element_blank(),
        plot.title = element_text(size = 16, hjust=0.5, face="bold.italic"),
        legend.key.size = unit(0.2, "cm"),
        legend.text = element_text(size = 6),
        legend.title = element_text(face="bold", size=12))
p7

```

### 4.5.4 Dotplot of MDM-inflam
```{r fig.height=3.5, fig.width=2}
# Read DE results
s <- qread(here("data","SCEs","analysis","bal_MDM-inflam.SEU.qs"),nthreads=16)

selected <- c("CCL7","CCL8","CXCL1","CXCL2","CXCL3","CXCL5","CXCL8","CCR7",
              "IL6","IL10","EREG","TNF","TNFAIP6","TNIP3","INHBA","ITGB8")
p5 <- DotPlot(s,
            features = rev(selected),
            group.by = "Condition2",
            col.max = 1,
            col.min = -1,
            dot.scale = 6
            ) + #coord_flip() +
    geom_point(aes(size=pct.exp), shape = 21, colour="black", stroke=0.5) +
    FontSize(y.text = 9, x.text = 8) + 
    labs(y = element_blank(), x = element_blank()) + 
      theme_bw() +
    theme(
          axis.text.x = element_text(size=10,
                                     
                                     angle = 45,
                                     hjust = 1,
                                     vjust = 1),
          axis.text.y = element_text(size=10,hjust=0,face="italic"),
          legend.direction="vertical",
          legend.position="right",
          legend.title.position = "left",
          legend.justification = "bottom",
          legend.box="horizontal",
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 8,angle=90),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank()
          ) + coord_flip() +
      
    scale_color_gradientn(colours = rev(brewer.rdbu(11)[2:10]),name = "Mean Expression") +
    scale_size_continuous(limits = c(0, 100),
                          breaks = c(0, 25, 50, 75, 100),
                          labels=c(0,"",50,"",100)) +
    guides(size=guide_legend(override.aes=list(shape=21, 
                                               colour="black", 
                                               fill="#adb5bd"),
                                               title="% of exp. cells",
                                               label.position="right",
                                               title.position="left",
                                               title.hjust=0.5
                                               ),
           colour = guide_colorbar(#barwidth       = unit(0.4, "cm"),
                                   #barheight      = unit(2,   "cm"),
                                   title.position = "left",
                                   title.hjust    = 0.5
                                  )) + NoLegend()

suppressWarnings(p5)

```

## 4.6 RNA velocity
### 4.6.1 Convert object to anndata
```{r}
out <- here("data","SCEs","analysis","RecM_velo.adata.h5ad")
if (!file.exists(out)) {
  sceasy::convertFormat(seu, from="seurat", to="anndata", outFile = here("data","SCEs","analysis","RecM.h5ad"))
  # proceed to scVelo
} else {
  s <- sceasy::convertFormat(out,from="anndata",to="seurat")
}

```

### 4.6.2 Run scVelo
see code/scVelo.ipynb

### 4.6.3 Plot diffusion map
```{r}
FeaturePlot(s, features=c("latent_time"), cols=viridis(256))
```

## 4.7 Gene-cytokine correlation
### 4.7.1 Aggregate expression
```{r}
# whole RecM
DefaultAssay(seu) <- "RNA"
seu <- NormalizeData(seu)

pseudo_bulk <- AggregateExpression(
  seu,
  assays = "RNA",
  slot = "counts",
  group.by = "sampleID",
  return.seurat = TRUE
)
pseudo_bulk <- NormalizeData(pseudo_bulk)

# MDM-PLA2G7hi
s <- subset(seu, subtype=="MDM-PLA2G7hi")
DefaultAssay(s) <- "RNA"

pseudo_bulk.fibro <- AggregateExpression(
  s,
  assays = "RNA",
  slot = "counts",
  group.by = "sampleID",
  return.seurat = TRUE
)
pseudo_bulk.fibro <- NormalizeData(pseudo_bulk.fibro)

# MDM-inflam
s <- subset(seu, subtype=="MDM-inflam")
DefaultAssay(s) <- "RNA"

pseudo_bulk.inflam <- AggregateExpression(
  seu,
  assays = "RNA",
  slot = "counts",
  group.by = "sampleID",
  return.seurat = TRUE
)
pseudo_bulk.inflam <- NormalizeData(pseudo_bulk.inflam)

# whole BAL
seu.combined <- qread(here("data","SCEs","manual.annot","bal.SEU.qs"),
                      nthreads=24) # process it in 04.annotation.Rmd
DefaultAssay(seu.combined) <- "RNA"

pseudo_bulk.all <- AggregateExpression(
  seu.combined,
  assays = "RNA",
  slot = "counts",
  group.by = "sampleID",
  return.seurat = TRUE
)
pseudo_bulk.all <- NormalizeData(pseudo_bulk.all)

# Use the assay/slot you aggregated into. Here we assume "RNA" and slot = "data"
pb_mat <- GetAssayData(pseudo_bulk, assay = "RNA", slot = "data")
pb_mat.fibro <- GetAssayData(pseudo_bulk.fibro, assay = "RNA", slot = "data")
pb_mat.inflam <- GetAssayData(pseudo_bulk.inflam, assay = "RNA", slot = "data")
pb_mat.all <- GetAssayData(pseudo_bulk.all, assay = "RNA", slot = "data")

```

### 4.7.2 Plot 
#### 4.7.2.1 Function
```{r}
plot_pb_gene_cyt <- function(pb_mat, gene = "SPP1", cyto = "Osteopontin",
                             res.final, pal = NULL, zero_filter = TRUE) {
  # minimal checks
  if (missing(pb_mat)) stop("pb_mat is required")
  if (!(is.matrix(pb_mat) || inherits(pb_mat, "Matrix") || is.data.frame(pb_mat))) {
  stop("pb_mat must be a base matrix, Matrix (sparse), or data.frame with genes as rows")
}
  if (!gene %in% rownames(pb_mat)) stop("gene '", gene, "' not found in pb_mat rownames")
  if (missing(res.final)) stop("res.final is required (must contain expr_combat and meta)")
  if (!"expr_combat" %in% names(res.final)) stop("res.final$expr_combat not found")
  if (!cyto %in% colnames(as.data.frame(res.final$expr_combat))) stop("cytokine '", cyto, "' not found in res.final$expr_combat")
  if (!"meta" %in% names(res.final) || !"Record_ID" %in% names(res.final$meta)) stop("res.final$meta with Record_ID required")
  if (!"Condition2" %in% names(res.final$meta)) stop("res.final$meta must contain Condition2")

  # build pb_df from pb_mat
  gene_vec <- as.numeric(pb_mat[gene, , drop = TRUE])
  sample_ids <- colnames(pb_mat)
  pb_df <- data.frame(sampleID = sample_ids, stringsAsFactors = FALSE)
  gene_col <- paste0(gene, "_expr")
  pb_df[[gene_col]] <- gene_vec

  # attach cytokine level: assume direct rowname match sampleID -> rowname(expr_combat)
  expr_c <- as.data.frame(res.final$expr_combat)
  matched_idx <- match(pb_df$sampleID, rownames(expr_c))
  if (!all(!is.na(matched_idx))) {
    # try normalized match for unmatched
    unmatched <- which(is.na(matched_idx))
    if (length(unmatched) > 0) {
      norm <- function(x) tolower(gsub("[^[:alnum:]]+", "", as.character(x)))
      row_norm <- norm(rownames(expr_c))
      pb_norm  <- norm(pb_df$sampleID[unmatched])
      norm_idx <- match(pb_norm, row_norm)
      matched_idx[unmatched] <- norm_idx
    }
  }
  cyto_col <- paste0(cyto, "_level")
  # assign (NA for no match)
  pb_df[[cyto_col]] <- as.numeric(ifelse(is.na(matched_idx), NA, expr_c[rownames(expr_c)[matched_idx], cyto]))
  pb_df$Record_ID_matched <- ifelse(is.na(matched_idx), NA, rownames(expr_c)[matched_idx])

  # attach Condition2 using matched Record_ID where possible, else try sampleID -> Record_ID
  meta_df <- as.data.frame(res.final$meta)
  pb_df$Condition2 <- meta_df$Condition2[ match(pb_df$Record_ID_matched, meta_df$Record_ID) ]
  nas <- which(is.na(pb_df$Condition2))
  if (length(nas) > 0) {
    pb_df$Condition2[nas] <- meta_df$Condition2[ match(pb_df$sampleID[nas], meta_df$Record_ID) ]
  }

  # optional: drop zero expression rows
  if (zero_filter) {
    pb_df <- pb_df[!is.na(pb_df[[gene_col]]) & pb_df[[gene_col]] != 0, ]
  } else {
    pb_df <- pb_df[!is.na(pb_df[[gene_col]]) & !is.na(pb_df[[cyto_col]]), ]
  }
  if (nrow(pb_df) == 0) stop("No rows available for plotting after matching/filtering")

  # palette resolution
  if (is.null(pal)) {
    if (exists("Condition2.colours")) pal <- get("Condition2.colours")
    else if (exists("condition2_cols")) pal <- get("condition2_cols")
  }
  if (is.null(pal)) {
    levs <- unique(pb_df$Condition2)
    pal <- setNames(RColorBrewer::brewer.pal(max(3, length(levs)), "Dark2")[seq_along(levs)], levs)
  }

  # compute numeric coords
  xr <- range(pb_df[[gene_col]], na.rm = TRUE)
  yr <- range(pb_df[[cyto_col]], na.rm = TRUE)
  x_pos <- xr[1] + 0.02 * diff(xr) - 0.1
  y_pos <- yr[2] - 0.01 * diff(yr) + 0.25

  # build plot using .data pronoun to safely reference potentially non-syntactic names
  p <- ggplot(pb_df, aes(x = .data[[gene_col]], y = .data[[cyto_col]], color = Condition2)) +
    geom_point(size = 3) +
    # geom_smooth(
    #   data = pb_df,
    #   mapping = aes(x = .data[[gene_col]], y = .data[[cyto_col]]),
    #   method = "lm",
    #   se = FALSE,
    #   color = "black",
    #   linewidth = 0.8,
    #   fullrange = TRUE,
    #   inherit.aes = FALSE,
    #   show.legend = FALSE
    # ) +
    # stat_ellipse(aes(fill = Condition2),
    #              geom = "polygon",
    #              level = 0.68,
    #              alpha = 0.18,
    #              show.legend = TRUE) +
    ggpubr::stat_cor(method = "pearson",
                     label.x = x_pos, label.y = y_pos,
                     size = 3, color = "gray10") +
    scale_color_manual(values = pal) +
    scale_fill_manual(values = pal) +
    coord_cartesian(
      xlim = range(pb_df[[gene_col]], na.rm = TRUE) + c(-0.1, 0.5),
      ylim = range(pb_df[[cyto_col]], na.rm = TRUE) + c(-0.1, 0.5)
    ) +
    labs(
      x = "Gene (log1p)",
      y = "Cytokine (log2 pg/mL)",
      color = NULL,
      title = gene
    ) +
    theme_minimal(base_size = 14) +
    theme(
      axis.text    = element_text(size = 12),
      axis.title   = element_text(size = 12),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.7),
      axis.line    = element_blank(),
      legend.position  = "bottom",
      plot.title   = element_text(hjust = 0.5, size = 16, face = "bold")
    ) +
    guides(fill = "none")

  return(p)
}

# load cytokine results
res.final <- qread(here("data","cytokine","res.master.qs"))
```

#### 4.7.2.2 Plot SPP1 and CHI3L1 [related to MDM-PLA2G7hi]
```{r fig.height=4, fig.width=7}
pc1 <- plot_pb_gene_cyt(pb_mat.fibro, "SPP1", "Osteopontin", res.final = res.final)
pc2 <- plot_pb_gene_cyt(pb_mat.fibro, "CHI3L1", "Chitinase-3-like1", res.final = res.final)
pc3 <- plot_pb_gene_cyt(pb_mat, "SPP1", "Osteopontin", res.final = res.final)
pc4 <- plot_pb_gene_cyt(pb_mat, "CHI3L1", "Chitinase-3-like1", res.final = res.final)

(pc1|pc2)

pc.chi3l1.fibro <- pc2
```

#### 4.7.2.3 Plot immune markers [RecM]
```{r fig.height=12, fig.width=10.5}
pc1 <- plot_pb_gene_cyt(pb_mat, "CCL2", "MCP-1(MCAF)", res.final = res.final)
pc2 <- plot_pb_gene_cyt(pb_mat, "CCL3", "MIP-1α", res.final = res.final)
pc3 <- plot_pb_gene_cyt(pb_mat, "CCL4", "MIP-1β", res.final = res.final)
# pc6 <- plot_pb_gene_cyt(pb_mat, "IL6", "IL-6", res.final = res.final)
pc4 <- plot_pb_gene_cyt(pb_mat, "CXCL8", "IL-8", res.final = res.final)
pc8 <- plot_pb_gene_cyt(pb_mat, "IL1RN", "IL-1ra", res.final = res.final)
# pc9 <- plot_pb_gene_cyt(pb_mat, "TNFSF13", "APRIL/TNFSF13", res.final = res.final)
# pc10 <- plot_pb_gene_cyt(pb_mat, "TNFSF12", "TWEAK/TNFSF12", res.final = res.final)
# pc11 <- plot_pb_gene_cyt(pb_mat, "TNFRSF1A", "sTNF-R1", res.final = res.final)
# pc12 <- plot_pb_gene_cyt(pb_mat, "TNFSF13B", "BAFF/TNFSF13B", res.final = res.final)
# pc13 <- plot_pb_gene_cyt(pb_mat, "TNF", "TNF-α", res.final = res.final)
pc5 <- plot_pb_gene_cyt(pb_mat, "CSF1", "M-CSF", res.final = res.final)
pc15 <- plot_pb_gene_cyt(pb_mat, "SPP1", "Osteopontin", res.final = res.final)
pc16 <- plot_pb_gene_cyt(pb_mat, "CHI3L1", "Chitinase-3-like1", res.final = res.final)
pc17 <- plot_pb_gene_cyt(pb_mat, "CD163", "sCD163", res.final = res.final)

plot.list.RecM <- list(pc1,pc2,pc3,pc4,pc5)
names(plot.list.RecM) <- c("CCL2","CCL3","CCL4","CXCL8","CSF1")

(pc1|pc2|pc3)/(pc4|pc8|pc5)/(pc15|pc16|pc17)

pc.chi3l1.RecM <- pc16
```

#### 4.7.2.4 Plot immune markers [all]
```{r fig.height=18, fig.width=10.5}
pc3 <- plot_pb_gene_cyt(pb_mat.all, "CCL2", "MCP-1(MCAF)", res.final = res.final)
pc4 <- plot_pb_gene_cyt(pb_mat.all, "CCL3", "MIP-1α", res.final = res.final)
pc5 <- plot_pb_gene_cyt(pb_mat.all, "CCL4", "MIP-1β", res.final = res.final)
pc6 <- plot_pb_gene_cyt(pb_mat.all, "IL6", "IL-6", res.final = res.final)
pc7 <- plot_pb_gene_cyt(pb_mat.all, "CXCL8", "IL-8", res.final = res.final)
pc8 <- plot_pb_gene_cyt(pb_mat.all, "IL1RN", "IL-1ra", res.final = res.final)
pc9 <- plot_pb_gene_cyt(pb_mat.all, "TNFSF13", "APRIL/TNFSF13", res.final = res.final)
pc10 <- plot_pb_gene_cyt(pb_mat.all, "TNFSF12", "TWEAK/TNFSF12", res.final = res.final)
pc11 <- plot_pb_gene_cyt(pb_mat.all, "TNFRSF1A", "sTNF-R1", res.final = res.final)
pc12 <- plot_pb_gene_cyt(pb_mat.all, "TNFSF13B", "BAFF/TNFSF13B", res.final = res.final)
pc13 <- plot_pb_gene_cyt(pb_mat.all, "TNF", "TNF-α", res.final = res.final)
pc14 <- plot_pb_gene_cyt(pb_mat.all, "CSF1", "M-CSF", res.final = res.final)
pc15 <- plot_pb_gene_cyt(pb_mat.all, "SPP1", "Osteopontin", res.final = res.final)
pc16 <- plot_pb_gene_cyt(pb_mat.all, "CHI3L1", "Chitinase-3-like1", res.final = res.final)
pc17 <- plot_pb_gene_cyt(pb_mat.all, "CD163", "sCD163", res.final = res.final)
(pc3|pc4|pc5)/(pc6|pc7|pc8)/(pc9|pc10|pc11)/(pc12|pc13|pc14)/(pc15|pc16|pc17)

pc.chi3l1.all <- pc16
```

#### 4.7.2.5 Plot CHI3L1 correlation across different levels
```{r fig.width=10.5, fig.height=4}
(pc.chi3l1.all + labs(x="Gene (log1p) - All")) | 
  (pc.chi3l1.RecM + labs(x="Gene (log1p) - RecM")) | 
  (pc.chi3l1.fibro + labs(x=expression("Gene (log1p) - MDM-" * italic("PLA2G7")^"hi"))) -> p.chi3l1
p.chi3l1
```


## 4.8 Pseudobulk gene expression
### 4.8.1 Plot Function
```{r}
plot_pb_gene <- function(pb_mat, gene, res.final, pal = NULL, zero_filter = TRUE, dodge_width = 0.8) {
  # minimal checks
  if (missing(pb_mat)) stop("pb_mat is required")
  if (missing(gene)) stop("gene is required")
  if (missing(res.final)) stop("res.final is required (must contain expr_combat and meta)")
  if (!(is.matrix(pb_mat) || inherits(pb_mat, "Matrix") || is.data.frame(pb_mat))) {
    stop("pb_mat must be a base matrix, Matrix (sparse), or data.frame with genes as rows")
  }
  if (!gene %in% rownames(pb_mat)) stop("gene '", gene, "' not found in pb_mat rownames")
  if (!"expr_combat" %in% names(res.final)) stop("res.final$expr_combat not found")
  if (!"meta" %in% names(res.final) || !"Record_ID" %in% names(res.final$meta)) stop("res.final$meta with Record_ID required")
  if (!"Condition2" %in% names(res.final$meta)) stop("res.final$meta must contain Condition2")

  # build pb_df
  sample_ids <- colnames(pb_mat)
  gene_col <- paste0(gene, "_expr")
  gene_vec <- as.numeric(pb_mat[gene, , drop = TRUE])
  pb_df <- data.frame(sampleID = sample_ids, stringsAsFactors = FALSE)
  pb_df[[gene_col]] <- gene_vec

  # attach cytokine/meta mapping using res.final (try exact rowname match, fallback to normalized match)
  expr_c <- as.data.frame(res.final$expr_combat)
  matched_idx <- match(pb_df$sampleID, rownames(expr_c))
  if (!all(!is.na(matched_idx))) {
    unmatched <- which(is.na(matched_idx))
    if (length(unmatched) > 0) {
      norm <- function(x) tolower(gsub("[^[:alnum:]]+", "", as.character(x)))
      row_norm <- norm(rownames(expr_c))
      pb_norm  <- norm(pb_df$sampleID[unmatched])
      norm_idx <- match(pb_norm, row_norm)
      matched_idx[unmatched] <- norm_idx
    }
  }
  pb_df$Record_ID_matched <- ifelse(is.na(matched_idx), NA, rownames(expr_c)[matched_idx])
  meta_df <- as.data.frame(res.final$meta)
  pb_df$Condition2 <- meta_df$Condition2[ match(pb_df$Record_ID_matched, meta_df$Record_ID) ]
  nas <- which(is.na(pb_df$Condition2))
  if (length(nas) > 0) {
    pb_df$Condition2[nas] <- meta_df$Condition2[ match(pb_df$sampleID[nas], meta_df$Record_ID) ]
  }
  if ("Condition2" %in% names(pb_df)) pb_df$Condition2 <- factor(pb_df$Condition2, levels = c("C","CFB","CFN"))

  # apply zero filter if requested
  if (zero_filter) {
    pb_df <- pb_df[!is.na(pb_df[[gene_col]]) & pb_df[[gene_col]] != 0, ]
  } else {
    pb_df <- pb_df[!is.na(pb_df[[gene_col]]) & !is.na(pb_df$Condition2), ]
  }
  if (nrow(pb_df) == 0) stop("No samples available after filtering/matching")

  # palette
  if (is.null(pal)) {
    if (exists("Condition2.colours")) pal <- get("Condition2.colours")
    else if (exists("condition2_cols")) pal <- get("condition2_cols")
  }
  if (is.null(pal)) {
    levs <- unique(pb_df$Condition2)
    pal <- setNames(RColorBrewer::brewer.pal(max(3, length(levs)), "Dark2")[seq_along(levs)], levs)
  }

  # Wilcoxon tests
  library(rstatix)
  groups_present <- unique(na.omit(pb_df$Condition2))
  stat.test.pb <- NULL
  if (length(groups_present) >= 2) {
    fmla <- as.formula(paste0(gene_col, " ~ Condition2"))
    stat.test.pb <- tryCatch(
      pb_df %>%
        wilcox_test(formula = fmla, p.adjust.method = "BH") %>%
        add_significance("p.adj") %>%
        add_xy_position(x = "Condition2"),
      error = function(e) {
        warning("Failed to compute pairwise wilcox_test: ", conditionMessage(e))
        NULL
      }
    )
  }

  # Plot
  library(ggplot2)
  library(scales)
  p <- ggplot(pb_df, aes(x = Condition2, y = .data[[gene_col]], color = Condition2)) +
    geom_boxplot(fill = NA, outlier.shape = NA, position = position_dodge(width = dodge_width)) +
    geom_point(position = position_dodge(width = dodge_width), size = 2, alpha = 0.9) +
    scale_color_manual(values = pal) +
    scale_y_continuous(labels = number_format(accuracy = 0.1)) +  # <-- force 1 decimal place
    labs(y = "Pseudobulk expression (log1p)", title = gene) +
    theme_classic(base_size = 16) +
    theme(
      axis.text.x      = element_text(angle = 40, hjust = 1, size = 10, colour = "black"),
      axis.text.y      = element_text(size = 10, colour = "black"),
      axis.title.y     = element_text(size = 12),
      axis.title.x     = element_blank(),
      axis.line        = element_blank(),
      panel.border     = element_rect(size = 0.5, fill = "transparent"),
      legend.position  = "none",
      plot.title       = element_text(hjust = 0.5, size = 16, face = "bold.italic")
    )

  if (!is.null(stat.test.pb) && nrow(stat.test.pb) > 0) {
    p <- p + stat_pvalue_manual(
      stat.test.pb,
      label = "p.adj.signif",
      tip.length = 0.01,
      bracket.size = 0.4,
      hide.ns = TRUE
    )
  }

  return(p)
}


```

### 4.8.2 Plot
```{r fig.height=4, fig.width=4}
fibro.genes <- c("SPP1","CHIT1","CHI3L1","ITGB8","MMP9","MMP19","PLA2G7","LGMN","MERTK")
inflam.genes <- c("CCL2","CCR2","CCL3","CCL4","IL6","CXCL8","IL1RN","TNFSF13","TNFSF12","TNFRSF1A","TNFSF13B","TNF","CSF1")
genes <- c(fibro.genes, inflam.genes)

# MDM-PLA2G7hi - profibrotic
plot.list.fibro <- lapply(fibro.genes, FUN=function(g){
  plot_pb_gene(pb_mat.fibro, g, res.final = res.final, zero_filter = FALSE)
})
names(plot.list.fibro) <- fibro.genes

# MDM-inflam
plot.list.inflam <- lapply(genes, FUN=function(g){
  plot_pb_gene(pb_mat.inflam, g, res.final = res.final, zero_filter = FALSE)
})
names(plot.list.inflam) <- genes

# RecM
plot.list <- list()
plot.list <- lapply(genes, FUN=function(g){
  plot_pb_gene(pb_mat, g, res.final = res.final)
})
names(plot.list) <- genes

# all
plot.list <- list()
plot.list <- lapply(genes, FUN=function(g){
  plot_pb_gene(pb_mat.all, g, res.final = res.final)
})
names(plot.list) <- genes

# check plots
plot.list.fibro$CHI3L1
plot.list$CHI3L1

```

## 4.9 Write tables
```{r}
props$Proportions %>% t() %>% as.data.frame() %>% write_csv(here("output","fig2D.csv"))

data.frame(pseudo_bulk.fibro@assays$RNA@data[c("CHI3L1","MMP9","CHIT1","SPP1"),] %>% t())  %>% write_csv(here("output","Fig2E.csv"))
```

# 5 Save figures
```{r}
# celltype <- "RecM"
# out <- here("output","plots",celltype)
# if(!dir.exists(out)) {
#   dir.create(out,recursive = TRUE)
# }
# 
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 1600)
# 
# # umap split by Condition2
# p1 +
#   theme(plot.tag = element_text(size = 14, face = "bold"))
# 
# ggsave(plot=last_plot(),
#        device="jpeg",
#        dpi=1600,
#        filename = paste0(out,"/","bal_",celltype,".umap.splitbyCondition2.jpeg"),
#        height=3.5,
#        width=10)
# 
# 
# # cell proportion
# p5 +
#   theme(plot.tag = element_text(size = 14, face = "bold"))
# 
# ggsave(plot=last_plot(),
#        device="jpeg",
#        dpi=1600,
#        filename = paste0(out,"/","bal_",celltype,".cellprop.jpeg"),
#        height=4,
#        width=5)
# 
# # dotplot
# suppressWarnings(print(p6 +
#   theme(plot.tag = element_text(size = 14, face = "bold"
#                                 ))))
# ggsave(plot=last_plot(),
#        device="jpeg",
#        dpi=1600,
#        filename = paste0(out,"/","bal_",celltype,".dotplot_vertical.jpeg"),
#        height=15,
#        width=4)
# 
# # MDM-inflam dotplot
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 1600)
# suppressWarnings(print(p5 +
#   theme(plot.tag = element_text(size = 14, face = "bold"
#                                 ))))
# ggsave(plot=last_plot(),
#        device="jpeg",
#        dpi=1600,
#        filename = paste0(out,"/","bal_MDM-inflam.dotplot.jpeg"),
#        height=3.5,
#        width=2)
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 100)
# 
# # Boxplot of MDM-inflam
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 1600)
# 
# for (g in genes) {
#   plot.list.inflam[[g]] +
#     theme(plot.tag = element_text(size = 14, face = "bold"))
# 
#   ggsave(plot=last_plot(),
#          device="jpeg",
#          dpi=1600,
#          filename = paste0(out,"/","MDM-inflam.pseudobulk.",g,".jpeg"),
#          height=3,
#          width=2.5)
# }
# 
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 100)
# 
# # Boxplot of RecM
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 1600)
# 
# for (g in genes) {
#   plot.list[[g]] +
#     theme(plot.tag = element_text(size = 14, face = "bold"))
# 
#   ggsave(plot=last_plot(),
#          device="jpeg",
#          dpi=1600,
#          filename = paste0(out,"/","RecM.pseudobulk.",g,".jpeg"),
#          height=3,
#          width=2.5)
# }
# 
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 100)
# 
# # Violin plot of all
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 1600)
# 
# for (g in genes) {
#   plot.list[[g]] +
#     theme(plot.tag = element_text(size = 14, face = "bold"))
# 
#   ggsave(plot=last_plot(),
#          device="jpeg",
#          dpi=1600,
#          filename = paste0(out,"/","combined.pseudobulk.",g,".jpeg"),
#          height=3,
#          width=2.5)
# }
# 
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 100)
# 
# # gene-cytokine plot for SPP1 and CHI3L1
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 1600)
# 
# # umap split by Condition2
# pc1|pc2 +
#   theme(plot.tag = element_text(size = 14, face = "bold"))
# 
# ggsave(plot=last_plot(),
#        device="jpeg",
#        dpi=1600,
#        filename = paste0(out,"/","MDM-PLA2G7hi_fibro",".pseudobulk.jpeg"),
#        height=4,
#        width=7)
# 
# pc3|pc4 +
#   theme(plot.tag = element_text(size = 14, face = "bold"))
# 
# ggsave(plot=last_plot(),
#        device="jpeg",
#        dpi=1600,
#        filename = paste0(out,"/","RecM_fibro",".pseudobulk.jpeg"),
#        height=4,
#        width=7)
# 
# 
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 100)
# 
# # gene-cytokine plot
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 1600)
# for (p in names(plot.list.RecM)) {
#   ggsave(plot=plot.list.RecM[[p]] +
#            theme(plot.tag = element_text(size = 14, face = "bold")),
#        device="jpeg",
#        dpi=1600,
#        filename = paste0(out,"/","RecM.",p,".gene-cyto_NoEllipse.jpeg"),
#        height=4,
#        width=3.5)
# }
# 
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 100)
# 
# # chi3l1
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 1600)
# ggsave(plot=p.chi3l1 +
#            theme(plot.tag = element_text(size = 14, face = "bold")),
#        device="jpeg",
#        dpi=1600,
#        filename = paste0(out,"/","CHI3L1.gene-cyto.jpeg"),
#        height=4,
#        width=10.5)
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 100)

```

# 6 Save object
```{r}
out <- here("data","SCEs","analysis",celltype,paste0("bal_",celltype,".analysis.SEU.qs"))

if (!file.exists(out)) {
  qsave(seu,file=out,nthreads=16)
}

```

# 7 Save subclustering results
```{r}
# 1) Write out annotations from the T cell subclustering seu
out_dir  <- here("output")
out_file <- file.path(out_dir, "bal_RecM_subtype_annotations.csv")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

ann_t <- tibble(
  cell      = rownames(seu@meta.data),
  sampleID  = seu$sampleID,
  subtype   = as.character(seu$subtype)
)

readr::write_csv(ann_t, out_file)
out_file

```
