---
title: 'Analysis of tissue-resident alveolar macrophages (AM)'
author:
- name: Anson Wong
date: "`r Sys.Date()`"
description: null
output: html_document
editor_options:
  chunk_output_type: inline
---

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
xaringanExtra::use_panelset()

```

# 1 Load packages
```{r}
suppressPackageStartupMessages({
library(Seurat)
library(here)
library(qs)
library(forcats)
library(pals)
library(ggplot2)
library(ggpubr)
library(speckle)
library(rstatix)
library(scCustomize)
library(patchwork)
library(dplyr)
library(Cepo)
library(phateR)
library(slingshot)
library(tradeSeq)
library(org.Hs.eg.db)
library(clusterProfiler)
library(ReactomePA)
library(readr)
library(cowplot)
library(harmony)
library(scattermore)
library(ggrastr)
library(viridis)
library(scater)
library(circlize)
library(ggplotify)
})
set.seed(1990)

```

# 2 Annotation, integration, clustering, finding marker genes
Cells annotated as AM, Prolif M and RecM in the combined dataset were extracted for subclustering.
RecM cluster was included for subclustering to aid a better TRM projection in UMAP, but they are not included for analysis here.
Methods for normalization, integration, clustering and annotation were the same.

For details, refer to code/bal_TRM.integration_clustering.R

# 3 Subclustering
## 3.1 Load clustered object
```{r}
seu <- qread(here("../archive_cf-eti-bal/SCEs/bal_TRM.integrated.clustered.SEU.qs"),
             nthreads=24) # with RecM

seu$predicted.ann_finest_level <- factor(seu$predicted.ann_finest_level)
seu$sampleID <- factor(seu$sampleID)
seu$Capture <- factor(seu$Capture)
seu$Condition <- factor(seu$Condition)
seu$Bronchiectasis <- factor(seu$Bronchiectasis)

# SCT_snn_res.2 was selected for subclustering
levels(seu$SCT_snn_res.3)
seu$SCT_snn_res.3 <- factor(seu$SCT_snn_res.3,
                            levels=as.character(seq(1,56)))

# flip the UMAP x axis for easier visualization
seu@reductions$umap@cell.embeddings[,2] <- -seu@reductions$umap@cell.embeddings[,2]

```

## 3.2 Visualize UMAP embeddings
```{r}
# prepare colours
colours <- c(brewer.paired(12), brewer.pastel1(9), brewer.accent(8), brewer.set1(9),rev(brewer.dark2(8)), 
             brewer.set3(12))

p1 <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        label = FALSE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2.5, 
        cols=colours,
        # cols=c("#054a91", "#fb6f92", "#81a4cd", "#b79ced","7d451b","#a3b18a", 
        #        "#0e9594", "#f77f00", "#b23a48"),
        split.by = "Condition",
        group.by = "predicted.ann_finest_level") +
  #scale_x_continuous(breaks = c(-5, 0, 5)) +
  #scale_color_manual(values=Annotation.colours) +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))

p2 <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        #label.box = TRUE,
        label = FALSE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2.5, 
        cols=colours,
        # cols=c("#054a91", "#fb6f92", "#81a4cd", "#b79ced","7d451b","#a3b18a", 
        #        "#0e9594", "#f77f00", "#b23a48"),
        split.by = "Condition",
        group.by = "SCT_snn_res.3") + #NoLegend() +
  #scale_x_continuous(breaks = c(-5, 0, 5)) +
  #scale_color_manual(values=Annotation.colours) +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))

p3 <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        label = FALSE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2.5, 
        cols=colours,
        # cols=c("#054a91", "#fb6f92", "#81a4cd", "#b79ced","7d451b","#a3b18a", 
        #        "#0e9594", "#f77f00", "#b23a48"),
        # split.by = "Condition",
        group.by = "celltype") +
  #scale_x_continuous(breaks = c(-5, 0, 5)) +
  #scale_color_manual(values=Annotation.colours) +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))

p4 <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        label.box = TRUE,
        label = TRUE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2, 
        cols=colours,
        # cols=c("#054a91", "#fb6f92", "#81a4cd", "#b79ced","7d451b","#a3b18a", 
        #        "#0e9594", "#f77f00", "#b23a48"),
        # split.by = "Condition",
        group.by = "SCT_snn_res.3") + NoLegend() +
  #scale_x_continuous(breaks = c(-5, 0, 5)) +
  #scale_color_manual(values=Annotation.colours) +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))

p1
p2
p3
p4

```

## 3.3 Visualize proportion of cells per cluster
```{r}

labels <- c("predicted.ann_finest_level","SCT_snn_res.3")
p <- vector("list",length(labels))

for(label in labels){
  seu@meta.data %>%
    ggplot(aes(x = !!sym(label), fill = !!sym(label))) +
    geom_bar() +
    geom_text(aes(label = ..count..), stat = "count",
              angle=90, colour = "black", size = 2) +
    scale_y_log10() +
    theme(axis.text.x = element_blank(),
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank()) +
    NoLegend() + scale_fill_manual(values = rev(colours)) +
    labs(y = "No. Cells (log scale)") -> p1
  
  seu@meta.data %>%
    dplyr::select(!!sym(label), Condition) %>%
    group_by(!!sym(label), Condition) %>%
    summarise(num = n()) %>%
    mutate(prop = num / sum(num)) %>%
  ggplot(aes(x = !!sym(label), y = prop * 100, 
             fill = Condition)) + 
    geom_bar(stat = "identity") +
    theme(axis.text.x = element_text(angle = 45, size=6,
                                     vjust = 1, 
                                     hjust = 1)) +
    labs(y = "% Cells", fill = "Condition") -> p2
  
  p1 / p2 -> p[[label]]
}
p

```

## 3.4 Manual annotation
```{r}
## this is using UMAP with harmony-adjusted embeddings.
# draft1
draft1 <- factor(
      case_when(
        seu$celltype == "RecM" ~ "RecM",
        seu$celltype == "Prolif M" ~ "Prolif M"))
# draft2
draft2 <- setNames(c("AM-IFI27hi","AM-IFI27hi","AM","AM","AM", # 1-5
                     "AM-CCL","AM","AM","RecM","AM", # 6-10
                     "AM-lipid","RecM","AM","AM-lipid","AM", # 11-15
                     "AM-CCL","AM-IFI27hi","AM","AM","AM-MT", # 16-20
                     "AM","AM","AM","AM-CCL","AM", # 21-25
                     "Prolif M","AM-IFI27hi","AM-CCL","AM","AM", # 26-30
                     "Prolif M","AM","AM","AM","AM", # 31-35
                     "AM-CCL","AM","AM-lipid","AM","AM", # 36-40
                     "AM-lipid","AM-CCL","AM","AM","AM-CCL", # 41-45
                     "AM","RecM","AM-CCL","AM","AM-CCL", # 46-50
                     "AM","AM","RecM","Prolif M","AM-IFI27hi", # 51-55
                     "AM"
                     ), levels(seu$SCT_snn_res.3))

draft2 <- factor(setNames(draft2[seu$SCT_snn_res.3],colnames(seu)),
                     levels=c("AM-CCL",
                              "AM-MT",
                              "AM-lipid",
                              "AM-IFI27hi",
                              "AM",
                              "RecM",
                              "Prolif M"
                              ))

subtype <- coalesce(draft1,draft2)
seu <- AddMetaData(seu, metadata = subtype, col.name = "subtype")

# Remove AM cells that are closely clustered with RecM
seu <- subset(seu, subtype %in% c("AM-CCL", "AM-MT", "AM-lipid", "AM-IFI27hi", "AM", "Prolif M", "RecM"))
seu$subtype <- fct_drop(seu$subtype)
seu$subtype <- factor(seu$subtype, levels=c("AM-CCL", "AM-MT", "AM-lipid", "AM-IFI27hi", "AM", "Prolif M", "RecM"))

```

## 3.5 Save object
```{r}
celltype <- "AM"
date <- Sys.Date()
if(!dir.exists(here("data","SCEs","final.annot"))) {
  dir.create(here("data","SCEs","final.annot"), recursive = TRUE)
}

out <- here("data","SCEs","final.annot",paste0("bal_",celltype,".final_annot.SEU.qs"))
if(!file.exists(out)) {
  qsave(seu,file=out,nthreads=24)
}

```


# 4 Analysis
## 4.1 Read processed object
```{r}
# load analysis object
celltype <- "AM"
seu <- qread(here("data","SCEs","final.annot",paste0("bal_",celltype,".final_annot.SEU.qs")),
             nthreads=24)

```

## 4.2 Prepare object for downstream analysis
```{r}
# Subset object
seu <- subset(seu, Condition != "CF.ETI" & 
              sampleID != "M1C170C" & sampleID != "M1C176" & sampleID != "M1C180")
gc()
seu$Condition <- fct_drop(seu$Condition)
seu$sampleID <- fct_drop(seu$sampleID)

# Add Condition2
draft1 <- factor(
      case_when(
        seu$Condition == "Control" ~ "C"))
draft2 <- factor(
  case_when(
    seu$Bronchiectasis == "No" ~ "CFN",
    seu$Bronchiectasis == "Yes" ~ "CFB"
  )
)

Condition2 <- coalesce(draft1,draft2)
names(Condition2) <- colnames(seu)
seu <- AddMetaData(seu, Condition2, col.name = "Condition2")
seu$Condition2 <- factor(seu$Condition2,levels=c("C","CFB","CFN"))

# Rename Condition
seu$Condition <- factor(
  case_when(
    seu$Condition == "Control" ~ "C",
    seu$Condition == "CF" ~ "CF"
  ),
  levels=c("C","CF")
)
# Assign ID for analysis
seu$sampleID <- factor(seu$sampleID,
                       levels=c("M1N066","M1N075","M1N078","M1N080","M1N087","M1N092", # C
                                "M1C160(1)","M1C188","M1C166(1)","M1C190C","M1C196B","M1C198B","M1C208A","M1C209A", # CFB
                                "M1C170D","M1C176C","M1C180D","M1C177","M1C191B","M1C199B","M1C201A","M1C207B" # CFN
                                ))

ID <- setNames(c(paste0("C-",1:6),paste0("CFB-",1:8),paste0("CFN-",1:8)),
               levels(seu$sampleID))

ID <- setNames(factor(ID[seu$sampleID],levels=c(paste0("C-",1:6),paste0("CFB-",1:8),paste0("CFN-",1:8)),),
               colnames(seu))

seu <- AddMetaData(seu, ID, col.name = "ID")

# Set colours
subtype.colours <- setNames(c( "#7209b7","#ef476f", "#ffd166", "#83c5be",
                                 "#3a6ea5", "#9a8c98", "#ced4da"),
                               levels(seu$subtype))
Condition2.colours <- setNames(c("#26547c","#ef476f","#ffd166"),
                               levels(seu$Condition2))
Condition.colours <- setNames(c("#26547c","#fb8b24"),
                              levels(seu$Condition))

subtype.labels <- c("AM-CCL","AM-MT","AM-lipid", 
                    expression('AM-'*italic('IFI27')^'hi'),
                    "AM","Prolif M","RecM")
```

## 4.3 UMAP embeddings
```{r fig.height=3.5, fig.width=10}
# split by severity
p1 <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.2,
        label.box = FALSE,
        label = FALSE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 4, 
        cols=subtype.colours,
        split.by = "Condition2",
        group.by = "subtype") + #NoLegend() +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text.align = 0,
        legend.text = element_text(size = 12),
        strip.text.x = element_text(face = "bold", vjust=2, size=16)) +
  scale_color_manual(values = subtype.colours,
                    labels = subtype.labels) +
  guides(
    color = guide_legend(
      override.aes = list(
        size  = 5    # make the points in the legend larger
        #shape = 15    # change the legend symbol shape if you like
      )
    )
  )

suppressWarnings(print(p1))

```

## 4.4 Proportion of cells
```{r fig.width=5, fig.height=4}
# calculate the proportion of cells --------------------------------------------
seu.trm <- subset(seu, subtype %in% c("Prolif M","RecM"),invert=TRUE)
seu.trm$subtype <- fct_drop(seu.trm$subtype)

props <- getTransformedProps(clusters=seu.trm$subtype,
                             sample=seu.trm$ID)
                             
props$Proportions %>%
  data.frame %>%
  dplyr::mutate(Condition = c(rep("C",30),rep("CF",80)),
                Condition2 = c(rep("C",30),rep("CFB",40),rep("CFN",40))) -> df

df$Condition <- factor(df$Condition, levels=c("C","CF"))
df$Condition2 <- factor(df$Condition2, levels=c("C","CFB","CFN"))

# stacked bar plot of cell proportion per sampleID
ggplot(df, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", position="stack") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=8,
                                   colour = "black"),
        axis.text.y = element_text(size=8,
                                   colour = "black"),
        legend.text = element_text(size = 8),
        legend.title = element_text(face="bold", size=12)) +
  labs( y = "Proportion", fill="Subtype", x="Participant") +
  guides(fill = guide_legend(keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = subtype.colours, labels = subtype.labels) +
  theme(legend.text.align = 0, legend.title = element_blank())

# Wilcox test on pairwise comparisons of control, CF mild and CF severe --------
stat.test <- df %>%
  group_by(clusters) %>%
  wilcox_test(Freq ~ Condition2, 
              ref.group = "C", 
              p.adjust.method = "BH")
stat.test <- stat.test %>%
  add_xy_position(x = "clusters", dodge = 0.8)

# box plot
dodge_width <- 0.8  # adjust this value to widen spacing between groups within clusters

p5 <- ggplot(df, aes(x = clusters, y = Freq, color = Condition2)) +
  geom_boxplot(
    fill = NA,
    outlier.shape = NA,
    position = position_dodge(width = dodge_width)
  ) +
  geom_point(
    position = position_dodge(width = dodge_width),
    size = 2, alpha = 0.8
  ) +
  stat_pvalue_manual(
    stat.test,
    label = "p.adj.signif",
    tip.length = 0.01,
    bracket.size = 0.4,
    hide.ns = TRUE
  ) +
  scale_color_manual(values = Condition2.colours) +
  labs(
    y = "Proportion"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_blank(),
    axis.text.x = element_text(angle = 40, hjust = 1, size = 10, colour = "black"),
    axis.text.y = element_text(size = 12, colour = "black"),
    axis.title.x = element_blank(),
    legend.title = element_blank(),
    legend.position = c(0.01, 0.99),
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.size = unit(1, "lines"),
    legend.text = element_text(size = 12)
  ) +
  scale_x_discrete(labels = subtype.labels)

p5

```


## 4.4 Marker gene analysis
### 4.4.1 Call marker genes with Cepo
```{r}
# Cepo
cepoMarkers <- Cepo(seu[["SCT"]]@data,
                    seu$subtype,
                    exprsPct = 0.2)
cepoMarkers$stats <- cepoMarkers$stats[, c("AM.CCL","AM.MT","AM.lipid","AM.IFI27hi","AM","Prolif.M","RecM")]
cepoMarkers$stats <- cepoMarkers$stats[!grepl('^A[C|F|P|L][0|1]', row.names(cepoMarkers$stats)),]

sapply(1:ncol(cepoMarkers$stats), function(i){
  names(sort(cepoMarkers$stats[,i], decreasing = TRUE))[1:300]
}) %>% data.frame() -> dat2

colnames(dat2) <- colnames(cepoMarkers$stats)
dat2

colnames(dat2) <- colnames(cepoMarkers$stats)
dat2 %>% data.frame() %>% write.csv(file=here("data","AM.markers.csv"))
```

### 4.4.2 DotPlot of selected features
```{r fig.height=7, fig.width=4}
# Selected ---------------------------------------------------------------------
universal <- c("FABP4","MRC1","MARCO","C1QA")
ccl <- c("CCL4","CCL3","CCL4L2","CCL20","CXCL5","MARCKS","TNFAIP6")
mt <- c("MT1X","MT1G","MT1F","MT1E","MT1H")
lipid <- c("LAMP1","CD36","ABCA1","LINC01146","HDAC9")
ifi27 <- c("IFI27","OVCH1","SLC19A3","AQP3","GPD1")
fibro <- c("SPP1","MMP9","CCL18","CHI3L1","CX3CR1")
#prolif <- c("PCLAF","MKI67","NUSAP1","CENPF","TOP2A")

selected <- c(universal,ccl,mt,lipid,ifi27)

# function colors
scavenger <- "#197278"
cytokine <- "#5A189A"
metallothionein <- "#d00000"
lipid <- "#003F88"
#cycling <- "#972d07"
ecm <- "#FF5D8F"
others <- "#000000"

gene.colours <- setNames(c(lipid, scavenger, scavenger, others, #universal
                           rep(cytokine,5),others,others, #ccl
                           rep(metallothionein,5), #mt
                           rep(lipid,5), #lipid
                           rep(others,4),lipid #ifi27
),
                         selected)
# plot
subtype.labels <- c("AM-CCL","AM-MT","AM-lipid", 
                    expression('AM-'*italic('IFI27')^'hi'),
                    "AM")

# version 2: vertical
p2 <- DotPlot(seu.trm,
        features = rev(selected),
        group.by = "subtype",
        col.max = 2,
        col.min = -2,
        dot.scale = 6
        ) + coord_flip() +
  geom_point(aes(size=pct.exp), shape = 21, colour="black", stroke=0.5) +
  FontSize(y.text = 10, x.text = 8) + 
  labs(y = element_blank(), x = element_blank()) + 
  theme(
        axis.text.y = element_text(size=10,
                                   face="italic",
                                   #angle = 45,
                                   #hjust = 1,
                                   #vjust = 1,
                                   color=rev(gene.colours)),
        axis.text.x = element_text(size=10,angle=45,vjust=1,hjust=1),
        legend.direction="vertical",
        legend.position="right",
        legend.justification = "bottom",
        legend.box="vertical",
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 8,family = "Arial",angle=90),
        ) +
  scale_color_gradient(low = brewer.ylorrd(9)[1], high=brewer.ylorrd(9)[8], name="Mean Expression") +
  scale_y_discrete(labels = subtype.labels) +
  scale_size_continuous(limits = c(0, 100), 
                        breaks = c(0, 25, 50, 75, 100),
                        labels=c(0,"",50,"",100)) +
  guides(size=guide_legend(override.aes=list(shape=21, 
                                               colour="black", 
                                               fill="#adb5bd"),
                                               title="% of exp. cells",
                                               label.position="right",
                                               title.position="left",
                                               title.hjust=0.5
                                               ),
           colour = guide_colorbar(#barwidth       = unit(0.4, "cm"),
                                   #barheight      = unit(2,   "cm"),
                                   title.position = "left",
                                   title.hjust    = 0.5
                                  ))

suppressWarnings(print(p2))
```


## 4.5 Group-based differential expression in AM-CCL
### 4.5.1 Prepare object
```{r}
s <- subset(seu, subtype=="AM-CCL")
celltype <- "AM-CCL"
```

Two comparisons:
CFN (n=8) vs C (n=6)
```{r}
celltype <- "AM-CCL"
# set idents
Idents(s) <- "Condition2" # Change idents here

# define condition
test <- "CFN"
control <- "C"

compare <- paste0(test,"_vs_",control)
if(!dir.exists(here("data", "DE",celltype, compare))) {
  dir.create(here("data", "DE",celltype, compare,"DEG"), recursive = TRUE)
  dir.create(here("data", "DE",celltype, compare,"Reactome"), recursive = TRUE)
  dir.create(here("data", "DE",celltype, compare,"WikiPathway"), recursive = TRUE)
  dir.create(here("data", "DE",celltype, compare,"GO_CC"), recursive = TRUE)
  dir.create(here("data", "DE",celltype, compare,"GO_BP"), recursive = TRUE)
  dir.create(here("data", "DE",celltype, compare,"GO_MF"), recursive = TRUE)
}
```

CFB (n=8) vs C (n=6)
```{r}
celltype <- "AM-CCL"
# set idents
Idents(s) <- "Condition2" # Change idents here

# define condition
test <- "CFB"
control <- "C"

compare <- paste0(test,"_vs_",control)
if(!dir.exists(here("data", "DE",celltype, compare))) {
  dir.create(here("data", "DE",celltype, compare,"DEG"), recursive = TRUE)
  dir.create(here("data", "DE",celltype, compare,"Reactome"), recursive = TRUE)
  dir.create(here("data", "DE",celltype, compare,"WikiPathway"), recursive = TRUE)
  dir.create(here("data", "DE",celltype, compare,"GO_CC"), recursive = TRUE)
  dir.create(here("data", "DE",celltype, compare,"GO_BP"), recursive = TRUE)
  dir.create(here("data", "DE",celltype, compare,"GO_MF"), recursive = TRUE)
}
```

Two comparisons:
CFN (n=8) vs CFB (n=8)
```{r}
celltype <- "AM-CCL"
# set idents
Idents(s) <- "Condition2" # Change idents here

# define condition
test <- "CFN"
control <- "CFB"

compare <- paste0(test,"_vs_",control)
if(!dir.exists(here("data", "DE",celltype, compare))) {
  dir.create(here("data", "DE",celltype, compare,"DEG"), recursive = TRUE)
  dir.create(here("data", "DE",celltype, compare,"Reactome"), recursive = TRUE)
  dir.create(here("data", "DE",celltype, compare,"WikiPathway"), recursive = TRUE)
  dir.create(here("data", "DE",celltype, compare,"GO_CC"), recursive = TRUE)
  dir.create(here("data", "DE",celltype, compare,"GO_BP"), recursive = TRUE)
  dir.create(here("data", "DE",celltype, compare,"GO_MF"), recursive = TRUE)
}
```

### 4.5.2 DE
DE results shown in Table E6 are written by code/writeDE.R
```{r}
# condition vs healthy
markers <- FindMarkers(s,
                       ident.1=test, # condition
                       ident.2=control, # healthy
                       test.use="MAST",
                       random.seed=1990,
                       recorrect_umi=FALSE, # already run PrepSCTFindMarkers()
                       min.pct=0.1,
                       logfc.threshold = 0.25,
                       only.pos = FALSE)

markers$gene <- rownames(markers)

# get sig up- and down-regulated genes
deg <- markers %>% dplyr::filter(p_val_adj < 0.05)

up <- deg %>% dplyr::filter(p_val_adj < 0.05 & avg_log2FC > 0.25) %>% dplyr::arrange(desc(avg_log2FC))
down <- deg %>% dplyr::filter(p_val_adj < 0.05 & avg_log2FC < -0.25) %>% dplyr::arrange(avg_log2FC)

# write results
write.csv(up %>%
            dplyr::relocate(gene),
            file = here("data","DE",celltype,compare,"DEG", paste0(celltype,".",compare,".DEG-up.csv")),
            row.names = FALSE)
write.csv(down %>%
            dplyr::relocate(gene),
            file = here("data","DE",celltype,compare,"DEG", paste0(celltype,".",compare,".DEG-down.csv")),
            row.names = FALSE)
write.csv(deg %>%
            dplyr::relocate(gene),
            file = here("data","DE",celltype,compare,"DEG", paste0(celltype,".",compare,".DEG-all.csv")),
            row.names = FALSE)
write.csv(markers %>%
            dplyr::relocate(gene),
            file = here("data","DE",celltype,compare,"DEG", paste0(celltype,".",compare,".DEG-raw.csv")),
            row.names = FALSE)
```

### 4.5.3 GO enrichment
Top 10 results shown in Table E11 are written by code/goEnrichment.R
```{r}
# get ensemblDB
getCols <- setNames(c("SYMBOL","ENTREZID"),c("SYMBOL","ENTREZID"))

genes <- data.frame(
  lapply(getCols, function(column) {
    mapIds(
      x = org.Hs.eg.db,
      keys = unique(deg$gene),
      keytype = "SYMBOL",
      column = column)
  }),
  row.names = unique(deg$gene))

# GO BP enrichment
go.up.bp <- enrichGO(genes[up$gene,]$ENTREZID, OrgDb=org.Hs.eg.db, ont="BP",
                       pAdjustMethod = "BH", readable=TRUE, pvalueCutoff = 0.05)

go.down.bp <- enrichGO(genes[down$gene,]$ENTREZID, OrgDb=org.Hs.eg.db, ont="BP",
                     pAdjustMethod = "BH", readable=TRUE, pvalueCutoff = 0.05)

# Remove redundant GO terms
go.up.bp <- clusterProfiler::simplify(go.up.bp)
go.down.bp <- clusterProfiler::simplify(go.down.bp)

# write results
out <- here("data","GOBP","ruv",celltype,compare)
if (!dir.exists(out)) {
  dir.create(here("data","GOBP","ruv",celltype,compare),recursive = TRUE)
}

go.up.bp@result %>%
  dplyr::rename(Pathway=Description) %>%
  dplyr::mutate(Celltype = "AM-CCL") %>%
  dplyr::mutate(Comparison=compare) %>%
  dplyr::mutate(Direction="Up") %>%
  dplyr::mutate(Genes=gsub("/","; ",go.up.bp@result$geneID)) %>%
  dplyr::select(Pathway, Celltype, Comparison, Direction, ID, Count, pvalue, p.adjust, Genes) %>%
  slice_head(n = 10) -> up.df
write_csv(up.df, file = here("data","GOBP","ruv",celltype,compare,
                        paste0(celltype,".",compare,".GO_BP-up.csv")))

go.down.bp@result %>%
  dplyr::rename(Pathway=Description) %>%
  dplyr::mutate(Celltype = "AM-CCL") %>%
  dplyr::mutate(Comparison=compare) %>%
  dplyr::mutate(Direction="Down") %>%
  dplyr::mutate(Genes=gsub("/","; ",go.down.bp@result$geneID)) %>%
  dplyr::select(Pathway, Celltype, Comparison, Direction, ID, Count, pvalue, p.adjust, Genes) %>%
  slice_head(n = 10) -> down.df
write_csv(down.df, file = here("data","GOBP","ruv",celltype,compare,
                        paste0(celltype,".",compare,".GO_BP-down.csv")))
```

### 4.5.4 Dotplot of AM-CCL markers
```{r fig.height=2.5, fig.width=8}
# DotPlot of TRM-CCL
genes <- c("EREG",
           "CCL2","CCL8",
           "CXCL1","CXCL2","CXCL3","CXCL8",
           "IL17RB","IL1A","IL6",
           "TNFAIP3","TNF","ZFP36","TRIB1",
           "EGR1","STAT4","CD274",
           "FOS","JUN","NFKBIA","NFKBIZ","NFKB1")

pt1 <- DotPlot(s,
            features = genes,
            group.by = "Condition2",
            #cols=c("#B2182B","#F7F7F7","#2166AC"),
            col.max = 1,
            col.min = -1,
            dot.scale = 6
            ) + #coord_flip() +
    geom_point(aes(size=pct.exp), shape = 21, colour="black", stroke=0.5) +
    FontSize(y.text = 10, x.text = 8) + 
    labs(y = element_blank(), x = element_blank()) + 
      theme_bw() +
    theme(
          axis.text.x = element_text(size=10,
                                     face="italic",
                                     angle = 45,
                                     hjust = 1,
                                     vjust = 1),
          axis.text.y = element_text(size=10,hjust=0),
          legend.direction="vertical",
          legend.position="right",
          legend.title.position = "left",
          legend.justification = "bottom",
          legend.box="horizontal",
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 8,angle=90),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank()
          ) +
      
    scale_color_gradientn(colours = rev(brewer.rdbu(11)[2:10]),name = "Mean Expression") +
    scale_size_continuous(limits = c(0, 100),
                          breaks = c(0, 25, 50, 75, 100),
                          labels=c(0,"",50,"",100)) +
    guides(size=guide_legend(override.aes=list(shape=21, 
                                               colour="black", 
                                               fill="#adb5bd"),
                                               title="% of exp. cells",
                                               label.position="right",
                                               title.position="left",
                                               title.hjust=0.5
                                               ),
           colour = guide_colorbar(#barwidth       = unit(0.4, "cm"),
                                   #barheight      = unit(2,   "cm"),
                                   title.position = "left",
                                   title.hjust    = 0.5
                                  ))
pt1

```

## 4.6 PHATE analysis of AM-CCL cluster
To further capture local transcriptional heterogeneity between groups within TRM-CCL cluster, we inferred trajectory from cell embeddings of PHATE, an unsupervised dimensionality reduction method which performs well on preserving the local and global structure of highly complex datasets.

### 4.6.1 PHATE embeddings
```{r fig.height=3, fig.width=8}
s <- subset(seu, subtype=="AM-CCL")
Idents(s) <- "Condition2"

# split object
obj.list <- SplitObject(s, split.by="Condition2")
obj.list <- obj.list[levels(s$Condition2)]

# check number of cells per Subject
lapply(obj.list,ncol)

# Downsample to the minimum number of cells among HC, CFN and CFB
# This ensures the PHATE project is not biased by dominated cell number in any group
set.seed(1990)
ds.obj.list <- lapply(X = obj.list, FUN = function(x) {
  n <- min(ncol(obj.list[[1]]),ncol(obj.list[[2]]),ncol(obj.list[[3]]))
  x <- subset(x, downsample=n)
})

ds.obj.list <- ds.obj.list[levels(s)]
lapply(ds.obj.list,ncol)

ds.cells <- lapply(X=ds.obj.list, FUN=function(x) {colnames(x)})
lapply(ds.cells,length)

s <- merge(x=ds.obj.list[[1]], y=ds.obj.list[2:length(ds.obj.list)])
s$Condition2 <- factor(s$Condition2, levels=levels(seu$Condition2))

# PHATE
var.features <- SelectIntegrationFeatures(ds.obj.list,nfeatures=3000)
phate_data <- t(as.data.frame(s@assays$SCT@data)[var.features,])
reticulate::py_config()
phate_output <- phate(phate_data,
                      knn=50,
                      npca=50,
                      t=50,
                      n.jobs = 1,
                      seed=1990
                      )

embeddings <- phate_output$embedding

s[["PHATE"]] <- CreateDimReducObject(embeddings = embeddings,
                                     key = "PHATE_",
                                     assay = DefaultAssay(s))
#s@reductions$PHATE@cell.embeddings[,2] <- -s@reductions$PHATE@cell.embeddings[,2]

# UMAP
# FindNeighbors
s <- FindNeighbors(s,reduction = "PHATE", dims = 1:2)
s <- FindClusters(s, resolution = 1, algorithm = 4, method="igraph",random.seed = 1990)  #method="matrix"
#s@reductions$PHATE@cell.embeddings[,1] <- -s@reductions$PHATE@cell.embeddings[,1]
s@reductions$PHATE@cell.embeddings[,2] <- -s@reductions$PHATE@cell.embeddings[,2]

# grouped by Condition
p5a <- DimPlot(s,
        reduction = "PHATE",
        group.by="Condition2",
        cols=Condition2.colours,
        pt.size = 0.01,
        shuffle = TRUE,
        label = FALSE) +
  labs(title = "Condition", x = NULL, y = NULL) +
  guides(x = "none", y = "none") +
  theme(plot.title = element_text(size=12,face="plain"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        legend.text.align = 0,
        strip.text.x = element_text(face = "bold", vjust=2, size=12))

DimPlot(s,
        reduction = "PHATE",
        group.by = "SCT_snn_res.1",
        shuffle = TRUE,
        label = TRUE,
        #cols=cluster.colours,
        ) + ggtitle(label = "PHATE") +
  labs(x = NULL, y = NULL) +
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        legend.text.align = 0,
        strip.text.x = element_text(face = "bold", vjust=2, size=12)) + p5a

# Group SCT_snn_res.1 clusters into 3 cell states
# 1: 1,2,3,4,5,7,9,12,14,15,17,19
# 2: 6,8,10,11,16
# 3: 13,18,20
cellstate <- setNames(c("1","1","1","1","1",
                        "2","1","2","1","2",
                        "2","1","3","1","1",
                        "2","1","3","1","3"
                        ),
                      levels(s$SCT_snn_res.1))

cellstate <- setNames(factor(cellstate[s$SCT_snn_res.1],
                             levels=c("1","2","3")),
                      colnames(s))

s <- AddMetaData(s, cellstate, col.name = "cellstate")

```

### 4.6.2 PHATE visualization
#### 4.6.2.1 Density plot of cells grouped by Condition2
```{r fig.width=8, fig.height=3}
s$Phate.UMAP1 <- s@reductions$PHATE@cell.embeddings[,1]
s$Phate.UMAP2 <- s@reductions$PHATE@cell.embeddings[,2]
s$CellID <- rownames(s@meta.data)
s$Condition2 <- factor(s$Condition2, levels=c("C","CFB","CFN"))

get_density <- function(x, y, ...) {
  dens <- MASS::kde2d(x, y, ...)
  ix <- findInterval(x, dens$x)
  iy <- findInterval(y, dens$y)
  ii <- cbind(ix, iy)
  return(dens$z[ii])
}

dens.umap.list <- list()
loop.cells <- levels(s$Condition2)

for(i in seq_along(loop.cells)){
    i.c <- loop.cells[i]
    # change this to 1% of all cells
    i.n <- ceiling(sum(s$Condition2 %in% i.c)/100)
    if(i.n < 100){
      i.n <- 100
    }
    i.dens <- get_density(s$Phate.UMAP1[s$Condition2 %in% i.c],
                          s$Phate.UMAP2[s$Condition2 %in% i.c], n=i.n)
    
    dens.umap.list[[i.c]] <- data.frame("Condition2"=i.c,
                                        "CellID"=s$CellID[s$Condition2 %in% i.c],
                                        "Dens"=i.dens)
}

umap.dens <- do.call(rbind.data.frame, dens.umap.list)
umap.dens.merge <- base::merge(s@meta.data, umap.dens, by=c('Condition2', 'CellID'))


condition.order <- levels(s$Condition2)
title.list <- setNames(c("HC","CFB","CFN"),
                       levels(s$Condition2))
pt.list <- list()
for(x in seq_along(condition.order)){
  x.c <- condition.order[x]
  
  p <- ggplot(umap.dens.merge,
         aes(x=Phate.UMAP1, y=Phate.UMAP2)) +
    geom_scattermore(data=umap.dens.merge[, c("Phate.UMAP1", "Phate.UMAP2")],
                     colour='#adb5bd',pointsize =2) +
    geom_point_rast(data=umap.dens.merge[umap.dens.merge$Condition2 %in% x.c, ],
                    aes(colour=Dens),size=0.00000001) +
    scale_colour_gradientn(colors = brewer.oranges(7),na.value = "#6c757d") +
    labs(title=title.list[x.c], x="PHATE_1", y="PHATE_2") +
    theme_cowplot() + #NoLegend() +
    guides(colour=guide_colourbar(title="Density")) +
    theme(plot.title=element_text(size=16,hjust=0.5,face="plain"),
          legend.key.size = unit(0.2, "cm"),
          legend.direction="horizontal",
          legend.position="bottom",
          legend.justification = "right",
          axis.text=element_blank(),
          axis.title=element_blank(),
          axis.line = element_blank(),
          axis.ticks=element_blank(),
          legend.title=element_text(size=8),
          legend.text=element_blank())
    pt.list[[x.c]] <- p
}

p4 <- (pt.list[[1]] + NoLegend()) | (pt.list[[2]] | NoLegend()) | pt.list[[3]]
p4
```

#### 4.6.2.2 Coloured by Condition and Cell State
```{r fig.width=6, fig.height=3}
state.colours <- setNames(c("#05a8aa","#780116","#ff6600"),levels(s$cellstate))

p5a <- DimPlot(s,
        reduction = "PHATE",
        group.by="Condition2",
        cols=Condition2.colours,
        pt.size = 0.01,
        shuffle = TRUE,
        label = FALSE) +
  labs(title = "Condition", x = NULL, y = NULL) +
  guides(x = "none", y = "none") +
  theme(plot.title = element_text(size=14,face="plain"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        legend.text.align = 0,
        strip.text.x = element_text(face = "bold", vjust=2, size=12)) +
  guides(
    color = guide_legend(
      override.aes = list(
        size  = 4    # make the points in the legend larger
        #shape = 15    # change the legend symbol shape if you like
      )
    )
  )

p5b <- DimPlot(s,
        reduction = "PHATE",
        group.by="cellstate",
        cols=state.colours,pt.size=0.01,
        shuffle = TRUE,
        label = FALSE) +
  labs(title = "Cell state", x = NULL, y = NULL) +
  guides(x = "none", y = "none") +
  theme(plot.title = element_text(size=14,face="plain"),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        legend.text.align = 0,
        strip.text.x = element_text(face = "bold", vjust=2, size=12)) +
  guides(
    color = guide_legend(
      override.aes = list(
        size  = 4    # make the points in the legend larger
        #shape = 15    # change the legend symbol shape if you like
      )
    )
  )

p5a|p5b
```


### 4.6.3 Stacked bar plot of cell state composition
```{r fig.width=3, fig.height=3}
# Stacked by cell states
props <- getTransformedProps(clusters=s$cellstate,
                             sample=s$Condition2,
                             transform = "asin")
p6 <- props$Proportions %>%
  data.frame %>%
  #inner_join(info, by = c("sample" = "orig.ident")) %>%
  ggplot(aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  labs(y = "Proportion", x="Condition", title="") +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=10,
                                   colour = "black"),
        axis.title.x = element_blank(),
        axis.text.y = element_text(size=8,
                                   colour = "black"),
        plot.title = element_text(hjust=0.5, size=12,face="plain"),
        legend.text = element_text(size = 8),
        legend.position="none",
        legend.title = element_text(face="bold", size=8)) +
  guides(fill = guide_legend(#override.aes = list(size = 0.01),
                             keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = state.colours)

p6

```
```{r fig.width=4, fig.height=3}
# 1) Compute proportions with sample = sampleID
props <- getTransformedProps(clusters = s$cellstate,
                             sample = s$sampleID,
                             transform = "asin")

# 2) Make a sample-level metadata table (keeps unique sampleIDs with their Condition2)
# If s is a Seurat object, you can use: meta <- s@meta.data %>% distinct(sampleID, Condition2)
meta <- data.frame(sample = as.character(s$sampleID),
                   Condition2 = as.character(s$Condition2),
                   stringsAsFactors = FALSE) %>%
  distinct(sample, .keep_all = TRUE)

# 3) Prepare the df used for stat tests / plotting
df <- props$Proportions %>%
  as.data.frame() %>%
  # 'sample' column from props now contains sampleID values, join to meta to get Condition2
  left_join(meta, by = "sample") %>%
  mutate(
    clusters = factor(clusters),
    Condition2 = factor(Condition2)
  )

# (Optional) check levels
# levels(df$Condition2)
# levels(df$clusters)

# 4) Pairwise Wilcoxon rank-sum tests per cluster (BH correction) using rstatix
# set ref.group to your control label (example uses "C")
stat.test <- df %>%
  group_by(clusters) %>%
  wilcox_test(Freq ~ Condition2, ref.group = "C", p.adjust.method = "BH") %>%
  add_xy_position(x = "clusters", dodge = 0.8)

# 5) Boxplot following your format, but using state.colours as the palette for Condition2
# NOTE: ensure `state.colours` is appropriate for mapping to Condition2 (names or same order)
dodge_width <- 0.8

p5 <- ggplot(df, aes(x = clusters, y = Freq, color = Condition2)) +
  geom_boxplot(
    fill = NA,
    outlier.shape = NA,
    position = position_dodge(width = dodge_width)
  ) +
  geom_point(
    position = position_dodge(width = dodge_width),
    size = 2, alpha = 0.8
  ) +
  stat_pvalue_manual(
    stat.test,
    label = "p.adj.signif",
    tip.length = 0.01,
    bracket.size = 0.4,
    hide.ns = TRUE
  ) +
  scale_color_manual(values = Condition2.colours) +    # using state.colours as requested
  labs(y = "Proportion") +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_blank(),
    axis.text.x = element_text(angle = 40, hjust = 1, size = 10, colour = "black"),
    axis.text.y = element_text(size = 12, colour = "black"),
    axis.title.x = element_blank(),
    legend.title = element_blank(),
    legend.position = c(0.01, 0.99),
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.size = unit(1, "lines"),
    legend.text = element_text(size = 12)
  ) + NoLegend()
  # if you have a vector of labels for clusters (subtype.labels), use it:
  #scale_x_discrete(labels = subtype.labels)

p5



```

### 4.6.4 Find markers in each cell state
```{r}
# Cepo
cepoMarkers <- Cepo(s[["SCT"]]@data,
                    s$cellstate,
                    exprsPct = 0.2)
cepoMarkers$stats
cepoMarkers$stats <- cepoMarkers$stats[!grepl('^A[C|F|P|L][0|1]', row.names(cepoMarkers$stats)),]

sapply(1:ncol(cepoMarkers$stats), function(i){
  names(sort(cepoMarkers$stats[,i], decreasing = TRUE))[1:300]
}) %>% data.frame() -> dat2

colnames(dat2) <- levels(s$cellstate)
dat2 %>% write.csv(file=here("data","AM-CCL.phate.markers.csv"))

dat2

```


### 4.6.5 Plot DEGs
```{r fig.width=8, fig.height=8}
features <- c("TNFAIP3","FOS","JUNB","NFKBIA","HLA-DQA2","HLA-DRB5",
              "EREG","CCL3","CCL4","CCL4L2","CTSL","SLC11A1",
              "CXCL2","CXCL3","CXCL5","CXCL8","LILRB4","IGFBP2"
              #"TREM2","SORL1","TMEM176A","TMEM176B","TNFAIP6"
              )
# CFB
features <- c("TNFAIP3","FOS","JUNB",
              "NFKBIA","EREG","CCL3",
              "CCL4","CCL4L2","CXCL2",
              "CXCL3","CXCL5","CXCL8"
              )
p7 <- suppressMessages(FeaturePlot(s,
            reduction = "PHATE",
            pt.size = 0.001,
            ncol = 4,
            #order = TRUE,
            features = features) &
  xlim(c(min(data.frame(s@reductions$PHATE@cell.embeddings)$PHATE_1),
         max(data.frame(s@reductions$PHATE@cell.embeddings)$PHATE_1))) &
  ylim(c(min(data.frame(s@reductions$PHATE@cell.embeddings)$PHATE_2),
         max(data.frame(s@reductions$PHATE@cell.embeddings)$PHATE_2))) &
  scale_color_gradientn(colours = inferno(256)) &
    theme_minimal() &
    theme(panel.border = element_blank(), 
        panel.grid = element_blank(),
        axis.text=element_blank(),
        axis.title = element_blank(),
        plot.title = element_text(size = 12, hjust=0.5, face="bold.italic"),
        legend.key.size = unit(0.2, "cm"),
        legend.text = element_text(size = 6),
        legend.title = element_text(face="bold", size=12)))

# CFN - antigen presentation
features <- c("HLA-DQA2","HLA-DRB5","CTSL","LILRB4"#,"S100A6","S100A8","S100A10","S100A13"
              )

p8 <- suppressMessages(FeaturePlot(s,
            reduction = "PHATE",
            pt.size = 0.001,
            ncol = 4,
            #order = TRUE,
            features = features) &
  xlim(c(min(data.frame(s@reductions$PHATE@cell.embeddings)$PHATE_1),
         max(data.frame(s@reductions$PHATE@cell.embeddings)$PHATE_1))) &
  ylim(c(min(data.frame(s@reductions$PHATE@cell.embeddings)$PHATE_2),
         max(data.frame(s@reductions$PHATE@cell.embeddings)$PHATE_2))) &
  scale_color_gradientn(colours = inferno(256)) &
    theme_minimal() &
    theme(panel.border = element_blank(), 
        panel.grid = element_blank(),
        axis.text=element_blank(),
        axis.title = element_blank(),
        plot.title = element_text(size = 12, hjust=0.5, face="bold.italic"),
        legend.key.size = unit(0.2, "cm"),
        legend.text = element_text(size = 6),
        legend.title = element_text(face="bold", size=12)))

p7/p8
```

# 5 Save figures
## 5.1 AM
```{r}
# celltype <- "AM"
# out <- here("data","plots", celltype)
# if(!dir.exists(out)) {
#   dir.create(out,recursive = TRUE)
# }
# 
# # umap split by Condition2
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 1600)
# p1 +
#   theme(plot.tag = element_text(size = 14, face = "bold"))
# 
# ggsave(plot=last_plot(),
#        device="jpeg",
#        dpi=1600,
#        filename = paste0(out,"/","bal_",celltype,".umap.splitbyCondition2.jpeg"),
#        height=3.5,
#        width=10)
# 
# # selected dot plot
# ## version 1: horizontal
# suppressWarnings(print(p2 +
#   theme(plot.tag = element_text(size = 14, face = "bold"
#                                 ))))
# 
# ggsave(plot=last_plot(),
#        device="jpeg",
#        dpi=1600,
#        filename = paste0(out,"/","bal_",celltype,".dotplot.jpeg"),
#        height=7,
#        width=4)
# 
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 100)

```

## 5.2 AM-CCL
```{r}
# celltype <- "AM-CCL"
# out <- here("data","plots",celltype)
# if(!dir.exists(out)) {
#   dir.create(out,recursive = TRUE)
# }
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 1600)
# 
# # TRM-CCL violin plot
# p3 +
#   theme(plot.tag = element_text(size = 14, face = "bold"))
# ggsave(plot=last_plot(), dpi=800, device="jpeg", limitsize = FALSE,
#        width=10, height=4, file=paste0(out,"/","bal_",celltype,"plots","FigE2.jpeg"))
# 
# # PHATE plot
# p4 +
#   theme(plot.tag = element_text(size = 14, face = "bold"))
# ggsave(last_plot(),
#        device="jpeg",
#        dpi=1600,
#        filename = paste0(out,"/","AM-CCL.phate.density.jpeg"),
#        height=4,
#        width=8)
# p5a +
#   theme(plot.tag = element_text(size = 14, face = "bold"))
# ggsave(last_plot(),
#        device="jpeg",
#        dpi=1600,
#        filename = paste0(out,"/","AM-CCL.phate.condition.jpeg"),
#        height=3,
#        width=3)
# 
# p5b +
#   theme(plot.tag = element_text(size = 14, face = "bold"))
# ggsave(last_plot(),
#        device="jpeg",
#        dpi=1600,
#        filename = paste0(out,"/","AML.phate.cellstate.jpeg"),
#        height=3,
#        width=3)
# 
# # proportion bar plot
# p6 +
#   theme(plot.tag = element_text(size = 14, face = "bold"))
# ggsave(last_plot(),
#        device="jpeg",
#        dpi=1600,
#        filename = paste0(out,"/","AM-CCL.barplot.jpeg"),
#        height=2.5,
#        width=2.5)
# 
# # AM-CCL dotplot
# pt1
# ggsave(last_plot(),filename=here("data","plots","AM-CCL","overlap.dotplot2.jpeg"),
#        device="jpeg",dpi=1600,height=2.5, width=7.5)
# 
# 
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 100)
```

# 6 Save object
```{r}
celltype <- "AM"
out <- here("data","SCEs","analysis",celltype,paste0("bal_",celltype,".analysis.SEU.qs"))
if (!file.exists(out)) {
  qsave(seu,file=out,nthreads=24)
}

celltype <- "AM-CCL"
out <- here("data","SCEs","analysis",celltype,paste0("bal_",celltype,".analysis.SEU.qs"))
if (!file.exists(out)) {
  qsave(s,file=out,nthreads=16)
}

```

# 7 Save annotation
```{r}
# 1) Write out annotations
out_dir  <- here("output")
out_file <- file.path(out_dir, "bal_AM_annotations.csv")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

ann <- tibble(
  cell      = rownames(seu@meta.data),
  sampleID  = seu$sampleID,
  subtype   = as.character(seu$celltype)
)

readr::write_csv(ann, out_file)
out_file

```
