---
title: 'Analysis of Proliferating macrophages (Prolif M)'
author:
- name: Anson Wong
date: "`r Sys.Date()`"
description: null
output: html_document
editor_options:
  chunk_output_type: inline
---

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
xaringanExtra::use_panelset()

```

# 1 Load packages
```{r}
suppressPackageStartupMessages({
library(Seurat)
library(here)
library(qs)
library(forcats)
library(pals)
library(ggplot2)
library(ggpubr)
library(speckle)
library(rstatix)
library(scCustomize)
library(patchwork)
library(tidyverse)
library(dplyr)
library(Cepo)
library(phateR)
library(slingshot)
library(tradeSeq)
library(org.Hs.eg.db)
library(clusterProfiler)
library(ReactomePA)
library(readr)
library(cowplot)
library(harmony)
library(scattermore)
library(ggrastr)
library(viridis)
library(scater)
library(circlize)
library(ggplotify)
library(ggh4x)
library(tidyr)
})
set.seed(1990)

```

# 2 Annotation, integration, clustering, finding marker genes
Cells annotated as TRM, Prolif M and RecM in the combined dataset were extracted for subclustering.
RecM cluster was included for subclustering to aid a better TRM projection in UMAP, but they are not included for analysis here.
Methods for normalization, integration, clustering and annotation were the same.

For details, refer to code/bal_TRM.integration_clustering.R

# 4 Analysis
## 4.1 Read processed object
```{r}
out <- here("data","SCEs","analysis","Prolif_M.analysis.SEU.qs")

if (!file.exists(out)) {
  seu <- qread(here("data","SCEs","analysis","bal.SEU.qs"),nthreads=16)
  s <- subset(seu, celltype=="Prolif M")
  s$celltype <- fct_drop(s$celltype)
  
  DefaultAssay(s) <- "RNA"
  
  # split object by experiment
  obj.list <- SplitObject(s, split.by="batchID")
  
  # normalization
  obj.list <- lapply(X = obj.list, FUN = function(x) {
    x <- SCTransform(x, vst.flavor = "v2", verbose=TRUE)
  })
  
  # find most variable features across samples to integrate
  var.features <- SelectIntegrationFeatures(object.list = obj.list, nfeatures=3000)
  
  # merge normalized samples
  s <- merge(obj.list[[1]], y=obj.list[2:length(obj.list)], merge.data=TRUE)
  
  DefaultAssay(s) <- "SCT"
  VariableFeatures(s) <- var.features
  
  # manually set variable features of merged Seurat object
  s <- RunPCA(s)
  s <- RunHarmony(s,
                  group.by.vars=c("batchID"),
                  reduction="pca",
                  assay.use="SCT",
                  reduction.save="harmony",
                  plot_convergence=TRUE,
                  .options = harmony_options(max.iter.cluster=50),
                  max_iter=100)
  s <- FindNeighbors(s, reduction="harmony", dims=1:30)
  s <- FindClusters(s, resolution=1, algorithm=4, method="igraph")
  s <- RunUMAP(s, reduction="harmony", dims=1:30)
  
  # cell cycle scoring
  s <- CellCycleScoring(s, 
                        g2m.features = cc.genes.updated.2019$g2m.genes, 
                        s.features = cc.genes.updated.2019$s.genes)
  qsave(s, file=out, nthreads=16)

} else {
  s <- qread(out, nthreads=16)
  s$cc.phase <- ifelse(s$Phase == "G1", "G1", "S-G2M")
  DefaultAssay(s) <- "RNA"
  s <- NormalizeData(s)
}


# 1) Recode Condition2
s$Condition2 <- recode(s$Condition2,
  "HC"="C",    # rename HC → C
  "CFB"="CFB",
  "CFN"="CFN"
)
# now turn into a factor with your desired order
s$Condition2 <- factor(s$Condition2, levels = c("C","CFB","CFN"))

s$Condition <- recode(s$Condition,
  "HC"="C",    # rename HC → C
  "CF"="CF"
)
# now turn into a factor with your desired order
s$Condition <- factor(s$Condition, levels = c("C","CF"))

# 2) Rewrite ID strings
s$ID <- s$ID %>%
  # HC-01 → C-1, HC-02 → C-2, etc
  sub("^HC-0*", "C-", .) %>%
  # CF-B1 → CFB-1, CF-B2 → CFB-2, etc
  sub("^CF-B", "CFB-", .) %>%
  # CF-N1 → CFN-1, CF-N2 → CFN-2, etc
  sub("^CF-N", "CFN-", .)

# now set the exact levels you want
new_id_levels <- c(
  paste0("C-",   1:6),
  paste0("CFB-", 1:8),
  paste0("CFN-", 1:8)
)
s$ID <- factor(s$ID, levels = new_id_levels)


```

## 4.2 UMAP
### 4.2.1 all UMAPs
```{r fig.height=6, fig.width=8}
Condition2.colours <- setNames(c("#26547c","#ef476f","#ffd166"),
                               levels(s$Condition2))
Condition.colours <- setNames(c("#26547c","#fb8b24"),
                              levels(s$Condition))

phase.colours <- setNames(c("#dd6e42","#e8dab2"),
                          levels(s$cc.phase))

(DimPlot(s, group.by = "Condition2", cols=Condition2.colours) |
  DimPlot(s, group.by = "cc.phase", cols=phase.colours)) /
  (DimPlot(s, group.by = "Condition",cols=Condition.colours) |
  DimPlot(s, group.by = "SCT_snn_res.1", cols = brewer.dark2(10)))
```

### 4.2.2 UMAP coloured by phase
```{r fig.height=3, fig.width=8}
DimPlot(s, 
        reduction = 'umap',
        pt.size = 0.5,
        label.box = FALSE,
        label = FALSE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 4, 
        cols=phase.colours,
        group.by = "cc.phase") + #NoLegend() +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text.align = 0,
        legend.text = element_text(size = 12),
        strip.text.x = element_text(face = "bold", vjust=2, size=16)) +
  scale_color_manual(values = phase.colours) +
  guides(
    color = guide_legend(
      override.aes = list(
        size  = 5,    # make the points in the legend larger
        shape = 15    # change the legend symbol shape if you like
      )
    )
  ) -> pp1


DimPlot(s, 
        reduction = 'umap',
        pt.size = 0.5,
        label.box = FALSE,
        label = FALSE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 4, 
        cols=Condition2.colours,
        group.by = "Condition2") + #NoLegend() +
  labs(x = NULL, y = NULL) + 
  guides(x = "none", y = "none") +
  theme(plot.title = element_blank(),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text.align = 0,
        legend.text = element_text(size = 12),
        strip.text.x = element_text(face = "bold", vjust=2, size=16)) +
  scale_color_manual(values = Condition2.colours) +
  guides(
    color = guide_legend(
      override.aes = list(
        size  = 5,    # make the points in the legend larger
        shape = 15    # change the legend symbol shape if you like
      )
    )
  ) -> pp2

pp1|pp2 -> p3
p3
```

## 4.3 RecM vs AM signature
### 4.3.1 Add module score
```{r fig.width=5, fig.height=3}
markers <- read.csv(here("data","Major.markers.csv"))

am.genes   <- markers$AM[1:300]
recm.genes <- markers$RecM[1:300]

# 3. Add module scores
s <- AddModuleScore(s,
  features = list(AM = am.genes, RecM = recm.genes),
  name     = c("AM_score", "RecM_score")
)

# → creates metadata columns "AM_score1" and "RecM_score1"

# 3. pull out those scores
scores_df <- FetchData(
  object = s,
  vars   = c("ID", "Condition2", "AM_score1", "RecM_score2")
) %>%
  tibble::rownames_to_column("cell") %>%
  dplyr::rename(
    AM_score   = AM_score1,
    RecM_score = RecM_score2
  )

# 2. Pivot to long form
long_df <- scores_df %>%
  tidyr::pivot_longer(
    cols      = c(AM_score, RecM_score),
    names_to  = "signature",
    values_to = "module_score"
  )

# 3. Pseudobulk by taking the mean per sample (ID) & signature
pseudo_df <- long_df %>%
  group_by(ID, Condition2, signature) %>%
  summarise(
    module_score = mean(module_score),
    .groups      = "drop"
  )

```

### 4.3.2 AM vs RecM signature within each group
```{r fig.width=5, fig.height=2}
# 1) Keep only IDs with both signatures
pair_counts <- pseudo_df %>%
  group_by(Condition2, ID) %>%
  summarise(n = n(), .groups = "drop")

pseudo_df2 <- pseudo_df %>%
  semi_join(
    filter(pair_counts, n == 2),
    by = c("Condition2", "ID")
  )

pseudo_df2$signature <- factor(pseudo_df2$signature,
                               levels = c("AM_score", "RecM_score"))

pseudo_df2$Condition2 <- factor(pseudo_df2$Condition2,
                               levels = c("C", "CFB", "CFN"))

# 2) Run paired t‑tests per Condition2
stat_df <- pseudo_df2 %>%
  group_by(Condition2) %>%
  pairwise_t_test(
    module_score ~ signature,
    paired           = TRUE,
    id               = "ID",
    p.adjust.method  = "BH"
  ) %>%
  add_xy_position(x = "signature", dodge = 0.8)

# 3) Compute y‑positions for each Condition2
lims <- pseudo_df2 %>%
  group_by(Condition2) %>%
  summarise(
    y_max = max(module_score, na.rm = TRUE),
    y_min = min(module_score, na.rm = TRUE),
    .groups = "drop"
  )

stat_df_wilcox <- pseudo_df2 %>%
  group_by(Condition2) %>%
  pairwise_wilcox_test(
    module_score ~ signature,
    paired           = TRUE,
    id               = "ID",
    p.adjust.method  = "BH"
  ) %>%
  add_xy_position(x = "signature", dodge = 0.8)

# then plot exactly as before, but using the Wilcoxon stats
ggplot(pseudo_df2, aes(x = signature, y = module_score, group = ID)) +
  geom_line(color = "grey50", size = 0.3, alpha = 0.7) +
  geom_point(aes(colour = signature), size = 1.5) +
  facet_wrap(~ Condition2) +
  stat_pvalue_manual(
    stat_df_wilcox,
    label        = "p.adj.signif",
    tip.length   = 0.03,
    bracket.size = 0.4,
    hide.ns      = TRUE,
    bracket.nudge.y = 0.02
  ) +
  scale_x_discrete(labels=c("AM","RecM")) +
  scale_colour_manual(
    name   = "Signature",
    values = c(AM_score   = "#054a91",
               RecM_score = "#dda15e"),
    labels = c("AM", "RecM")
  ) +
  labs(x = NULL, y = "Module score") +
  coord_cartesian(ylim = c(-0.1, 0.1)) +
  theme_classic(base_size = 12) +
  theme(
    strip.text       = element_text(size = 10),
    axis.title.x     = element_blank(),
    axis.title.y     = element_text(size = 10),
    axis.text.x      = element_text(size = 10),
    axis.text.y      = element_text(size = 8),
    axis.ticks.x     = element_blank(),
    panel.border     = element_rect(color = "black", fill = NA, size = 0.5),
    strip.background = element_rect(color = "black", fill = "white", size = 0.5),
    axis.line.x      = element_blank(),
    axis.line.y      = element_blank(),
    legend.position  = "right",
    legend.title     = element_blank(),
    legend.text      = element_text(size = 10)
  ) + NoLegend()
```

### 4.3.3 AM vs RecM signature within each group, separated by Cell Cycle Phase
```{r fig.width=4, fig.height=3}
# 1) Fetch & pseudobulk by ID × Condition2 × cc.phase × signature
pseudo_df <- FetchData(
  object = s,
  vars   = c("ID", "Condition2", "cc.phase", "AM_score1", "RecM_score2")
) %>%
  rownames_to_column("cell") %>%
  dplyr::rename(
    AM_score   = AM_score1,
    RecM_score = RecM_score2
  ) %>%
  pivot_longer(
    cols      = c(AM_score, RecM_score),
    names_to  = "signature",
    values_to = "module_score"
  ) %>%
  group_by(ID, Condition2, cc.phase, signature) %>%
  summarise(module_score = mean(module_score), .groups = "drop") %>%
  # enforce factor order
  mutate(
    signature = factor(signature, levels = c("AM_score","RecM_score")),
    Condition2 = factor(Condition2, levels = c("C","CFB","CFN")),
    cc.phase   = factor(cc.phase,   levels = c("G1","S-G2M"))
  )

# 2) Paired Wilcoxon per phase × condition, BH‐adjusted
stat_df <- pseudo_df %>%
  group_by(cc.phase, Condition2) %>%
  wilcox_test(
    module_score ~ signature,
    paired = TRUE
  ) %>%
  adjust_pvalue(method="BH", p.col="p", output.col="p.adj") %>%
  add_significance("p.adj") %>%               # gives p.adj.signif
  add_xy_position(x="signature", dodge = 0)    # dodge=0 for aligned

# 3) Compute y‑positions by phase × condition
lims <- pseudo_df %>%
  group_by(cc.phase, Condition2) %>%
  summarise(
    y_max = max(module_score, na.rm = TRUE),
    y_min = min(module_score, na.rm = TRUE),
    .groups = "drop"
  )

stat_df <- stat_df %>%
  left_join(lims, by = c("cc.phase","Condition2")) %>%
  group_by(cc.phase, Condition2) %>%
  mutate(
    y.position = y_max + (row_number()-1) * (y_max-y_min) * 0.1
  ) %>%
  ungroup()

# 4) Plot: only lines + points, aligned, faceted by phase (rows) × condition (cols)
ggplot(pseudo_df, aes(
    x     = signature,
    y     = module_score,
    group = ID
  )) +
  # grey connecting lines
  geom_line(
    color = "grey50",
    size  = 0.3,
    alpha = 0.7
  ) +
  # coloured dots
  geom_point(
    aes(colour = signature),
    size = 2
  ) +
  # facets: phase × condition
  ggh4x::facet_grid2(
    cc.phase ~ Condition2,
    strip = ggh4x::strip_themed(
      # column strips (Condition2)…
      background_x = elem_list_rect(
        fill = Condition2.colours
      ),
      # row strips (cc.phase)…
      background_y = elem_list_rect(
        fill = c("G1"    = "#dd6e42",
                 "S-G2M" = "#e8dab2")
      ),
      # map each element to its own strip, not to layers
      by_layer_x = FALSE,
      by_layer_y = FALSE
    )
  ) +
  # p‑value brackets
  stat_pvalue_manual(
    stat_df,
    label        = "p.adj.signif",
    tip.length   = 0.03,
    bracket.size = 0.4,
    hide.ns = TRUE,
    bracket.nudge.y = 0.03,
    vjust = 0
  ) +
  # manual colours & labels
  scale_colour_manual(
    name   = "Signature",
    values = c(AM_score   = "#054a91",
               RecM_score = "#dda15e"),
    labels = c("AM","RecM")
  ) +
  scale_x_discrete(labels=c("AM","RecM")) +
  labs(x = NULL, y = "Score") +
  coord_cartesian(ylim = c(-0.1, 0.12)) +
  theme_bw(base_size = 12) +
  guides(
    colour = guide_legend(
      override.aes = list(size =3)   # points in legend
    )
  ) +
  theme(
    strip.text       = element_text(size = 10, colour="white", face="bold"),
    axis.title.x     = element_blank(),
    axis.title.y     = element_text(size=10),
    axis.text.x      = element_text(size=10),
    axis.text.y      = element_text(size=8),
    #axis.ticks.x     = element_blank(),
    panel.border     = element_rect(color = "black", fill = NA, size = 0.5),
    strip.background = element_rect(color = "black", fill = "white", size = 0.5),
    axis.line.x      = element_blank(),
    axis.line.y      = element_blank(),
    legend.position  = "right",
    legend.title     = element_text(size=11),
    legend.text      = element_text(size=10)
  ) + NoLegend() -> p2
p2
```

## 4.4 Other plots
### 4.4.1 feature plot of markers
```{r fig.width=9, fig.height=9}
FeaturePlot(s, features = c("SPP1","CHI3L1","MERTK","MMP9","PLA2G7","GPNMB","TGFBI","CD36",
                            "MMP19"), 
            #order = TRUE,
            cols=brewer.reds(9)[2:9])
```

### 4.4.2 Kernal density of cell cycle phase
```{r fig.height=2, fig.width=4}
s$Harmony.UMAP1 <- s@reductions$umap@cell.embeddings[,1]
s$Harmony.UMAP2 <- s@reductions$umap@cell.embeddings[,2]
s$CellID <- rownames(s@meta.data)

get_density <- function(x, y, ...) {
  dens <- MASS::kde2d(x, y, ...)
  ix <- findInterval(x, dens$x)
  iy <- findInterval(y, dens$y)
  ii <- cbind(ix, iy)
  return(dens$z[ii])
}


dens.umap.list <- list()
loop.cells <- levels(factor(s$cc.phase))

for(i in seq_along(loop.cells)){
    i.c <- loop.cells[i]
    # change this to 1% of all cells
    i.n <- ceiling(sum(s$cc.phase %in% i.c)/100)
    if(i.n < 100){
      i.n <- 100
    }
    i.dens <- get_density(s$Harmony.UMAP1[s$cc.phase %in% i.c],
                          s$Harmony.UMAP2[s$cc.phase %in% i.c], n=i.n)
    
    dens.umap.list[[i.c]] <- data.frame("cc.phase"=i.c,
                                        "CellID"=s$CellID[s$cc.phase %in% i.c],
                                        "Dens"=i.dens)
}

umap.dens <- do.call(rbind.data.frame, dens.umap.list)
umap.dens.merge <- base::merge(s@meta.data, umap.dens, by=c('cc.phase', 'CellID'))


phase.order <- c("G1","S-G2M")
title.list <- setNames(phase.order,
                       phase.order)
pt.list <- list()
for(x in seq_along(phase.order)){
  x.c <- phase.order[x]
  
  p <- ggplot(umap.dens.merge,
         aes(x=Harmony.UMAP1, y=Harmony.UMAP2)) +
    geom_scattermore(data=umap.dens.merge[, c("Harmony.UMAP1", "Harmony.UMAP2")],
                     colour='#adb5bd',) +
    geom_point_rast(data=umap.dens.merge[umap.dens.merge$cc.phase %in% x.c, ],
                    aes(colour=Dens), size=0.001) +
    scale_colour_viridis() +
    labs(title=title.list[x.c], x="UMAP1", y="UMAP2") +
    theme_cowplot() +
    guides(colour=guide_colourbar(title="Density")) +
    theme(plot.title=element_text(size=8,hjust=0.5,face="plain"),
          legend.key.size = unit(0.2, "cm"),
          axis.text=element_blank(),
          axis.title=element_blank(),
          axis.line = element_blank(),
          axis.ticks=element_blank(),
          legend.title=element_text(size=6),
          legend.text=element_blank())
    pt.list[[x.c]] <- p
}


p4 <- pt.list[[1]] + pt.list[[2]]
p4

```

### 4.4.3 Kernal density of marker genes
#### 4.4.3.1 Define function
```{r}
# helper to compute 2D density at each (x,y)
get_density <- function(x, y, n = 100, ...) {
  dens <- MASS::kde2d(x, y, n = n, ...)
  ix   <- findInterval(x, dens$x)
  iy   <- findInterval(y, dens$y)
  dens$z[cbind(ix, iy)]
}

plot_marker_density <- function(
  seurat_obj,
  markers,
  umap1 = "Harmony.UMAP1",
  umap2 = "Harmony.UMAP2",
  cellid = "CellID",
  min_grid = 100
) {
  # 0) pull & augment metadata
  meta <- seurat_obj@meta.data
  if (!cellid %in% colnames(meta)) {
    meta[[cellid]] <- rownames(meta)
  }
  
  # 1) add “_pos” columns for each gene
  for (g in markers) {
    expr_vals <- FetchData(seurat_obj, vars = g)[[g]]
    meta[[paste0(g, "_pos")]] <- expr_vals > 0
  }
  
  # 2) compute densities per marker
  dens_list <- vector("list", length(markers))
  names(dens_list) <- markers
  
  for (g in markers) {
    flag_col <- paste0(g, "_pos")
    cells    <- which(meta[[flag_col]])
    grid_n   <- max(ceiling(length(cells) / min_grid), min_grid)
    
    dens_vals <- get_density(
      meta[[umap1]][cells],
      meta[[umap2]][cells],
      n = grid_n
    )
    
    dens_list[[g]] <- data.frame(
      CellID = meta[[cellid]][cells],
      Marker = g,
      Dens   = dens_vals,
      stringsAsFactors = FALSE
    )
  }
  
  dens_df <- do.call(rbind, dens_list)
  
  # merge back onto metadata
  meta2 <- merge(meta, dens_df, by = cellid)
  
  # 3) prepare bare‐UMAP background
  umap_all <- data.frame(
    Harmony.UMAP1 = meta[[umap1]],
    Harmony.UMAP2 = meta[[umap2]]
  )
  
  # 4) build label expressions, e.g. italic(GENE)+
  lab_expr <- lapply(markers, function(m)
    bquote(italic(.(m))^"+")
  )
  names(lab_expr) <- markers
  
  # 5) loop to make one plot per marker
  plots <- lapply(markers, function(g) {
    ggplot() +
      # all cells in grey
      geom_scattermore(
        data   = umap_all,
        aes(x = Harmony.UMAP1, y = Harmony.UMAP2),
        colour = "#adb5bd"
      ) +
      # positives coloured by density
      geom_point_rast(
        data = meta2[meta2$Marker == g, ],
        aes_string(x = umap1, y = umap2, colour = "Dens"),
        size = 0.001
      ) +
      scale_colour_viridis() +
      labs(title = lab_expr[[g]], x = "UMAP1", y = "UMAP2") +
      theme_cowplot() +
      theme(
        plot.title      = element_text(size = 12, hjust = 0.5, face = "plain"),
        legend.key.size = unit(0.2, "cm"),
        axis.text       = element_blank(),
        axis.title      = element_blank(),
        axis.line       = element_blank(),
        axis.ticks      = element_blank(),
        legend.title    = element_text(size = 6),
        legend.text     = element_blank()
      )
  })
  
  # name the list by gene for easy indexing
  names(plots) <- markers
  return(plots)
}


```

#### 4.4.3.2 Plot
```{r fig.width=6, fig.height=3.5}
markers <- c("VCAN","FABP4","CCL2","CXCL8","SPP1","PLA2G7")
pt.list <- plot_marker_density(s, markers)

(pt.list[["VCAN"]] + pt.list[["FABP4"]] + pt.list[["CCL2"]]) / 
  (pt.list[["CXCL8"]] + pt.list[["SPP1"]] + pt.list[["PLA2G7"]]) -> p4

p4
```


# 5 Save figures
```{r}
celltype <- "ProlifM"
out <- here("output","plots",celltype)
if(!dir.exists(out)) {
  dir.create(out,recursive = TRUE)
}

showtext::showtext_auto()
showtext::showtext_opts(dpi = 1600)


# # cc phase proportion
# p1 +
#   theme(plot.tag = element_text(size = 14, face = "bold"))
# 
# ggsave(plot=last_plot(),
#        device="jpeg",
#        dpi=1600,
#        filename = paste0(out,"/","bal_",celltype,".ccphase.prop.jpeg"),
#        height=2,
#        width=6)

# paired T test of RecM signature
p2 +
  theme(plot.tag = element_text(size = 14, face = "bold"))

ggsave(plot=last_plot(),
       device="jpeg",
       dpi=1600,
       filename = paste0(out,"/","bal_",celltype,".signature_pairedTtest.jpeg"),
       height=3,
       width=4)

# umap
p3 +
  theme(plot.tag = element_text(size = 14, face = "bold"))

ggsave(plot=last_plot(),
       device="jpeg",
       dpi=1600,
       filename = paste0(out,"/","bal_",celltype,".umap.jpeg"),
       height=3,
       width=8)

# density plot
p4 +
  theme(plot.tag = element_text(size = 14, face = "bold"))

ggsave(plot=last_plot(),
       device="jpeg",
       dpi=1600,
       filename = paste0(out,"/","bal_",celltype,".markerDensity.jpeg"),
       height=3.5,
       width=6)

# # dot plot
# p5 +
#   theme(plot.tag = element_text(size = 14, face = "bold"))
# 
# ggsave(plot=last_plot(),
#        device="jpeg",
#        dpi=1600,
#        filename = paste0(out,"/","bal_",celltype,".DE.dotplot.jpeg"),
#        height=2.5,
#        width=8)
# 
# p5 + NoLegend() +
#   theme(plot.tag = element_text(size = 14, face = "bold"))
# 
# ggsave(plot=last_plot(),
#        device="jpeg",
#        dpi=1600,
#        filename = paste0(out,"/","bal_",celltype,".DE.dotplot_NoLegend.jpeg"),
#        height=1.8,
#        width=6)

showtext::showtext_auto()
showtext::showtext_opts(dpi = 100)

```

# 6 Save raw data
```{r}
# correlation to pc1
stat_df %>% as.data.frame() %>% WriteXLS::WriteXLS(here("output","ProlifM_module_stat.xlsx"))

write_summary_xlsx <- function(pseudo_df, file = "cyt_summary.xlsx") {
  library(dplyr)
  library(writexl)

  out <- pseudo_df %>%
    # remove suffix so "AM_score" -> "AM", "RecM_score" -> "RecM"
    mutate(signature = sub("_score$", "", signature)) %>%
    group_by(cc.phase, Condition2, signature) %>%
    summarise(
      `Min value`    = quantile(module_score, probs = 0,    na.rm = TRUE),
      `1st Quatile`  = quantile(module_score, probs = 0.25, na.rm = TRUE),
      `Median`       = quantile(module_score, probs = 0.5,  na.rm = TRUE),
      `3rd Quatile`  = quantile(module_score, probs = 0.75, na.rm = TRUE),
      `Max value`    = quantile(module_score, probs = 1,    na.rm = TRUE),
      .groups = "drop"
    ) %>%
    # rename to the exact column headers you requested
    rename(
      `Cell cycle phase` = cc.phase,
      `Study group`      = Condition2,
      `Gene set`         = signature
    )

  writexl::write_xlsx(list(Summary = out), path = file)
  invisible(out)
}

write_summary_xlsx(pseudo_df = pseudo_df,file = here("output","ProlifM_module.xlsx"))

pseudo_df %>% data.frame() %>% write_csv(here("output","Fig5B.csv"))
```
