---
title: 'ETI single-cell analysis'
author:
- name: Anson Wong
date: "`r Sys.Date()`"
description: null
output: html_document
editor_options:
  chunk_output_type: inline
---

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
xaringanExtra::use_panelset()

```

# 1 Load packages
```{r}
suppressPackageStartupMessages({
library(Seurat)
library(here)
library(qs)
library(forcats)
library(pals)
library(ggplot2)
library(ggpubr)
library(cowplot)
library(speckle)
library(rstatix)
library(scCustomize)
library(patchwork)
library(dplyr)
library(Cepo)
library(phateR)
library(slingshot)
library(tradeSeq)
library(org.Hs.eg.db)
library(clusterProfiler)
library(ReactomePA)
library(readr)
library(DOSE)
library(parallelDist)
library(ggrepel)
library(tibble)
})
set.seed(1990)

```

# 2 Analysis
Annotation was performed on longitudinal samples.

The sample size per condition for this analysis are as follows:
n=3 CF.UT.BL
n=3 CF.UT
n=3 CF.ETI.BL
n=3 CF.ETI

## 2.1 Read processed object
RecM
```{r}
# load analysis object
celltype <- "RecM"
seu <- qread(here("data","SCEs","final.annot",paste0("bal_",celltype,".final_annot.SEU.qs")),
                  nthreads=16)
latent.vars <- c("Bronchiectasis")
```

AM-CCL
```{r}
# load analysis object
celltype <- "TRM-CCL"
seu <- qread(here("data","SCEs","final.annot",paste0("bal_",celltype,".final_annot.SEU.qs")),
                  nthreads=16)
latent.vars <- c("Bronchiectasis")
```

Combined
```{r}
# load analysis object
seu <- qread(here("data","SCEs","manual.annot",paste0("bal.SEU.qs")),
                  nthreads=24)
latent.vars <- c("Bronchiectasis")
seu.backup <- seu
```

Cytokine
```{r}
res.final <- qread(here("data","cytokine","res_ETI.master.qs"))

```

## 2.2 Prepare object for downstream analysis
Run this for RecM, AM-CCL, and Combined
```{r}
# Subset object
seu.backup <- seu
seu <- subset(seu.backup, 
              sampleID %in% c("M1C160(1)","M1C160F", # CF-S1, treated
                              "M1C177","M1C177C", # CF-M3, treated
                              "M1C188","M1C188B", # CF-S3, treated
                              "M1C170C","M1C170D", # CF-M1, untreated
                              "M1C176", "M1C176C", # CF-M2
                              "M1C180","M1C180D", # CF-M4
                              "M1N066","M1N075","M1N078","M1N080","M1N087","M1N092" #HC1-6
                              ))
seu$Condition <- fct_drop(seu$Condition)
seu$sampleID <- fct_drop(seu$sampleID)

# Assign ID for analysis
seu$sampleID <- factor(seu$sampleID,
                       levels=c("M1N066","M1N075","M1N078","M1N080","M1N087","M1N092", #HC1-6
                              "M1C170C","M1C170D", # CF-M1, untreated
                              "M1C176","M1C176C", # CF-M2, untreated
                              "M1C180","M1C180D", # CF-M4, untreated
                              "M1C160(1)","M1C160F", # CF-S1, treated
                              "M1C177","M1C177C", # CF-M3, treated
                              "M1C188","M1C188B" # CF-S3, treated
                              ))

# Add Condition2
draft1 <- factor(
      case_when(
        seu$Condition == "Control" ~ "C"))
draft2 <- factor(
  case_when(
    seu$Bronchiectasis == "No" ~ "CFN",
    seu$Bronchiectasis == "Yes" ~ "CFB"
  )
)

Condition2 <- coalesce(draft1,draft2)
names(Condition2) <- colnames(seu)
seu <- AddMetaData(seu, Condition2, col.name = "Condition2")
seu$Condition2 <- factor(seu$Condition2,levels=c("C","CFB","CFN"))

# Add Timepoint
Timepoint <- setNames(c(rep("t1",6),rep(c("t1","t2"),6)),
                      levels(seu$sampleID))

Timepoint <- setNames(factor(Timepoint[seu$sampleID],
                             levels=c("t1","t2")),
                      colnames(seu))
seu <- AddMetaData(seu, Timepoint, col.name = "Timepoint")

# Add Condition3, which is CF.BL at timepoint 1, and either CF.UT or CF.ETI at timepoint 2
Condition3 <- setNames(c(rep("C",6),rep(c("UT1","UT2"),3),rep(c("pre-ETI","ETI"),3)),
                      levels(seu$sampleID))

Condition3 <- setNames(factor(Condition3[seu$sampleID],
                             levels=c("C","UT1","UT2","pre-ETI","ETI")),
                      colnames(seu))
seu <- AddMetaData(seu, Condition3, col.name = "Condition3")


# Assign analysis ID
ID <- setNames(c(paste0("C-0",1:6),
                 rep("CF-M1",2),rep("CF-M2",2),rep("CF-M4",2),
                 rep("CF-S1",2),rep("CF-M3",2),rep("CF-S3",2)),
               levels(seu$sampleID))

ID <- setNames(factor(ID[seu$sampleID],
                      levels=c(paste0("HC-0",1:6),
                               "CF-M1","CF-M2","CF-M4",
                               "CF-S1","CF-M3","CF-S3")),
               colnames(seu))

seu <- AddMetaData(seu, ID, col.name = "ID")

# Assign analysis ID2
ID2 <- setNames(c(paste0("HC-0",1:6),
                  "CF-01_BL","CF-01_UT",
                  "CF-02_BL","CF-02_UT",
                  "CF-03_BL","CF-03_UT",
                  "CF-04_BL","CF-04_ETI",
                  "CF-05_BL","CF-05_ETI",
                  "CF-06_BL","CF-06_ETI"),
                 levels(seu$sampleID))

ID2 <- setNames(factor(ID2[seu$sampleID],
                       levels=c(paste0("HC-0",1:6),
                                paste0("CF-0",1:3,"_BL"),
                                paste0("CF-0",1:3,"_UT"),
                                paste0("CF-0",4:6,"_BL"),
                                paste0("CF-0",4:6,"_ETI"))),
                colnames(seu))

seu <- AddMetaData(seu, ID2, col.name = "ID2")

# Assign batchID 2
draft1 <- factor(
      case_when(
        seu$sampleID %in% c("M1C170C","M1C176") ~ "batch 1"))

seu <- AddMetaData(seu, draft1, col.name = "draft1")

seu$batchID <- factor(seu$batchID,
                       levels=c("A","B","C","D"))

draft2 <- setNames(c("batch 3","batch 4","batch 2", "batch 1"),
                     levels(seu$batchID))

draft2 <- setNames(factor(draft2[seu$batchID],levels=paste0("batch ",1:4)),
                     colnames(seu))

seu <- AddMetaData(seu, draft2, col.name = "draft2")

# coalesce two draft columns
seu@meta.data %>%
  mutate(batchID2=coalesce(draft1,draft2)) -> seu@meta.data

# reorder levels
seu$batchID2 <- factor(seu$batchID2,
                       levels=paste0("batch ",1:4))

# remove draft columns
seu$draft2 <- NULL

# Factorize latent variables
seu$Bronchiectasis <- factor(seu$Bronchiectasis)
#seu$subtype <- factor(seu$subtype)

seu.backup <- NULL
gc()

s <- seu

# Set colors
Condition2.colours <- setNames(c("#26547c","#ffd166","#ef476f"),
                               levels(seu$Condition2))
Condition3.colours <- setNames(c("#26547c","#dda15e","#bc6c25","#a3b18a","#588157"),
                               levels(seu$Condition3))
batch.colours <- setNames(rev(c("#e07a5f", "#3d405b", "#81b29a", "#f2cc8f")),
                          levels(seu$batchID2))
```

Backup objects: RecM
```{r}
seu.RecM <- s
```

Backup objects: AM-CCL
```{r}
seu.AMCCL <- s
```

## 2.3 MDS
### 2.3.1 Aggregate expression
```{r}
# Aggregate expression
DefaultAssay(seu) <- "RNA"
#seu <- NormalizeData(seu)
#seu <- ScaleData(seu)

seu.pseudobulk <- AggregateExpression(seu,
                           group.by = c("sampleID"),
                           #features = deg,
                           slot = "counts",
                           return.seurat = TRUE,
                           assays = "RNA")
seu.pseudobulk <- NormalizeData(seu.pseudobulk)
seu.pseudobulk <- ScaleData(seu.pseudobulk)

```

### 2.3.2 Define genes
```{r}
# cytokine genes n=48
mapping_csv <- read.csv(here("data","cytokine","human_cytokine_gene_map.csv"))
deg <- (mapping_csv %>% filter(Cytokine %in% res.final$cyto_vars))$Gene
deg <- deg[deg %in% rownames(seu.pseudobulk)]
length(deg)

# profibrotic cytokine genes
fibro.cytokines <- c(
  "BAFF/TNFSF13B", "sCD30/TNFRSF8", "sCD163", "Chitinase-3-like1",
  "gp130/sIL-6Rβ", "sIL-6Rα", "Osteopontin", "Pentraxin-3", "TSLP",
  "TWEAK/TNFSF12", "FGFbasic", "Eotaxin", "IL-1β", "IL-5",
  "IL-8", "IL-13", "MCP-1(MCAF)", "SDF-1α", "PDGF-BB","M-CSF"
)
deg <- (mapping_csv %>% dplyr::filter(Cytokine %in% fibro.cytokines))$Gene
deg <- deg[deg %in% rownames(seu.pseudobulk)]

# # combined CF.ETI vs CF.ETI.BL DEGs
# celltype <- "combined"
# 
# # prepare DEGs
# # compare.list <- setNames(c("CF.ETI vs CF.ETI.BL","CF.UT vs CF.UT.BL","CF.ETI vs HC","CF.UT vs HC"),
# #                          c("CF.ETI_vs_CF.ETI.BL","CF.UT_vs_CF.UT.BL","CF.ETI_vs_HC","CF.UT_vs_HC"))
# compare.list <- setNames(c("CF.ETI vs CF.ETI.BL"),
#                          c("CF.ETI_vs_CF.ETI.BL"))
# # compare.list <- setNames(c("CF.ETI vs CF.ETI.BL","CF.UT vs CF.UT.BL"),
# #                          c("CF.ETI_vs_CF.ETI.BL","CF.UT_vs_CF.UT.BL"))
# 
# compare.list <- setNames(c("CF.UT vs CF.UT.BL"),
#                          c("CF.UT_vs_CF.UT.BL"))
# 
# deg <- c()
# for (compare in names(compare.list)) {
#   celltype <- "combined"
#   # read DE files
#   up <- read.csv(here("data","DE",celltype,compare,"DEG",paste0(celltype,".",compare,".DEG-up.csv")))
#   down <- read.csv(here("data","DE",celltype,compare,"DEG",paste0(celltype,".",compare,".DEG-down.csv")))
#   deg <- unique(c(deg, up$gene, down$gene))
# }
# 
# length(deg)

```

### 2.3.3 Plot MDS [whole BAL]
#### 2.3.3.1 All cytokine genes
```{r fig.width=9, fig.height=4}
# Get matrix
mtx <- as.matrix(t(GetAssayData(seu.pseudobulk, assay = "RNA", slot = "scale.data")))[,deg]
df <- data.frame(mtx)
# Before running MDS, we first calculate a distance matrix between all pairs of cells.  Here
# we use a simple euclidean distance metric on all genes, using scale.data as input
d <- parallelDist(mtx,threads=16)
# Run the MDS procedure, k determines the number of dimensions
mds <- cmdscale(d = d, k = 2)

# mds1 = single-cell
#mds1 <- mds

# cmdscale returns the cell embeddings, we first label the columns to ensure downstream
# consistency
colnames(mds) <- paste0("MDS_", 1:2)
# We will now store this as a custom dimensional reduction called 'mds'
seu.pseudobulk[["mds"]] <- CreateDimReducObject(embeddings = mds, key = "MDS_", assay = DefaultAssay(seu.pseudobulk))
seu.pseudobulk <- AddMetaData(seu.pseudobulk, seu.pseudobulk@reductions$mds@cell.embeddings[,1],col.name = "MDS_1")
seu.pseudobulk <- AddMetaData(seu.pseudobulk, seu.pseudobulk@reductions$mds@cell.embeddings[,2],col.name = "MDS_2")

# We can now use this as you would any other dimensional reduction in all downstream functions
seu.pseudobulk <- AddMetaData(seu.pseudobulk, factor(rownames(seu.pseudobulk@meta.data),
                                                     levels=rownames(seu.pseudobulk@meta.data)),col.name = "sampleID")
seu.pseudobulk <- AddMetaData(seu.pseudobulk, factor(c(rep("C",6),rep(c("CF.UT.BL","CF.UT"),3),rep(c("CF.ETI.BL","CF.ETI"),3)),
                                                     levels=c("C","CF.UT.BL","CF.UT","CF.ETI.BL","CF.ETI")),
                                                     col.name = "Condition")
seu.pseudobulk <- AddMetaData(seu.pseudobulk, factor(c("batch 4",
                                                       "batch 1",
                                                       "batch 4",
                                                       "batch 3",
                                                       "batch 3",
                                                       "batch 1",
                                                       "batch 1",
                                                       "batch 1",
                                                       "batch 1",
                                                       "batch 1",
                                                       "batch 2",
                                                       "batch 2",
                                                       "batch 1",
                                                       "batch 1",
                                                       "batch 2",
                                                       "batch 2",
                                                       "batch 1",
                                                       "batch 1")),
                                                     col.name = "batchID2")
seu.pseudobulk$batchID2 <- factor(seu.pseudobulk$batchID2, levels=c("batch 1","batch 2","batch 3","batch 4"))

Condition.colours <- setNames(c("#26547c","#dda15e","#bc6c25","#a3b18a","#588157"),
                               levels(seu.pseudobulk$Condition))
sample.colours <- setNames(c(rep("#26547c",6),rep(c("#dda15e","#bc6c25"),3),rep(c("#a3b18a","#588157"),3)),
                           levels(seu.pseudobulk$sampleID))

# --- Extract embeddings and metadata safely (avoid duplicate MDS_* columns) ---
emb <- as.data.frame(seu.pseudobulk@reductions$mds@cell.embeddings)
colnames(emb) <- c("MDS_1", "MDS_2")
emb$sampleID <- rownames(emb)

meta <- as.data.frame(seu.pseudobulk@meta.data)

# drop any MDS_* in meta to avoid .x/.y names on join
meta <- meta %>% select(-matches("^MDS_\\d+$"), everything())

# join by sampleID (meta may already have sampleID column)
if ("sampleID" %in% names(meta)) {
  emb <- emb %>% left_join(meta, by = "sampleID")
} else {
  emb <- emb %>% left_join(meta %>% tibble::rownames_to_column("sampleID"), by = "sampleID")
}

# Ensure canonical MDS_1 / MDS_2 exist
if (!("MDS_1" %in% names(emb)) && any(grepl("^MDS_1", names(emb)))) {
  emb$MDS_1 <- emb[[ grep("^MDS_1", names(emb))[1] ]]
}
if (!("MDS_2" %in% names(emb)) && any(grepl("^MDS_2", names(emb)))) {
  emb$MDS_2 <- emb[[ grep("^MDS_2", names(emb))[1] ]]
}

# Ensure Condition and sampleID factors match Seurat object definitions (keeps legend order)
emb$Condition <- factor(emb$Condition, levels = levels(seu.pseudobulk$Condition))
emb$sampleID  <- factor(emb$sampleID,  levels = levels(seu.pseudobulk$sampleID))

# Create SubjectID (first 6 chars for IDs starting with "M") to find pairs
emb <- emb %>%
  mutate(SubjectID = ifelse(startsWith(as.character(sampleID), "M"),
                            substr(as.character(sampleID), 1, 6),
                            as.character(sampleID)))

# --- Aggregate treatment to only C / ETI / UT and derive timepoint (TP1/TP2) ---
emb <- emb %>%
  mutate(
    # aggregated treatment only
    treatment = case_when(
      Condition == "C"                      ~ "C",
      Condition %in% c("CF.ETI.BL", "CF.ETI")~ "ETI",
      Condition %in% c("CF.UT.BL",  "CF.UT") ~ "UT",
      TRUE                                    ~ NA_character_
    ),
    # timepoint: baseline (.BL) -> TP1, otherwise TP2 (controls treated as TP1)
    timepoint = case_when(
      grepl("\\.BL$", Condition) ~ "TP1",
      Condition == "HC"          ~ "TP1",
      TRUE                       ~ "TP2"
    )
  )

# Factor ordering
emb$treatment <- factor(emb$treatment, levels = c("C", "UT", "ETI"))
emb$timepoint <- factor(emb$timepoint, levels = c("TP1", "TP2"))

# --- Colour, fill and shape mapping (following cytokine PCA) ---
stroke_cols <- c("C" = "#26547c", "ETI" = "#3a5a40", "UT" = "#936639")
shape_map   <- c("C" = 4, "ETI" = 24, "UT" = 21)

# fill: TP1 -> white (unfilled), TP2 -> treatment colour
emb <- emb %>%
  mutate(fill_col = ifelse(timepoint == "TP1", "white", stroke_cols[as.character(treatment)]))

# --- Prepare lines for paired samples (connect TP1 -> TP2) ---
lines_df <- emb %>%
  group_by(SubjectID) %>%
  filter(n() > 1) %>%
  arrange(SubjectID, timepoint) %>% # ensure ordering TP1 -> TP2
  ungroup()

# --- Sample-labelled panel (optional) ---
centroids_sample <- emb %>%
  group_by(sampleID) %>%
  summarize(x = mean(MDS_1, na.rm = TRUE), y = mean(MDS_2, na.rm = TRUE), .groups = "drop")

p_sample <- ggplot(emb, aes(x = MDS_1, y = MDS_2, color = sampleID)) +
  geom_point(size = 4) +
  geom_text_repel(
    data = centroids_sample,
    aes(x = x, y = y, label = sampleID),
    inherit.aes = FALSE,
    size = 3.2,
    box.padding = 0.3,
    point.padding = 0.3,
    segment.size = 0.3,
    max.overlaps = Inf
  ) +
  scale_color_manual(values = setNames(rep("#666666", length(levels(emb$sampleID))), levels(emb$sampleID))) +
  labs(title = "SampleID") +
  NoLegend() +
  theme_classic(base_size = 12) +
  theme(
    axis.text        = element_text(size = 12),
    axis.title       = element_text(size = 14),
    panel.border     = element_rect(color = "black", fill = NA, size = 0.7),
    axis.line        = element_blank(),
    legend.position  = "none",
    plot.title       = element_text(hjust = 0.5)
  )

# --- Treatment-styled MDS plot: shapes = treatment, stroke = treatment colour, fill = TP1 (white) or TP2 (colour) ---
title <- "Profibrotic cytokine genes"
#title <- "Cytokine level"
p_treatment <- ggplot(emb, aes(x = MDS_1, y = MDS_2)) +
  # paired connectors (behind points)
  geom_path(
    data = lines_df,
    aes(x = MDS_1, y = MDS_2, group = SubjectID),
    inherit.aes = FALSE,
    color = "grey40",
    size = 0.7,
    alpha = 0.85
  ) +

  # points on top: map shape & stroke by treatment, fill by precomputed fill_col (identity)
  geom_point(aes(shape = treatment, color = treatment, fill = fill_col),
             size = 3.8, stroke = 0.9, alpha = 0.98) +

  # ellipse only for controls (treatment == "C")
  stat_ellipse(
    data = emb %>% filter(treatment == "C"),
    aes(x = MDS_1, y = MDS_2, group = treatment),
    type = "norm", level = 0.68,
    geom = "polygon", alpha = 0.18, colour = NA,fill="#26547c", inherit.aes = FALSE, show.legend = FALSE
  ) +

  # NOTE: removed geom_text_repel that labeled treatments inside the plot (C/ETI/UT)

  # scales: stroke colours by treatment, fill uses identity (per-sample)
  scale_color_manual(values = stroke_cols, guide = "none") +
  scale_fill_identity(guide = "none") +

  # shape legend only; rename to "Treatment"
  scale_shape_manual(name = "Treatment", values = shape_map,
                     breaks = c("C", "UT", "ETI"),
                     labels = c("C", "UT", "ETI")) +

  labs(title = title) +
  theme_classic(base_size = 12) +
  theme(
    axis.text        = element_text(size = 12),
    axis.title       = element_text(size = 14),
    panel.border     = element_rect(color = "black", fill = NA, size = 0.7),
    axis.line        = element_blank(),
    legend.position  = "right",
    plot.title       = element_text(hjust = 0.5)
  )

# Combine and display side-by-side
p_combined <- p_sample | p_treatment
print(p_combined)

p_treatment.combined <- p_treatment
```

#### 2.3.3.2 All cytokine genes [batch]
```{r fig.height=4, fig.width=5}
# palette you provided (names "batch 1" ... "batch 4")
batch.colours <- setNames(
  rev(c("#e07a5f", "#3d405b", "#81b29a", "#f2cc8f")),
  c(paste0("batch ", 1:4))
)

# ensure batchID2 strings match the palette names (add "batch " prefix only if missing)
emb$batchID2 <- as.character(emb$batchID2)
if (!any(grepl("^batch", emb$batchID2))) {
  emb$batchID2 <- paste("batch", emb$batchID2)
}

# show batch only for UT and ETI, NA for controls (C)
emb$batch_shown <- ifelse(emb$treatment %in% c("UT", "ETI"), emb$batchID2, NA_character_)

# keep only palette entries actually present in batch_shown
present_batches <- intersect(names(batch.colours), unique(na.omit(emb$batch_shown)))
batch.colours2 <- batch.colours[present_batches]

p_batch <- ggplot(emb, aes(x = MDS_1, y = MDS_2)) +
  # points: shape by treatment (unchanged), fill by batch_shown (NA for C)
  geom_point(aes(shape = treatment, fill = batch_shown),
             size = 3.8, stroke = 0.9, alpha = 0.98) +

  # paired connectors (behind points)
  geom_path(
    data = lines_df,
    aes(x = MDS_1, y = MDS_2, group = SubjectID),
    inherit.aes = FALSE,
    color = "grey10",
    size = 0.7,
    alpha = 0.85
  ) +

  # ellipse only for controls (treatment == "C")
  stat_ellipse(
    data = emb %>% filter(treatment == "C"),
    aes(x = MDS_1, y = MDS_2, group = treatment),
    type = "norm", level = 0.68,
    geom = "polygon", alpha = 0.18, colour = NA, fill = "#26547c",
    inherit.aes = FALSE, show.legend = FALSE
  ) +

  # manual fill for Batch: don't show NA, and draw filled legend keys
  scale_fill_manual(
    name = "Batch",
    values = batch.colours2,
    na.translate = FALSE,
    guide = guide_legend(override.aes = list(shape = 21, colour = "black"))
  ) +

  # keep your existing shape mapping (shape_map unchanged)
  scale_shape_manual(name = "Treatment", values = shape_map,
                     breaks = c("C", "UT", "ETI"),
                     labels = c("C", "UT", "ETI")) +

  labs(title = "Cytokine gene expression") +
  theme_classic(base_size = 12) +
  theme(
    axis.text        = element_text(size = 12),
    axis.title       = element_text(size = 14),
    panel.border     = element_rect(color = "black", fill = NA, size = 0.7),
    axis.line        = element_blank(),
    legend.position  = "right",
    plot.title       = element_text(hjust = 0.5)
  )

p_batch
```


## 2.6 Pseudobulk gene expression in RecM
### 2.6.1 Function
```{r}
plot_pb_gene <- function(pb_seu,
                          gene,
                          assay = "RNA",
                          slot = "data",
                          condition_col = "Condition3",
                          pairing_col = "subjectID",
                          pal = Condition3.colours,
                          zero_filter = TRUE,
                          connect_pairs = TRUE,
                          show_stats = TRUE,
                          step_increase = 0.5,
                          treatment_levels = c("UT1","UT2","pre-ETI","ETI")) {
  # minimal checks
  if (missing(pb_seu)) stop("pb_seu required")
  if (missing(gene)) stop("gene required")
  if (!requireNamespace("Seurat", quietly = TRUE)) stop("Seurat required")
  if (!requireNamespace("ggplot2", quietly = TRUE)) stop("ggplot2 required")
  if (show_stats && !requireNamespace("ggpubr", quietly = TRUE)) {
    warning("ggpubr not installed; stat brackets disabled")
    show_stats <- FALSE
  }

  # get expression matrix and metadata
  mat <- Seurat::GetAssayData(pb_seu, assay = assay, slot = slot)
  if (!gene %in% rownames(mat)) stop("gene not found in assay rows")
  meta_df <- as.data.frame(pb_seu@meta.data, stringsAsFactors = FALSE)

  # align metadata to columns
  sample_ids <- colnames(mat)
  if (!is.null(rownames(meta_df)) && all(sample_ids %in% rownames(meta_df))) {
    meta_df <- meta_df[sample_ids, , drop = FALSE]
  } else if ("sampleID" %in% names(meta_df) && all(sample_ids %in% meta_df$sampleID)) {
    rownames(meta_df) <- meta_df$sampleID
    meta_df <- meta_df[sample_ids, , drop = FALSE]
  } else {
    # best-effort normalized match or intersection
    norm <- function(x) tolower(gsub("[^[:alnum:]]+", "", as.character(x)))
    mm <- match(norm(sample_ids), norm(rownames(meta_df)))
    if (all(!is.na(mm))) {
      meta_df <- meta_df[mm, , drop = FALSE]
      rownames(meta_df) <- sample_ids
    } else {
      common <- intersect(sample_ids, rownames(meta_df))
      if (length(common) == 0) stop("No overlap between assay cols and metadata rownames")
      keep <- sample_ids %in% common
      mat <- mat[, keep, drop = FALSE]
      sample_ids <- colnames(mat)
      meta_df <- meta_df[sample_ids, , drop = FALSE]
    }
  }

  if (!condition_col %in% names(meta_df)) stop("condition_col not found in metadata")

  # build plotting dataframe; protect against duplicate names
  gene_col <- paste0(gene, "_expr")
  expr_vec <- as.numeric(mat[gene, , drop = TRUE])
  plot_df <- data.frame(sampleID = sample_ids, stringsAsFactors = FALSE)
  plot_df[[gene_col]] <- expr_vec
  meta_for_bind <- meta_df
  if ("sampleID" %in% names(meta_for_bind)) meta_for_bind$sampleID <- NULL
  if (gene_col %in% names(meta_for_bind)) meta_for_bind[[gene_col]] <- NULL
  plot_df <- cbind(plot_df, meta_for_bind[plot_df$sampleID, , drop = FALSE])

  # simple filtering: remove NAs and optional zero filter
  if (zero_filter) plot_df <- plot_df[!is.na(plot_df[[gene_col]]) & plot_df[[gene_col]] != 0, , drop = FALSE]
  else plot_df <- plot_df[!is.na(plot_df[[gene_col]]), , drop = FALSE]

  # remove controls ("C") entirely
  plot_df <- plot_df[plot_df[[condition_col]] != "C", , drop = FALSE]

  # require condition present
  plot_df <- plot_df[!is.na(plot_df[[condition_col]]), , drop = FALSE]
  if (nrow(plot_df) == 0) stop("No samples available after filtering/matching")

  # factorize Treatment using provided treatment_levels (ensures numeric x ordering)
  plot_df$Treatment <- factor(as.character(plot_df[[condition_col]]), levels = treatment_levels)
  # drop samples with NA Treatment (i.e., not in treatment_levels)
  plot_df <- plot_df[!is.na(plot_df$Treatment), , drop = FALSE]
  if (nrow(plot_df) == 0) stop("No samples with requested treatment levels present")

  # pairing column
  if (!pairing_col %in% names(plot_df)) pairing_col <- NULL
  if (!is.null(pairing_col)) plot_df[[pairing_col]] <- as.factor(plot_df[[pairing_col]])

  # palette: ensure it maps to treatment_levels (subset if needed)
  levs <- levels(plot_df$Treatment)
  if (!is.null(pal) && !is.null(names(pal))) {
    pal <- pal[intersect(levs, names(pal))]
    missing_levs <- setdiff(levs, names(pal))
    if (length(missing_levs) > 0) {
      extras <- RColorBrewer::brewer.pal(max(3, length(missing_levs)), "Dark2")
      pal <- c(pal, setNames(extras[seq_along(missing_levs)], missing_levs))
    }
    pal <- pal[levs]
  } else {
    pal_vals <- RColorBrewer::brewer.pal(max(3, length(levs)), "Dark2")
    pal <- setNames(pal_vals[seq_along(levs)], levs)
  }

  # compute pairwise tests for two comparisons of interest (ETI vs pre-ETI, UT2 vs UT1)
  stat.test.pb <- NULL
  if (show_stats) {
    comps <- list(c("ETI","pre-ETI"), c("UT2","UT1"))
    res_list <- list()
    for (cmp in comps) {
      gA <- cmp[1]; gB <- cmp[2]
      if (!(gA %in% levs && gB %in% levs)) next
      valsA <- plot_df[[gene_col]][plot_df$Treatment == gA]
      valsB <- plot_df[[gene_col]][plot_df$Treatment == gB]
      pval <- NA_real_
      # try paired test if pairing_col exists and there are >=2 shared subjects
      if (!is.null(pairing_col)) {
        idsA <- unique(plot_df[[pairing_col]][plot_df$Treatment == gA])
        idsB <- unique(plot_df[[pairing_col]][plot_df$Treatment == gB])
        shared <- intersect(idsA, idsB)
        if (length(shared) >= 2) {
          xa <- sapply(shared, function(s) plot_df[[gene_col]][plot_df$Treatment == gA & plot_df[[pairing_col]] == s][1])
          xb <- sapply(shared, function(s) plot_df[[gene_col]][plot_df$Treatment == gB & plot_df[[pairing_col]] == s][1])
          keep <- !is.na(xa) & !is.na(xb)
          xa <- xa[keep]; xb <- xb[keep]
          if (length(xa) >= 2) pval <- tryCatch(wilcox.test(xa, xb, paired = TRUE)$p.value, error = function(e) NA_real_)
        }
      }
      # fallback to unpaired if paired couldn't be run
      if (is.na(pval)) {
        if (length(valsA) >= 2 && length(valsB) >= 2) {
          pval <- tryCatch(wilcox.test(valsA, valsB, paired = FALSE)$p.value, error = function(e) NA_real_)
        } else {
          pval <- NA_real_
        }
      }
      res_list[[paste(gA,gB,sep="_")]] <- data.frame(group1 = gA, group2 = gB, p = pval, stringsAsFactors = FALSE)
    }
    if (length(res_list) > 0) {
      stat.test.pb <- do.call(rbind, res_list)
      stat.test.pb$p.adj <- p.adjust(stat.test.pb$p, method = "BH")
      stat.test.pb$p.adj.signif <- sapply(stat.test.pb$p.adj, function(pv) {
        if (is.na(pv)) return(NA_character_)
        if (pv < 1e-4) "****" else if (pv < 1e-3) "***" else if (pv < 1e-2) "**" else if (pv < 0.05) "*" else "ns"
      })
      # compute y positions using fixed scalar step_increase
      # compute per-comparison y from the global max of plotted values
      y_max <- max(plot_df[[gene_col]], na.rm = TRUE)
      ncomp <- nrow(stat.test.pb)
      if (ncomp > 0) {
        stat.test.pb$xmin <- match(stat.test.pb$group1, levs)
        stat.test.pb$xmax <- match(stat.test.pb$group2, levs)
        stat.test.pb$y.position <- y_max + seq_len(ncomp) * step_increase
      } else stat.test.pb <- NULL
    }
  }

  # Build the plot: numeric x for exact line connection, points colored by Treatment
  library(ggplot2)
  library(ggrepel)
  # minimal additions: map three ETI subjectIDs to short labels and compute label positions
  label_map <- c(M1C177 = "E1", M1C160 = "E2", M1C188 = "E3")
  label_df <- subset(plot_df, Treatment == "ETI" & subjectID %in% names(label_map))
  if (nrow(label_df) > 0) {
    label_df$ETI_label <- label_map[as.character(label_df$subjectID)]
    # place label to the right of the point (same y as point)
    label_df$label_y <- label_df[[gene_col]]
  }
  
  p <- ggplot() +
  # lines (numeric x) connecting paired points exactly
  { if (connect_pairs && !is.null(pairing_col)) {
      pair_counts <- table(plot_df[[pairing_col]])
      valid_pairs <- names(pair_counts)[pair_counts >= 2]
      if (length(valid_pairs) > 0) {
        lines_df <- plot_df[plot_df[[pairing_col]] %in% valid_pairs, , drop = FALSE]
        lines_df <- lines_df[order(lines_df[[pairing_col]], as.numeric(lines_df$Treatment)), , drop = FALSE]
        geom_line(data = lines_df,
                  aes(x = as.numeric(Treatment), y = .data[[gene_col]], group = .data[[pairing_col]]),
                  color = "grey70", size = 0.5)
      } else NULL
    } else NULL } +
  # points at numeric x
  geom_point(data = plot_df,
             aes(x = as.numeric(Treatment), y = .data[[gene_col]], color = Treatment),
             size = 2.5) +
  # use ggrepel to place labels and draw connecting segments to the points
  { if (exists("label_df") && nrow(label_df) > 0) {
      geom_text_repel(
        data = label_df,
        aes(x = as.numeric(Treatment), y = .data[[gene_col]], label = ETI_label),
        nudge_x = 1,            # push labels to the right of the points
        hjust = 0.5,
        size = 3.5,
        fontface = "bold",        
        color = "#3A5A40",
        segment.size = 0.35,       # line thickness connecting label -> point
        segment.color = "grey60",
        box.padding = 0.25,
        point.padding = 0.3,
        max.overlaps = Inf,
        show.legend = FALSE
      )
    } else NULL } +
  # x axis: numeric positions labeled by treatments
  scale_x_continuous(breaks = seq_along(treatment_levels),
                     labels = c("Baseline","UT","Baseline","ETI"),
                     limits = c(0.5, length(treatment_levels) + 0.5)) +
  scale_color_manual(values = pal) +
  labs(y = "log1p exprssion", x="Treatment",color = "Treatment", title = gene) +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1, size = 10),
        legend.position = "right",
        plot.title = element_text(hjust = 0.5, face = "bold.italic")) + NoLegend()

  # add p-value brackets if computed
  if (show_stats && !is.null(stat.test.pb) && nrow(stat.test.pb) > 0) {
    p <- p + ggpubr::stat_pvalue_manual(stat.test.pb,
                                        label = "p.adj.signif",
                                        y.position = "y.position",
                                        tip.length = 0.01, bracket.size = 0.4, hide.ns = TRUE)
  }

  return(p)
}

# plot with controls
combine_with_control_box <- function(plot_a,
                                     pb_seu,
                                     gene,
                                     assay = "RNA",
                                     slot = "data",
                                     condition_col = "Condition3",
                                     control_label = "C",
                                     control_colour = "#26547c",
                                     box_width = 0.6,
                                     dodge_width = 0.6,
                                     patchwork_widths = c(3, 1)) {
  if (missing(plot_a) || !inherits(plot_a, "ggplot")) stop("plot_a (ggplot) is required")
  if (missing(pb_seu) || missing(gene)) stop("pb_seu and gene are required")

  if (!requireNamespace("Seurat", quietly = TRUE)) stop("Seurat required")

  # 1. extract expression + metadata
  mat <- tryCatch(Seurat::GetAssayData(pb_seu, assay = assay, slot = slot),
                  error = function(e) stop("Failed to GetAssayData(): ", conditionMessage(e)))
  if (!gene %in% rownames(mat)) stop(sprintf("Gene '%s' not found in assay rows", gene))

  meta_df <- as.data.frame(pb_seu@meta.data, stringsAsFactors = FALSE)
  sample_ids <- colnames(mat)
  if (!is.null(rownames(meta_df)) && all(sample_ids %in% rownames(meta_df))) {
    meta_df <- meta_df[sample_ids, , drop = FALSE]
  } else if ("sampleID" %in% names(meta_df) && all(sample_ids %in% meta_df$sampleID)) {
    rownames(meta_df) <- meta_df$sampleID
    meta_df <- meta_df[sample_ids, , drop = FALSE]
  } else {
    common <- intersect(sample_ids, rownames(meta_df))
    if (length(common) == 0) stop("No overlap between assay columns and pb_seu metadata rownames")
    mat <- mat[, common, drop = FALSE]
    sample_ids <- colnames(mat)
    meta_df <- meta_df[sample_ids, , drop = FALSE]
  }

  if (!condition_col %in% names(meta_df)) stop(sprintf("condition_col '%s' not found in pb_seu metadata", condition_col))

  # avoid name conflicts when cbind'ing
  conflicts <- intersect(names(meta_df), c("sampleID", "expr", gene))
  if (length(conflicts) > 0) meta_df <- meta_df[, setdiff(names(meta_df), conflicts), drop = FALSE]

  expr_vec <- as.numeric(mat[gene, , drop = TRUE])
  df_all <- data.frame(sampleID = sample_ids, expr = expr_vec, stringsAsFactors = FALSE)
  df_all <- cbind(df_all, meta_df[sample_ids, , drop = FALSE])
  if (any(duplicated(names(df_all)))) names(df_all) <- make.unique(names(df_all))

  # subset controls
  df_ctrl <- df_all[df_all[[condition_col]] == control_label, , drop = FALSE]
  if (nrow(df_ctrl) == 0) stop(sprintf("No control samples with %s == '%s' found", condition_col, control_label))

  # 2. infer y-limits and y-label from plot_a
  y_limits <- NULL
  y_label  <- NULL
  built <- tryCatch(ggplot2::ggplot_build(plot_a), error = function(e) NULL)
  if (!is.null(built)) {
    yvals <- unlist(lapply(built$data, function(ld) if (!is.null(ld$y)) ld$y else NULL))
    if (length(yvals) > 0 && any(!is.na(yvals))) y_limits <- range(yvals, na.rm = TRUE)
    if (is.null(y_limits) && length(built$layout$panel_scales_y) >= 1) {
      rng <- tryCatch(built$layout$panel_scales_y[[1]]$range$range, error = function(e) NULL)
      if (!is.null(rng) && length(rng) == 2) y_limits <- rng
    }
    if (!is.null(plot_a$labels$y) && nzchar(as.character(plot_a$labels$y))) y_label <- plot_a$labels$y
  }
  if (is.null(y_limits) || any(!is.finite(y_limits))) y_limits <- range(c(df_ctrl$expr, 0), na.rm = TRUE)
  if (is.null(y_label) || !nzchar(as.character(y_label))) y_label <- if (!is.null(plot_a$labels$y)) plot_a$labels$y else "Expression"

  # 3. build the control-only boxplot (no jitter) using position_dodge
  df_ctrl$Control_x <- control_label

  b <- ggplot(df_ctrl, aes(x = factor(Control_x, levels = control_label), y = expr)) +
    geom_boxplot(width = box_width, outlier.shape = NA, fill = NA, colour = control_colour, position = position_dodge(width = dodge_width)) +
    geom_point(position = position_dodge(width = dodge_width), size = 2.5, color = control_colour) +
    scale_x_discrete(labels = control_label, name = "") +
    coord_cartesian(ylim = y_limits) +
    labs(y = y_label, title = NULL) +
    theme_classic(base_size = 12) +
    theme(
      plot.title = element_blank(),
      axis.text.x = element_text(angle = 40, hjust = 1, size = 10, colour = "black"),
      axis.text.y = element_text(size = 12, colour = "black"),
      axis.title.x = element_blank(),
      legend.position = "none"
    )

  # 4. combine with patchwork
  combined <- plot_a + b + patchwork::plot_layout(widths = patchwork_widths)
  combined
}

```

### 2.6.2 Prepare pseudobulk object
```{r}
# Aggregate raw RNA counts by BAL ID (pseudobulk)
pseudo_bulk <- AggregateExpression(
  seu, # or s
  assays = "RNA",
  slot = "counts",
  group.by = "sampleID",
  return.seurat = TRUE
)

pseudo_bulk <- NormalizeData(pseudo_bulk)

# one minimal chunk: add Condition3, sampleID, and subjectID to pseudo_bulk (aligned to pseudo_bulk columns)
# assumes `seu` (single-cell Seurat) and `pseudo_bulk` (AggregateExpression -> return.seurat = TRUE) exist

sample_order <- colnames(pseudo_bulk)
seu_sample_vec <- as.character(seu$sampleID)

# align seu -> pseudo_bulk using exact match, fall back to normalized matching if needed
idx <- match(sample_order, seu_sample_vec)
if (any(is.na(idx))) {
  norm <- function(x) tolower(gsub("[^[:alnum:]]+", "", as.character(x)))
  idx2 <- match(norm(sample_order), norm(seu_sample_vec))
  idx[is.na(idx)] <- idx2[is.na(idx)]
}
if (any(is.na(idx))) warning("Some pseudo_bulk samples were not found in seu$sampleID; corresponding Condition3 will be NA")

# build metadata frame (rows in same order as pseudo_bulk columns)
meta_df <- data.frame(
  sampleID   = sample_order,
  Condition3 = if ("Condition3" %in% colnames(seu@meta.data)) seu$Condition3[idx] else NA,
  subjectID  = substr(sample_order, 1, 6),
  row.names  = sample_order,
  stringsAsFactors = FALSE
)

# preserve factor levels for Condition3 if present in seu
if ("Condition3" %in% names(seu) && is.factor(seu$Condition3)) {
  meta_df$Condition3 <- factor(meta_df$Condition3, levels = levels(seu$Condition3))
}

# add to pseudo_bulk and coerce subjectID to factor for pairing
pseudo_bulk <- AddMetaData(pseudo_bulk, metadata = meta_df)
pseudo_bulk$subjectID <- factor(pseudo_bulk$subjectID)

# quick sanity checks
pseudo_bulk@meta.data[, c("Condition3", "sampleID", "subjectID")]
table(pseudo_bulk$subjectID, useNA = "ifany")
```

### 2.6.2 Plot
```{r fig.height=3, fig.width=4}
fibro.genes <- c("SPP1","CHIT1","CHI3L1","ITGB8","MMP9","MMP19","PLA2G7","LGMN","MERTK")
inflam.genes <- c("CCL2","CCR2","CCL3","CCL4","IL6","CXCL8","IL1RN","TNFSF13","TNFSF12","TNFRSF1A","TNFSF13B","TNF","CSF1")
genes <- c(fibro.genes, inflam.genes)

# RecM
plot.list <- list()
plot.list <- lapply(genes, FUN = function(g) {
  plot_pb_gene(
    pb_seu = pseudo_bulk,
    gene = g,
    condition_col = "Condition3",            # metadata column name
    pairing_col = "subjectID",
    pal = Condition3.colours,                # palette vector
  )
})
names(plot.list) <- genes

plotC.list <- list()
for (g in genes) {
  p <- combine_with_control_box(plot_a = plot.list[[g]], pb_seu = pseudo_bulk, gene = g, condition_col = "Condition3")
  plotC.list[[g]] <- p
}

# check plot
plot.list$CHI3L1
```

# 3 Save figures
```{r}
# celltype <- "ETI"
# out <- here("output","plots",celltype)
# out <- here("output","plots",celltype,"expr_plot")
# if(!dir.exists(out)) {
#   dir.create(out,recursive = TRUE)
# }
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 1600)
# 
# 
# # MDS plots
# # combined cytokine gene
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 1600)
# 
# ggsave(plot=p_sample + p_treatment +
#   theme(plot.tag = element_text(size = 14, face = "bold", family="Arial")),
#        device="jpeg",
#        dpi=1600,
#        filename = here("data","plots","ETI","bal_ETI.combined.FibroCytoGenes.MDSplot.jpeg"),
#        height=4,
#        width=9)
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 100)
# 
# # combined cytokine gene by batch
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 1600)
# 
# ggsave(plot=p_batch +
#          theme(plot.tag = element_text(size = 14, face = "bold", family="Arial")),
#        device="jpeg",
#        dpi=1600,
#        filename = here("data","plots","ETI","bal_ETI.combined.batch.MDSplot.jpeg"),
#        height=4,
#        width=5)
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 100)
# 
# # Box plot of pseudobulk marker genes in RecM
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 1600)
# 
# for (g in genes.am) {
#   plotC.list[[g]] +
#     theme(plot.tag = element_text(size = 14, face = "bold"))
# 
#   ggsave(plot=last_plot(),
#          device="jpeg",
#          dpi=1600,
#          filename = paste0(out,"/","RecM.pseudobulk.",g,".jpeg"),
#          height=4,
#          width=4)
# }
# 
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 100)
# 
# 
# # Box plot of pseudobulk marker genes in whole BAL
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 1600)
# 
# for (g in genes) {
#   plotC.list[[g]] +
#     theme(plot.tag = element_text(size = 14, face = "bold"))
# 
#   ggsave(plot=last_plot(),
#          device="jpeg",
#          dpi=1600,
#          filename = paste0(out,"/","combined.pseudobulk.",g,".jpeg"),
#          height=4,
#          width=4)
# }
# 
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 100)
```

# 4 Save object
```{r}
celltype <- "ETI"
out <- here("data","SCEs","analysis",celltype)
if(!dir.exists(out)) {
  dir.create(out,recursive = TRUE)
}

out <- here("data","SCEs","analysis",celltype,paste0("bal_",celltype,".analysis.SEUlist.qs"))

if (!file.exists(out)) {
  qsave(seu.list,file=out,nthreads=16)
}
```

# 5 Save tables
```{r}
emb %>% write_csv(here("output","Fig6B.csv"))

# Figure 6D
genes <- c("CHI3L1", "MMP9", "SPP1", "CCL2", "CCL3", "CXCL8")

mat <- GetAssayData(pseudo_bulk, assay = "RNA", slot = "data")

genes_ok <- intersect(genes, rownames(mat))
stopifnot(length(genes_ok) > 0)

df_expr <- as.data.frame(t(as.matrix(mat[genes_ok, , drop = FALSE]))) %>%
  rownames_to_column("sampleID")

meta_df <- pseudo_bulk@meta.data

df <- df_expr %>%
  left_join(meta_df, by = "sampleID") %>%
  #filter(Condition3 != "C") %>%    # matches your plot function’s control removal :contentReference[oaicite:1]{index=1}
  mutate(Treatment = factor(as.character(Condition3),
                            levels = c("UT1","UT2","pre-ETI","ETI","C")))

df

df %>% write_csv(here("output","Fig6D.csv"))
```
