---
title: 'BAL cytokine analysis (ETI)'
author:
- name: Anson Wong
date: "`r Sys.Date()`"
description: null
output: html_document
editor_options:
  chunk_output_type: inline
---

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
xaringanExtra::use_panelset()

```

# 1 Load packages
```{r}
suppressPackageStartupMessages({
library(Seurat)
library(here)
library(qs)
library(forcats)
library(pals)
library(ggplot2)
library(ggpubr)
library(rstatix)
library(patchwork)
library(dplyr)
library(readr)
library(cowplot)
library(ComplexHeatmap)
library(circlize)
library(RColorBrewer)
library(grid) 
library(tibble)
library(purrr)
library(ggrepel)
library(scales)
library(tidyr)
library(ggrepel)
library(scales)
library(lubridate)
library(sva)
library(stringr)
library(vctrs)
library(limma)
library(kableExtra)
library(aod)
})
set.seed(1990)

```

# 2 Prepare 37-plex and 48-plex cytokine data

## 2.1 Configuration and helpers

```{r}
plexes <- c("37" = "37-Plex_assay", "48" = "48-Plex_assay")
base_data_dir <- "/group/canc2/anson/working/cf-eti-bal/data/cytokine/250806"

clean_metadata <- function(x) {
  x <- sub("\\s+$", "", x)
  x <- gsub(" ", "_", x)
  x <- gsub("[[:punct:]&&[^\\-]]", "", x)
  x
}
clean_cytokine <- function(x) {
  x <- gsub("\\s*\\(pg/mL\\)", "", x)
  x <- gsub(" ", "", x)
  x
}
format_batch_date <- function(d) {
  fmt <- as.Date(d, "%Y-%m-%d")
  ifelse(is.na(fmt), d, format(fmt, "%-d/%-m/%Y"))
}
```

## 2.2 Processing function

```{r}
process_cytokine <- function(plex_size) {
  data_dir <- file.path(base_data_dir, plexes[plex_size])
  files    <- list.files(data_dir, "\\.csv$", full.names = TRUE)
  if (length(files)==0) stop("No CSVs found in ", data_dir)

  lst <- lapply(files, function(path) {
    df <- read.csv(path, stringsAsFactors = FALSE, fileEncoding = "UTF-8", check.names = FALSE)
    orig    <- names(df)
    is_cyto <- grepl("pg/mL", orig, fixed = TRUE)
    md      <- orig[!is_cyto]
    cy      <- orig[ is_cyto]
    new_md <- clean_metadata(md)
    new_cy <- clean_cytokine(cy)
    names(df) <- c(new_md, new_cy)
    
    # # For each cytokine column in this file, compute the minimal observed non-NA value.
    # # If >50% of non-missing samples equal that minimal value (after rounding to 2 dp),
    # # treat the entire column as below-detection for this run by setting it to NA.
    # digits <- 2
    # for (cn in new_cy) {
    #   vals <- suppressWarnings(as.numeric(df[[cn]]))
    #   non_na <- !is.na(vals)
    #   n_non_na <- sum(non_na)
    #   if (n_non_na == 0) next
    # 
    #   min_val <- min(vals[non_na], na.rm = TRUE)
    #   # count how many non-NA values equal the minimal value after rounding to `digits`
    #   cnt_min <- sum(round(vals[non_na], digits) == round(min_val, digits), na.rm = TRUE)
    #   frac_min <- cnt_min / n_non_na
    # 
    #   # if >50% of samples equal the minimal (rounded) value, mark column as all NA for this file
    #   if (frac_min > 0.5) {
    #     df[[cn]] <- NA_real_
    #   }
    # }
    # 
    date_col  <- "Date_performed"
    plate_col <- grep("^What_plate", names(df), value = TRUE)[1]
    date_str  <- format_batch_date(df[[date_col]])
    plate_str <- if (!is.na(plate_col)) as.character(df[[plate_col]]) else ""
    df$Batch  <- ifelse(plate_str=="",
                        date_str,
                        paste0(date_str, "_p", plate_str))
    first_cyto <- new_cy[1]
    df <- df %>% relocate(Batch, .before = all_of(first_cyto))
    list(df = df, cytokines = new_cy)
  })

  all_cyto <- unique(unlist(lapply(lst, "[[", "cytokines")))
  merged   <- do.call(rbind, lapply(lst, "[[", "df"))

  keep_cyto <- sapply(all_cyto, function(cn) {
    non_na_by_batch <- tapply(!is.na(merged[[cn]]), merged$Batch, any)
    all(non_na_by_batch)
  })
  to_remove <- setdiff(all_cyto, names(keep_cyto)[keep_cyto])
  raw_backup <- merged[, setdiff(names(merged), to_remove)]

  # edit metadata (infection, condition, etc)
  raw <- raw_backup
  raw$Infection_bacteria <- ifelse(!is.na(raw$Bacteria) & raw$Bacteria!="", "Yes", "No")
  raw$Infection_virus    <- ifelse(!is.na(raw$Virus)    & raw$Virus!="",    "Yes", "No")
  raw$Infection_fungi    <- ifelse(!is.na(raw$Fungi)    & raw$Fungi!="",    "Yes", "No")
  raw <- raw %>%
    relocate(Infection_bacteria, .before="Bacteria") %>%
    relocate(Infection_virus,    .before="Virus")    %>%
    relocate(Infection_fungi,    .before="Fungi") %>%
    mutate(
      CF_Severity = ifelse(!is.na(CF_Severity) & CF_Severity!="",
                           paste0("CF_", CF_Severity), NA_character_),
      Condition   = coalesce(CF_Severity, Control_Subgroup),
      Condition   = gsub(" ", "_", Condition)
    ) %>%
    relocate(Condition, .before=CF_Severity) %>%
    mutate(
      Condition1 = case_when(
        grepl("^CF_mild$",   Condition, ignore.case=TRUE) ~ "CFN",
        grepl("^CF_severe$", Condition, ignore.case=TRUE)       ~ "CFB",
        TRUE                                              ~ "non-CF"
      ),
      Condition2 = case_when(
        grepl('eAIR',      Record_ID, ignore.case=TRUE)  ~ "C",
        grepl("true",      Condition, ignore.case=TRUE) ~ "C",
        grepl("CF_mild",   Condition, ignore.case=TRUE) ~ "CFN",
        grepl("CF_severe", Condition, ignore.case=TRUE) ~ "CFB",
        grepl("CSLD",      Condition, ignore.case=TRUE) &
        grepl("wheeze",    Condition, ignore.case=TRUE) ~ "CSLD_wheeze",
        grepl("CSLD",      Condition, ignore.case=TRUE) ~ "CSLD",
        grepl("wheeze",    Condition, ignore.case=TRUE) ~ "Wheeze",
        grepl("oncology",  Condition, ignore.case=TRUE) ~ "Oncology",
        grepl("BMT",       Condition, ignore.case=TRUE) ~ "BMT",
        TRUE                                               ~ sub("_$", "", Condition)
      )
    ) %>%
    relocate(Condition1, Condition2, .after=Condition) %>%
    mutate(
    #CF_Severity  = ifelse(is.na(CF_Severity)  | CF_Severity  == "", "NaN", CF_Severity),
    CF_treatment = ifelse(is.na(CF_treatment) | CF_treatment == "", "NaN", CF_treatment)
  )

  # PCA & ComBat
  metadata_vars <- c("Record_ID","Batch","Age","Sex",
                     "Condition1","Condition2","CF_treatment",
                     "Infection_bacteria","Infection_virus","Infection_fungi")

  cyto_vars <- intersect(all_cyto, colnames(raw))
  expr_mat  <- as.matrix(raw[, cyto_vars])
  expr_log  <- log2(expr_mat)

  run_pca_and_tests <- function(mat, meta, pcs=1:10) {
    pca    <- prcomp(mat, center=TRUE, scale.=TRUE)
    scores <- as.data.frame(pca$x[,pcs,drop=FALSE])
    scores_meta <- bind_cols(scores, meta)
    vars_to_test <- setdiff(names(meta), "Record_ID")
    pvals <- sapply(vars_to_test, function(var) {
      sapply(paste0("PC", pcs), function(pc) {
        summary(aov(as.formula(paste(pc,"~",var)), data=scores_meta))[[1]]["Pr(>F)"][1,]
      })
    }, simplify="matrix")
    colnames(pvals) <- vars_to_test
    rownames(pvals) <- paste0("PC", pcs)
    list(pca=pca, scores=scores_meta, pvals=pvals)
  }

  pre_res <- run_pca_and_tests(expr_log, raw[metadata_vars])
  mod     <- model.matrix(~ Age + Sex + Condition2 +
                           Infection_bacteria + Infection_virus + Infection_fungi,
                         data=raw)
  combat_out   <- ComBat(dat = t(expr_log), batch=raw$Batch,
                         mod=mod, par.prior=TRUE)
  expr_combat  <- t(combat_out)
  post_res     <- run_pca_and_tests(expr_combat, raw[metadata_vars])

  list(
    raw=raw, expr_log=expr_log, expr_combat=expr_combat,
    pre_res=pre_res, post_res=post_res,
    cyto_vars=cyto_vars
  )
}
```

## 2.3 Run both 37-plex and 48-plex

```{r}
results <- lapply(names(plexes), process_cytokine)
names(results) <- names(plexes)

raw37         <- results[["37"]]$raw
expr_log37    <- results[["37"]]$expr_log
expr_combat37 <- results[["37"]]$expr_combat

raw48         <- results[["48"]]$raw
expr_log48    <- results[["48"]]$expr_log
expr_combat48 <- results[["48"]]$expr_combat
```

# 3 Plots for 37-plex and 48-plex

## 3.1 PCA plots

```{r fig.width=10, fig.height=8}
for (sz in names(results)) {
  res <- results[[sz]]
  pre <- res$pre_res; post <- res$post_res; raw <- res$raw
  batch_cols <- brewer.pal(length(unique(raw$Batch)), "Dark2")
  condition1_cols <- setNames(c("#26547c","#ffd166","#ef476f"),
                              c("non-CF","CFN","CFB"))
  condition2_levels <- levels(factor(raw$Condition2))
  fixed <- c("C","CFN","CFB")
  others <- setdiff(condition2_levels, fixed)
  other_cols <- brewer.pal(max(3, length(others)), name="Dark2")
  condition2_cols <- c(
    C = "#26547c",
    CFN = "#ffd166",
    CFB = "#ef476f",
    setNames(other_cols, others)
  )
  plot_pca <- function(res, title, color_var, color_values) {
    ggplot(res$scores, aes_string("PC1", "PC2", color = color_var)) +
      geom_point(size = 2) +
      scale_color_manual(name = color_var, values = color_values) +
      labs(title = title, x = "PC1", y = "PC2") +
      theme_classic()
  }
  p1 <- plot_pca(pre,  paste0("Pre-ComBat by Batch (", sz, "-plex)"),      "Batch",      batch_cols)
  p2 <- plot_pca(post, paste0("Post-ComBat by Batch (", sz, "-plex)"),     "Batch",      batch_cols)
  p3 <- plot_pca(pre,  paste0("Pre-ComBat by Condition1 (", sz, "-plex)"),"Condition1", condition1_cols)
  p4 <- plot_pca(post, paste0("Post-ComBat by Condition1 (", sz, "-plex)"),"Condition1", condition1_cols)
  print((p1 | p2) / (p3 | p4))
}
```

## 3.2 PCA by Condition2

```{r fig.width=14, fig.height=4}
for (sz in names(results)) {
  res <- results[[sz]]
  pre <- res$pre_res; post <- res$post_res; raw <- res$raw
  condition2_levels <- levels(factor(raw$Condition2))
  fixed <- c("C","CFN","CFB")
  others <- setdiff(condition2_levels, fixed)
  other_cols <- brewer.pal(max(3, length(others)), name="Dark2")
  condition2_cols <- c(
    C = "#26547c",
    CFN = "#ffd166",
    CFB = "#ef476f",
    setNames(other_cols, others)
  )
  plot_pca <- function(res, title, color_var, color_values) {
    ggplot(res$scores, aes_string("PC1", "PC2", color = color_var)) +
      geom_point(size = 2) +
      scale_color_manual(name = color_var, values = color_values) +
      labs(title = title, x = "PC1", y = "PC2") +
      theme_classic()
  }
  p5 <- plot_pca(pre,  paste0("Pre-ComBat by Condition2 (", sz, "-plex)"), "Condition2", condition2_cols)
  p6 <- plot_pca(post, paste0("Post-ComBat by Condition2 (", sz, "-plex)"), "Condition2", condition2_cols)
  print(p5 | p6)
}
```

## 3.3 RLE plots
### 3.3.1 37-plex
```{r fig.height=3, fig.width=6}
sz <- "37"
res <- results[[sz]]
raw <- res$raw
expr_log    <- res$expr_log
expr_combat <- res$expr_combat

# 1) Prepare your matrices
mat1 <- t(expr_log)
mat2 <- t(expr_combat)

# 2) compute per-feature medians
ref1 <- apply(mat1, 1, median)
ref2 <- apply(mat2, 1, median)

# 3) subtract to get RLE
rle1 <- mat1 - ref1
rle2 <- mat2 - ref2

# 4) recompute palette for this raw$Batch
batch_lvls <- sort(unique(raw$Batch))
batch_cols <- setNames(brewer.pal(length(batch_lvls), "Dark2"), batch_lvls)

# 5) Order samples by batch and map to colours
ord        <- order(raw$Batch)
rle1_ord   <- rle1[, ord]
rle2_ord   <- rle2[, ord]
col_vec    <- batch_cols[ raw$Batch[ord] ]

# 6) side-by-side boxplots
par(mfrow = c(1,2), mar = c(4,4,4,1))

boxplot(rle1_ord,
        outline = FALSE,
        col     = col_vec,
        border  = NA,
        xaxt    = "n",
        main    = "RLE before ComBat",
        ylab    = "Deviation from median")

boxplot(rle2_ord,
        outline = FALSE,
        col     = col_vec,
        border  = NA,
        xaxt    = "n",
        main    = "RLE after ComBat",
        ylab    = "Deviation from median")


```

### 3.3.2 48-plex
```{r fig.height=3, fig.width=6}
sz <- "48"
res <- results[[sz]]
raw <- res$raw
expr_log    <- res$expr_log
expr_combat <- res$expr_combat

# 1) Prepare your matrices
mat1 <- t(expr_log)
mat2 <- t(expr_combat)

# 2) compute per-feature medians
ref1 <- apply(mat1, 1, median)
ref2 <- apply(mat2, 1, median)

# 3) subtract to get RLE
rle1 <- mat1 - ref1
rle2 <- mat2 - ref2

# 4) recompute palette for this raw$Batch
batch_lvls <- sort(unique(raw$Batch))
batch_cols <- setNames(brewer.pal(length(batch_lvls), "Dark2"), batch_lvls)

# 5) Order samples by batch and map to colours
ord        <- order(raw$Batch)
rle1_ord   <- rle1[, ord]
rle2_ord   <- rle2[, ord]
col_vec    <- batch_cols[ raw$Batch[ord] ]

# 6) side-by-side boxplots
par(mfrow = c(1,2), mar = c(4,4,4,1))

boxplot(rle1_ord,
        outline = FALSE,
        col     = col_vec,
        border  = NA,
        xaxt    = "n",
        main    = "RLE before ComBat",
        ylab    = "Deviation from median")

boxplot(rle2_ord,
        outline = FALSE,
        col     = col_vec,
        border  = NA,
        xaxt    = "n",
        main    = "RLE after ComBat",
        ylab    = "Deviation from median")


```


## 3.4 37-plex: Deviation from median and Log2 Concentration

```{r fig.width=9, fig.height=12.5}
library(patchwork)
sz <- "37"
res <- results[[sz]]
raw <- res$raw
expr_log <- res$expr_log
expr_combat <- res$expr_combat
cyto_vars <- res$cyto_vars

# Deviation from median
df_pre <- raw %>%
  select(Record_ID, Batch) %>%
  bind_cols(as.data.frame(expr_log)) %>%
  pivot_longer(cols = all_of(cyto_vars), names_to = "Cytokine", values_to = "Expression") %>%
  mutate(ComBat = "pre")
df_post <- raw %>%
  select(Record_ID, Batch) %>%
  bind_cols(as.data.frame(expr_combat)) %>%
  pivot_longer(cols = all_of(cyto_vars), names_to = "Cytokine", values_to = "Expression") %>%
  mutate(ComBat = "post")
df_all <- bind_rows(df_pre, df_post)
cyto_median <- df_all %>%
  group_by(Cytokine) %>% summarize(MedianAll = median(Expression, na.rm=TRUE), .groups="drop")
df_dev <- df_all %>%
  left_join(cyto_median, by="Cytokine") %>%
  mutate(Deviation = Expression - MedianAll,
         ComBat = factor(ComBat, levels=c("pre","post")),
         group = interaction(ComBat, Batch, drop=TRUE))
batch_lvls <- sort(unique(df_dev$Batch))
batch_cols <- brewer.pal(length(batch_lvls), name="Dark2"); names(batch_cols) <- batch_lvls
p_dev <- ggplot(df_dev, aes(x=ComBat, y=Deviation, fill=ComBat, group=group)) +
  geom_boxplot(aes(color=Batch), position=position_dodge(width=0.75),
               alpha=0.6, outlier.size=0.5) +
  scale_fill_manual(name="ComBat step", values=c("pre"="black","post"="white")) +
  scale_color_manual(name="Batch", values=batch_cols) +
  facet_wrap(~ Cytokine, scales="free_y", ncol=3) +
  labs(
    title    = "37-plex: Per‐cytokine deviation from overall median by Batch",
    subtitle = "Left = pre-ComBat, Right = post-ComBat",
    x        = "ComBat step",
    y        = "Expression − median(Expression)"
  ) +
  theme_bw(base_size=12) +
  theme(
    strip.text      = element_text(size=8),
    axis.text.x     = element_text(angle=45, hjust=1),
    legend.position = "bottom"
  )

# Log2 concentration
df_all2 <- bind_rows(df_pre, df_post) %>%
  mutate(ComBat = factor(ComBat, levels=c("pre","post")),
         group = interaction(ComBat, Batch, drop=TRUE))
batch_lvls2 <- sort(unique(df_all2$Batch))
batch_cols2 <- brewer.pal(length(batch_lvls2), name="Dark2"); names(batch_cols2) <- batch_lvls2
p_log2 <- ggplot(df_all2, aes(x=ComBat, y=Expression, fill=ComBat, group=group)) +
  geom_boxplot(position=position_dodge(width=0.75),
               alpha=0.6, outlier.size=0.5) +
  geom_jitter(aes(color=Batch),
              position=position_jitterdodge(jitter.width=0.2, dodge.width=0.75),
              size=1.2, alpha=0.8) +
  scale_fill_manual(name="ComBat step", values=c("pre"="grey80", "post"="white")) +
  scale_color_manual(name="Batch", values=batch_cols2) +
  facet_wrap(~ Cytokine, scales="free_y", ncol=3) +
  labs(
    title    = "37-plex: Log2‐concentration distributions of each cytokine, pre‐ vs post‐ComBat",
    subtitle = "Grey = pre‐ComBat, white = post‐ComBat",
    x        = "ComBat",
    y        = "Log2 Concentration"
  ) +
  theme_bw(base_size=12) +
  theme(
    strip.text      = element_text(size=8),
    axis.text.x     = element_text(angle=45, hjust=1),
    legend.position = "bottom"
  )

# Display both plots vertically
p_dev 
p_log2
```

## 3.5 48-plex: Deviation from median and Log2 Concentration

```{r fig.width=15, fig.height=15}
library(patchwork)
sz <- "48"
res <- results[[sz]]
raw <- res$raw
expr_log <- res$expr_log
expr_combat <- res$expr_combat
cyto_vars <- res$cyto_vars

# Deviation from median
df_pre <- raw %>%
  select(Record_ID, Batch) %>%
  bind_cols(as.data.frame(expr_log)) %>%
  pivot_longer(cols = all_of(cyto_vars), names_to = "Cytokine", values_to = "Expression") %>%
  mutate(ComBat = "pre")
df_post <- raw %>%
  select(Record_ID, Batch) %>%
  bind_cols(as.data.frame(expr_combat)) %>%
  pivot_longer(cols = all_of(cyto_vars), names_to = "Cytokine", values_to = "Expression") %>%
  mutate(ComBat = "post")
df_all <- bind_rows(df_pre, df_post)
cyto_median <- df_all %>%
  group_by(Cytokine) %>% summarize(MedianAll = median(Expression, na.rm=TRUE), .groups="drop")
df_dev <- df_all %>%
  left_join(cyto_median, by="Cytokine") %>%
  mutate(Deviation = Expression - MedianAll,
         ComBat = factor(ComBat, levels=c("pre","post")),
         group = interaction(ComBat, Batch, drop=TRUE))
batch_lvls <- sort(unique(df_dev$Batch))
batch_cols <- brewer.pal(length(batch_lvls), name="Dark2"); names(batch_cols) <- batch_lvls
p_dev <- ggplot(df_dev, aes(x=ComBat, y=Deviation, fill=ComBat, group=group)) +
  geom_boxplot(aes(color=Batch), position=position_dodge(width=0.75),
               alpha=0.6, outlier.size=0.5) +
  scale_fill_manual(name="ComBat step", values=c("pre"="black","post"="white")) +
  scale_color_manual(name="Batch", values=batch_cols) +
  facet_wrap(~ Cytokine, scales="free_y", ncol=6) +
  labs(
    title    = "48-plex: Per‐cytokine deviation from overall median by Batch",
    subtitle = "Left = pre-ComBat, Right = post-ComBat",
    x        = "ComBat step",
    y        = "Expression − median(Expression)"
  ) +
  theme_bw(base_size=12) +
  theme(
    strip.text      = element_text(size=8),
    axis.text.x     = element_text(angle=45, hjust=1),
    legend.position = "bottom"
  )

# Log2 concentration
df_all2 <- bind_rows(df_pre, df_post) %>%
  mutate(ComBat = factor(ComBat, levels=c("pre","post")),
         group = interaction(ComBat, Batch, drop=TRUE))
batch_lvls2 <- sort(unique(df_all2$Batch))
batch_cols2 <- brewer.pal(length(batch_lvls2), name="Dark2"); names(batch_cols2) <- batch_lvls2
p_log2 <- ggplot(df_all2, aes(x=ComBat, y=Expression, fill=ComBat, group=group)) +
  geom_boxplot(position=position_dodge(width=0.75),
               alpha=0.6, outlier.size=0.5) +
  geom_jitter(aes(color=Batch),
              position=position_jitterdodge(jitter.width=0.2, dodge.width=0.75),
              size=1.2, alpha=0.8) +
  scale_fill_manual(name="ComBat step", values=c("pre"="grey80", "post"="white")) +
  scale_color_manual(name="Batch", values=batch_cols2) +
  facet_wrap(~ Cytokine, scales="free_y", ncol=6) +
  labs(
    title    = "48-plex: Log2‐concentration distributions of each cytokine, pre‐ vs post‐ComBat",
    subtitle = "Grey = pre‐ComBat, white = post‐ComBat",
    x        = "ComBat",
    y        = "Log2 Concentration"
  ) +
  theme_bw(base_size=12) +
  theme(
    strip.text      = element_text(size=8),
    axis.text.x     = element_text(angle=45, hjust=1),
    legend.position = "bottom"
  )

# Display both plots vertically
p_dev
p_log2
```

## 3.6 Heatmap of p-values

```{r fig.width=10, fig.height=5}
for (sz in names(results)) {
  res    <- results[[sz]]
  pre    <- res$pre_res; post <- res$post_res
  t_pre  <- paste0("Pre-ComBat: metadata vs PCs (", sz, "-plex)")
  t_post <- paste0("Post-ComBat: metadata vs PCs (", sz, "-plex)")
  brpal <- brewer.pal(5, "RdBu")
  col_fun <- colorRamp2(c(0,0.01,0.05,0.2,1), brpal)
  fmt_sci2 <- function(x) sprintf("%.2e", x)
  ht_pre <- Heatmap(pre$pvals, name="p-value", col=col_fun,
                    cluster_rows=FALSE, cluster_columns=FALSE,
                    column_title=t_pre, row_title="PCs",
                    cell_fun=function(j,i,x,y,w,h,fill) {
                      grid.rect(x,y,w*0.9,h*0.9, gp=gpar(fill=fill,col=NA))
                      grid.text(fmt_sci2(pre$pvals[i,j]), x,y,
                                gp=gpar(fontsize=7))
                    })
  ht_post <- Heatmap(post$pvals, name="p-value", col=col_fun,
                     cluster_rows=FALSE, cluster_columns=FALSE,
                     column_title=t_post, row_title="PCs",
                     cell_fun=function(j,i,x,y,w,h,fill) {
                       grid.rect(x,y,w*0.9,h*0.9, gp=gpar(fill=fill,col=NA))
                       grid.text(fmt_sci2(post$pvals[i,j]), x,y,
                                 gp=gpar(fontsize=7))
                     })
  draw(ht_pre + ht_post, heatmap_legend_side="right")
}
```



# 4 Analysis
```{r}
res.final <- qread(here("data","cytokine","res_ETI.master.qs"))
```

## 4.1 Subset CF and true controls
```{r}
res37 <- results[["37"]]
res48 <- results[["48"]]

# 1) Shared IDs & allowed conditions
common_ids  <- intersect(res37$raw$Record_ID, res48$raw$Record_ID)
keep_conds  <- c("CFB","CFN","C")
#keep_treatment <- c("None","NaN")
```

## 4.2 Remove overlapping cytokines
```{r}
# 2) Cytokines in both → drop from res37 only
common_cyto <- intersect(res37$cyto_vars, res48$cyto_vars)

# 3) One‐liner helper
filter_cyto_df <- function(res, drop_overlap=FALSE) {
  # a) filter rows
  raw_f <- res$raw %>%
    filter(Record_ID %in% common_ids,
           Condition2 %in% keep_conds,
           #CF_treatment %in% keep_treatment,
           Age < 8)

  # b) pick which cyto to keep
  cyto_f <- if (drop_overlap) setdiff(res$cyto_vars, common_cyto) else res$cyto_vars
  # c) meta = all non‐cytokine cols
  meta_vars <- setdiff(names(raw_f), res$cyto_vars)
  meta_f    <- raw_f %>% select(all_of(meta_vars))
  # d) get original row‐indices for subsetting expr_mats
  idx <- match(raw_f$Record_ID, res$raw$Record_ID)
  expr_log_f    <- if (!is.null(res$expr_log))    res$expr_log[idx, cyto_f, drop=FALSE]
  expr_combat_f <- if (!is.null(res$expr_combat)) res$expr_combat[idx, cyto_f, drop=FALSE]
  # e) return
  list(
    raw         = raw_f,
    meta        = meta_f,
    cyto_vars   = cyto_f,
    expr_log    = expr_log_f,
    expr_combat = expr_combat_f
  )
}

# 4) Build two filtered objects
res37.filter <- filter_cyto_df(res37, drop_overlap=TRUE)
res48.filter <- filter_cyto_df(res48, drop_overlap=FALSE)
```

## 4.3 Cross-check metadata columns
```{r}
# 1. figure out which meta‐columns to compare
ignore_patterns <- "^Batch$|^Date_performed$|^What_plate|^Complete\\?$|^Analysis_performed$"
common_meta   <- intersect(names(res37.filter$meta), names(res48.filter$meta))
compare_cols  <- setdiff(common_meta, grep(ignore_patterns, common_meta, value = TRUE))

# 2. pull the joined table with suffixes
cmp <- inner_join(
  res37.filter$meta %>% select(Record_ID, all_of(compare_cols)),
  res48.filter$meta %>% select(Record_ID, all_of(compare_cols)),
  by = "Record_ID",
  suffix = c(".37", ".48")
)

# 3. for each column, find mismatches
mismatches <- map_dfr(compare_cols, function(col) {
  col37 <- paste0(col, ".37")
  col48 <- paste0(col, ".48")
  diffs <- which(cmp[[col37]] != cmp[[col48]])
  if (length(diffs)==0) return(NULL)
  tibble(
    Record_ID = cmp$Record_ID[diffs],
    column    = col,
    val.37    = cmp[[col37]][diffs],
    val.48    = cmp[[col48]][diffs]
  )
})

# 4. report
if (nrow(mismatches)==0) {
  message("All compared meta‐columns match exactly!")
} else {
  message("Found mismatches:")
  print(mismatches)
}

```

## 4.4 Merge data
```{r}
# 2) Desired metadata columns
keep_meta <- c(
  "Record_ID","Age","Sex",
  "Condition","Condition1","Condition2","CF_Severity","CF_treatment",
  "Collection_date","Control_Subgroup",
  "Infection_bacteria","Bacteria",
  "Infection_virus","Virus",
  "Infection_fungi","Fungi",
  "Batch"
)

# 1) put Record_ID on the rownames of each raw‐matrix
rownames(res37.filter$raw) <- res37.filter$meta$Record_ID
rownames(res48.filter$raw) <- res48.filter$meta$Record_ID

# 2) pick off just the analyte columns (drop the meta ones)
analytes37 <- res37.filter$cyto_vars
analytes48 <- res48.filter$cyto_vars

# 3) find the samples in common
common <- intersect(
  rownames(res37.filter$raw),
  rownames(res48.filter$raw)
)

# 4) cbind only the analyte cols for those common samples
res.final.raw <- cbind(
  res37.filter$raw[common, analytes37, drop=FALSE],
  res48.filter$raw[common, analytes48, drop=FALSE]
)

# 5) rebuild meta frame to the same order
res.final.meta <- res37.filter$meta[ match(common, res37.filter$meta$Record_ID), keep_meta, drop = FALSE ]

# 4) Record_ID rownames on each expr matrix
rownames(res37.filter$expr_log)    <- res37.filter$meta$Record_ID
rownames(res48.filter$expr_log)    <- res48.filter$meta$Record_ID
rownames(res37.filter$expr_combat) <- res37.filter$meta$Record_ID
rownames(res48.filter$expr_combat) <- res48.filter$meta$Record_ID

# 5) Find the common 136 samples (same order for both plexes)
common <- intersect(
  rownames(res37.filter$expr_log),
  rownames(res48.filter$expr_log)
)

# 6) Subset & cbind the log and ComBat matrices
res.final.expr_log <- cbind(
  res37.filter$expr_log[common, , drop=FALSE],
  res48.filter$expr_log[common, , drop=FALSE]
)

res.final.expr_combat <- cbind(
  res37.filter$expr_combat[common, , drop=FALSE],
  res48.filter$expr_combat[common, , drop=FALSE]
)

# 8) Recompute cytokine list
res.final.cyto_vars <- colnames(res.final.expr_combat)

# 9) Final sanity‐checks
dim(res.final.raw)    # 136 × 62
dim(res.final.expr_log)    # 136 × 49
dim(res.final.expr_combat) # 136 × 49
dim(res.final.meta)        # 136 × 16

# 10) Assemble final list
res.final <- list(
  raw         = res.final.raw,
  meta        = res.final.meta,
  cyto_vars   = res.final.cyto_vars,
  expr_log    = res.final.expr_log,
  expr_combat = res.final.expr_combat
)

# ensure character
res.final$meta$Record_ID <- as.character(res.final$meta$Record_ID)

# create Subject_ID (first 6 chars if Record_ID starts with "M", else whole Record_ID)
res.final$meta$Subject_ID <- ifelse(startsWith(res.final$meta$Record_ID, "M"),
                                    substr(res.final$meta$Record_ID, 1, 6),
                                    res.final$meta$Record_ID)

# move Subject_ID immediately after Record_ID
res.final$meta <- res.final$meta[, c("Record_ID", "Subject_ID", setdiff(names(res.final$meta), c("Record_ID", "Subject_ID")))]

# Inspect
#str(res.final)

# backup object
res.final.backup <- res.final

```

## 4.5 Filter ETI-related samples
```{r}
records <- c("M1C160(1)","M1C160F", # CF-S1, treated
            "M1C177","M1C177C", # CF-M3, treated
            "M1C188","M1C188B", # CF-S3, treated
            "M1C170C","M1C170D", # CF-M1, untreated
            "M1C176", "M1C176C", # CF-M2
            "M1C180","M1C180D", # CF-M4
            "M1N066","M1N075","M1N078","M1N080","M1N087","M1N092" #HC1-6
            )
if (!exists("res.final.backup")) stop("res.final.backup not found in the environment")

# preserve the order given in `records`, but only keep those that actually exist
available_ids <- rownames(res.final.backup$raw)
ordered_found <- records[records %in% available_ids]

if (length(ordered_found) == 0) stop("None of the requested record IDs were found in res.final.backup$raw")

missing_ids <- setdiff(records, ordered_found)
if (length(missing_ids) > 0) warning("Some requested Record_IDs were not found and will be skipped: ",
                                    paste(missing_ids, collapse = ", "))

# build res.final by subsetting known components; be tolerant if some components missing
res.final <- list()

if ("raw" %in% names(res.final.backup)) {
  res.final$raw <- res.final.backup$raw[ordered_found, , drop = FALSE]
  # keep rownames as Record_ID
}

if ("meta" %in% names(res.final.backup)) {
  # meta is a data.frame with Record_ID column
  md <- res.final.backup$meta
  # subset and preserve order
  res.final$meta <- md[match(ordered_found, md$Record_ID), , drop = FALSE]
  # ensure Record_ID is character
  if ("Record_ID" %in% names(res.final$meta)) res.final$meta$Record_ID <- as.character(res.final$meta$Record_ID)
}

if ("cyto_vars" %in% names(res.final.backup)) {
  res.final$cyto_vars <- res.final.backup$cyto_vars
}

if ("expr_log" %in% names(res.final.backup)) {
  # expr_log rows should be Record_ID rownames
  res.final$expr_log <- res.final.backup$expr_log[ordered_found, , drop = FALSE]
}

if ("expr_combat" %in% names(res.final.backup)) {
  res.final$expr_combat <- res.final.backup$expr_combat[ordered_found, , drop = FALSE]
}

# simple sanity checks
message("Constructed res.final with ", nrow(res.final$meta), " samples (rows).")
if (!is.null(res.final$raw))  message("res.final$raw dims: ", paste(dim(res.final$raw), collapse = " x "))
if (!is.null(res.final$expr_log)) message("res.final$expr_log dims: ", paste(dim(res.final$expr_log), collapse = " x "))
if (!is.null(res.final$expr_combat)) message("res.final$expr_combat dims: ", paste(dim(res.final$expr_combat), collapse = " x "))

# add_treatment_to_res_final_meta.R
pre_ids <- c("M1C160(1)", "M1C177", "M1C188")
ut1_ids <- c("M1C170C", "M1C176", "M1C180")
ut2_ids <- c("M1C170D", "M1C176C", "M1C180D")

res.final$meta <- res.final$meta %>%
  dplyr::mutate(
    treatment = dplyr::case_when(
      CF_treatment == "Trikafta"              ~ "ETI",
      CF_treatment == "NaN"                   ~ "C",
      CF_treatment == "None" & Record_ID %in% pre_ids  ~ "pre-ETI",
      CF_treatment == "None" & Record_ID %in% ut1_ids ~ "UT1",
      CF_treatment == "None" & Record_ID %in% ut2_ids ~ "UT2",
      TRUE                                    ~ NA_character_
    )) %>%
  dplyr::mutate(
    treatment_group = dplyr::case_when(
      treatment %in% c("pre-ETI", "ETI") ~ "ETI",
      treatment %in% c("UT1", "UT2")     ~ "UT",
      treatment == "C"                   ~ "C"
    )
  )

# sanity check
table(res.final$meta$treatment, useNA = "ifany")
# sanity check
table(res.final$meta$treatment_group, useNA = "ifany")

# save object
qsave(res.final, file=here("data","cytokine","res_ETI.master.qs"))
```

## 4.6 PCA
### 4.6.1 fibro cytokines
```{r fig.height=4, fig.width=9}
# 1) define your fibro‐related cytokines
fibro.cytokines <- c(
  "BAFF/TNFSF13B", "sCD30/TNFRSF8", "sCD163", "Chitinase-3-like1",
  "gp130/sIL-6Rβ", "sIL-6Rα", "Osteopontin", "Pentraxin-3", "TSLP",
  "TWEAK/TNFSF12", "FGFbasic", "Eotaxin", "IL-1β", "IL-5",
  "IL-8", "IL-13", "MCP-1(MCAF)", "SDF-1α", "PDGF-BB","M-CSF"
)

# 2) subset the ComBat‐corrected data
expr_fibro <- res.final$expr_combat[, fibro.cytokines, drop = FALSE]

# 1) PCA on the merged ComBat‐corrected matrix
#    rows = samples, cols = cytokines
pca_final <- prcomp(expr_fibro, center = TRUE, scale. = TRUE)
# Percentage variance explained
var_exp <- summary(pca_final)$importance[2, 1:2] * 100

# 2) turn scores into a data.frame and join metadata
scores_final <- as.data.frame(pca_final$x[, 1:10], check.names = FALSE)
scores_final <- bind_cols(scores_final, res.final$meta)

# define stroke (border) colours and fill colours per treatment
stroke_cols <- c(
  "pre-ETI" = "#3a5a40",
  "ETI"     = "#3a5a40",
  "UT1"     = "#936639",
  "UT2"     = "#936639",
  "C"       = "#26547c"
)
fill_cols <- c(
  "pre-ETI" = "white",      # empty-looking (white interior) with coloured border
  "ETI"     = "#3a5a40",    # filled with ETI colour
  "UT1"     = "white",      # empty-looking
  "UT2"     = "#936639",    # filled with UT2 colour
  "C"       = "#26547c"     # control (cross) — fill ignored for non-fillable shape
)

# shapes: C = cross (4), T = upward triangle (24, fillable), UT = circle (21, fillable)
shape_map <- c("C" = 4, "ETI" = 24, "UT" = 21)

# ensure factors have the expected levels (optional, but keeps legend order predictable)
scores_final$treatment_group <- factor(scores_final$treatment_group, levels = names(shape_map))
scores_final$treatment <- factor(scores_final$treatment, levels = names(stroke_cols))

# prepare small df of subjects with >1 sample (no date ordering)
lines_df <- scores_final %>%
  group_by(Subject_ID) %>%
  filter(n() > 1) %>%
  arrange(Subject_ID, PC1) %>%   # deterministic order; change to Collection_date if you later want date ordering
  ungroup()

# ensure Record_ID is a factor for consistent colour mapping
scores_final$Record_ID <- as.factor(scores_final$Record_ID)

# centroids_sample: label coordinates taken directly from scores_final
centroids_sample <- scores_final %>%
  transmute(x = PC1, y = PC2, sampleID = Record_ID)

p_sample <- ggplot(scores_final, aes(x = PC1, y = PC2, color = Record_ID)) +
  geom_point(size = 4) +
  ggrepel::geom_text_repel(
    data = centroids_sample,
    aes(x = x, y = y, label = sampleID),
    inherit.aes = FALSE,
    size = 3.2,
    box.padding = 0.3,
    point.padding = 0.3,
    segment.size = 0.3,
    max.overlaps = Inf
  ) +
  scale_color_manual(values = setNames(rep("#666666", length(levels(scores_final$Record_ID))),
                                      levels(scores_final$Record_ID))) +
  labs(title = "Record_ID") +
  NoLegend() +
  theme_classic(base_size = 12) +
  theme(
    axis.text        = element_text(size = 12),
    axis.title       = element_text(size = 14),
    panel.border     = element_rect(color = "black", fill = NA, size = 0.7),
    axis.line        = element_blank(),
    legend.position  = "none",
    plot.title       = element_text(hjust = 0.5)
  )

# plot
ggplot(scores_final, aes(x = PC1, y = PC2,
                         shape = treatment_group,
                         color = treatment,
                         fill = treatment)) +
  # draw lines connecting longitudinal samples (behind points)
  geom_line(data = lines_df,
            aes(x = PC1, y = PC2, group = Subject_ID),
            inherit.aes = FALSE,
            color = "grey20",
            linetype = "solid",
            size = 0.7,
            alpha = 0.8) +
  # points on top
  geom_point(size = 3.5, stroke = 0.9, alpha = 0.95) +
  stat_ellipse(aes(group = treatment), type = "norm", level = 0.68,
               geom = "polygon", alpha = 0.18, colour = NA, show.legend = FALSE) +
  # rename shape legend to "Treatment"
  scale_shape_manual(name = "Treatment", values = shape_map) +
  # hide color & fill legends (only keep the shape legend)
  scale_color_manual(values = stroke_cols, guide = "none") +
  scale_fill_manual(values = fill_cols, guide = "none") +
  # optionally tweak how the shape legend keys appear
  guides(shape = guide_legend(override.aes = list(colour = "black", fill = NA, size = 4))) +
  labs(
    title = "Profibrotic cytokines",
    x     = paste0("PC1 (", round(var_exp[1], 1), "%)"),
    y     = paste0("PC2 (", round(var_exp[2], 1), "%)")
  ) +
  theme_classic(base_size = 12) +
  theme(
    axis.text        = element_text(size = 12),
    axis.title       = element_text(size = 14),
    panel.border     = element_rect(color = "black", fill = NA, size = 0.7),
    axis.line        = element_blank(),
    legend.position  = "right",
    legend.direction = "vertical",
    plot.title       = element_text(hjust = 0.5)
  ) -> p2
p_sample | p2
```

### 4.6.2 all cytokines
```{r fig.height=4, fig.width=9}
# 1) PCA on the merged ComBat‐corrected matrix
#    rows = samples, cols = cytokines
pca_final <- prcomp(res.final$expr_combat, center = TRUE, scale. = TRUE)
# Percentage variance explained
var_exp <- summary(pca_final)$importance[2, 1:2] * 100

# 2) turn scores into a data.frame and join metadata
scores_final <- as.data.frame(pca_final$x[, 1:10], check.names = FALSE)
scores_final <- bind_cols(scores_final, res.final$meta)

# define stroke (border) colours and fill colours per treatment
stroke_cols <- c(
  "pre-ETI" = "#3a5a40",
  "ETI"     = "#3a5a40",
  "UT1"     = "#936639",
  "UT2"     = "#936639",
  "C"       = "#26547c"
)
fill_cols <- c(
  "pre-ETI" = "white",      # empty-looking (white interior) with coloured border
  "ETI"     = "#3a5a40",    # filled with ETI colour
  "UT1"     = "white",      # empty-looking
  "UT2"     = "#936639",    # filled with UT2 colour
  "C"       = "#26547c"     # control (cross) — fill ignored for non-fillable shape
)

# shapes: C = cross (4), T = upward triangle (24, fillable), UT = circle (21, fillable)
shape_map <- c("C" = 4, "ETI" = 24, "UT" = 21)

# ensure factors have the expected levels (optional, but keeps legend order predictable)
scores_final$treatment_group <- factor(scores_final$treatment_group, levels = names(shape_map))
scores_final$treatment <- factor(scores_final$treatment, levels = names(stroke_cols))

# prepare small df of subjects with >1 sample (no date ordering)
lines_df <- scores_final %>%
  group_by(Subject_ID) %>%
  filter(n() > 1) %>%
  arrange(Subject_ID, PC1) %>%   # deterministic order; change to Collection_date if you later want date ordering
  ungroup()

# ensure Record_ID is a factor for consistent colour mapping
scores_final$Record_ID <- as.factor(scores_final$Record_ID)

# centroids_sample: label coordinates taken directly from scores_final
centroids_sample <- scores_final %>%
  transmute(x = PC1, y = PC2, sampleID = Record_ID)

p_sample <- ggplot(scores_final, aes(x = PC1, y = PC2, color = Record_ID)) +
  geom_point(size = 4) +
  ggrepel::geom_text_repel(
    data = centroids_sample,
    aes(x = x, y = y, label = sampleID),
    inherit.aes = FALSE,
    size = 3.2,
    box.padding = 0.3,
    point.padding = 0.3,
    segment.size = 0.3,
    max.overlaps = Inf
  ) +
  scale_color_manual(values = setNames(rep("#666666", length(levels(scores_final$Record_ID))),
                                      levels(scores_final$Record_ID))) +
  labs(title = "Record_ID") +
  NoLegend() +
  theme_classic(base_size = 12) +
  theme(
    axis.text        = element_text(size = 12),
    axis.title       = element_text(size = 14),
    panel.border     = element_rect(color = "black", fill = NA, size = 0.7),
    axis.line        = element_blank(),
    legend.position  = "none",
    plot.title       = element_text(hjust = 0.5)
  )

# plot
ggplot(scores_final, aes(x = PC1, y = PC2,
                         shape = treatment_group,
                         color = treatment,
                         fill = treatment)) +
  # draw lines connecting longitudinal samples (behind points)
  geom_line(data = lines_df,
            aes(x = PC1, y = PC2, group = Subject_ID),
            inherit.aes = FALSE,
            color = "grey20",
            linetype = "solid",
            size = 0.7,
            alpha = 0.8) +
  # points on top
  geom_point(size = 3.5, stroke = 0.9, alpha = 0.95) +
  stat_ellipse(aes(group = treatment), type = "norm", level = 0.68,
               geom = "polygon", alpha = 0.18, colour = NA, show.legend = FALSE) +
  # rename shape legend to "Treatment"
  scale_shape_manual(name = "Treatment", values = shape_map) +
  # hide color & fill legends (only keep the shape legend)
  scale_color_manual(values = stroke_cols, guide = "none") +
  scale_fill_manual(values = fill_cols, guide = "none") +
  # optionally tweak how the shape legend keys appear
  guides(shape = guide_legend(override.aes = list(colour = "black", fill = NA, size = 4))) +
  labs(
    title = "Profibrotic cytokines",
    x     = paste0("PC1 (", round(var_exp[1], 1), "%)"),
    y     = paste0("PC2 (", round(var_exp[2], 1), "%)")
  ) +
  theme_classic(base_size = 12) +
  theme(
    axis.text        = element_text(size = 12),
    axis.title       = element_text(size = 14),
    panel.border     = element_rect(color = "black", fill = NA, size = 0.7),
    axis.line        = element_blank(),
    legend.position  = "right",
    legend.direction = "vertical",
    plot.title       = element_text(hjust = 0.5)
  ) -> p2
p_sample | p2
```



## 4.7 Cytokine concentration
```{r fig.width=6, fig.height=3}
# 0) quick checks
stopifnot(exists("res.final"))
stopifnot(is.data.frame(res.final$meta))
stopifnot(is.matrix(res.final$expr_combat) || is.data.frame(res.final$expr_combat))

# 1) build a long dataframe: Record_ID, Subject_ID, treatment, cytokine, Value
expr_df <- as.data.frame(res.final$expr_combat) %>%
  tibble::rownames_to_column("Record_ID")

meta_df <- as.data.frame(res.final$meta) %>%
  select(Record_ID, Subject_ID, treatment)

df <- expr_df %>%
  left_join(meta_df, by = "Record_ID")

# 2) pivot to long; expr_combat is already log2-transformed (from your pipeline), so do not re-transform
cytokines <- setdiff(names(df), c("Record_ID", "Subject_ID", "treatment"))

df_long <- df %>%
  pivot_longer(cols = all_of(cytokines),
               names_to  = "Cytokine",
               values_to = "Value") %>%
  filter(treatment %in% c("pre-ETI", "ETI", "UT1", "UT2")) %>%
  filter(!is.na(Value)) %>%
  mutate(
    Treatment = treatment,           # use a clearer column name for compatibility with previous code
    Subject_ID = as.character(Subject_ID)
  )

# 3) Per-cytokine, per-subject counts of observations (we want paired samples)
pair_counts <- df_long %>%
  group_by(Cytokine, Subject_ID) %>%
  summarise(n = n(), .groups = "drop")

# 4) Keep cytokines where all subjects present in this subset have exactly 2 observations
#    (i.e. both timepoints for their pairing)
good_cytokines <- pair_counts %>%
  group_by(Cytokine) %>%
  summarise(all_paired = all(n == 2), .groups = "drop") %>%
  filter(all_paired) %>%
  pull(Cytokine)

message("Kept ", length(good_cytokines), " cytokines with complete pairs across subjects.")

df_long2 <- df_long %>%
  filter(Cytokine %in% good_cytokines) %>%
  # order factors so UT1 -> UT2 and pre-ETI -> ETI in the x-axis
  mutate(Treatment = factor(Treatment, levels = c("UT1", "UT2", "pre-ETI", "ETI")),
         PairPanel = ifelse(grepl("^UT", as.character(Treatment)), "UT", "ETI"))

# 5) paired t-tests per cytokine for the two comparisons of interest
comparisons <- list(
  c("ETI",    "pre-ETI"),
  c("UT2",    "UT1")
)

stat.tests <- df_long2 %>%
  group_by(Cytokine) %>%
  pairwise_t_test(
    Value ~ Treatment,
    paired           = TRUE,
    paired.id        = "Subject_ID",
    comparisons      = comparisons,
    p.adjust.method  = "BH"
  ) %>%
  ungroup()

# 6) select significant results (BH-adjusted p < 0.05)
sig_tests <- stat.tests %>% filter(p.adj < 0.05)
sig_cytokines <- unique(sig_tests$Cytokine)
message("Significant cytokines (FDR < 0.05): ", paste(sig_cytokines, collapse = ", "))

# 7) subset data to only the significant cytokines (or a chosen set)
if (length(sig_cytokines) == 0) {
  warning("No significant cytokines found. Plotting top 6 by smallest p.adj instead.")
  top_cytos <- stat.tests %>% arrange(p.adj) %>% pull(Cytokine) %>% unique() %>% head(6)
  plot_cytokines <- top_cytos
} else {
  plot_cytokines <- sig_cytokines
}

df_sig <- df_long2 %>% filter(Cytokine %in% plot_cytokines)
df_sig$Cytokine <- factor(df_sig$Cytokine, levels = plot_cytokines)

# 8) prepare stat.test for plotting (filter to those cytokines and non-NA p)
plot_stat <- stat.tests %>%
  filter(Cytokine %in% plot_cytokines) %>%
  filter(!is.na(p.adj))

# 9) compute y positions (per-cytokine) to draw brackets above the data
lims <- df_sig %>%
  group_by(Cytokine) %>%
  summarise(
    y_max = max(Value, na.rm = TRUE),
    y_min = min(Value, na.rm = TRUE),
    .groups = "drop"
  )

plot_stat <- plot_stat %>%
  left_join(lims, by = "Cytokine") %>%
  group_by(Cytokine) %>%
  arrange(p.adj) %>%   # smaller p on top (optional)
  mutate(
    y.position = y_max + (row_number() - 1) * (y_max - y_min) * 0.15
  ) %>%
  ungroup()

# 10) plotting palette for treatments (reuse your border colours)
treatment_pal <- c("pre-ETI" = "#3a5a40", "ETI" = "#3a5a40", "UT1" = "#936639", "UT2" = "#936639")

p <- ggplot(df_sig, aes(x = Treatment, y = Value)) +

  # draw lines at the *true* x positions (as numeric), so they connect in the intended order
  geom_line(
    data = df_sig,
    aes(x = as.numeric(Treatment), y = Value, group = interaction(Subject_ID, PairPanel)),
    inherit.aes = FALSE,linetype="solid",
    color = "grey70", size = 0.5
  ) +

  # draw jittered points on top (so overlapping points become visible)
  geom_point(
    aes(color = Treatment),
    position = position_jitterdodge(jitter.width = 0.5),
    size = 2, alpha = 0.95
  ) +
  # allow brackets/annotations outside panel
  { if (nrow(plot_stat) > 0) stat_pvalue_manual(plot_stat,
                                               label = "p.adj.signif",
                                               tip.length = 0.02,bracket.nudge.y = 1,
                                               bracket.size = 0.4,
                                               hide.ns = TRUE)
    else NULL } +
  theme_classic(base_size = 12) +
  theme(
    strip.text       = element_text(size = 8),
    axis.title.x     = element_blank(),
    axis.text.x      = element_text(angle = 45, hjust = 1),
    panel.border     = element_rect(color = "black", fill = NA, size = 0.5),
    strip.background = element_rect(color = "black", fill = "white", size = 0.5),
    axis.line.x      = element_blank(),
    axis.line.y      = element_blank(),
    legend.position  = "right"
  ) +

  # restore axis ticks/labels from numeric x used by lines
  scale_x_continuous(breaks = 1:4, labels = c("UT1","UT2","pre-ETI","ETI")) +
  scale_color_manual(values = treatment_pal) +
  facet_wrap(~ Cytokine, scales = "free_y", ncol = 3) +
  scale_x_discrete(limits = c("UT1", "UT2", "pre-ETI", "ETI")) +
  scale_color_manual(values = treatment_pal) +
  scale_y_continuous(expand = expansion(mult = c(0.12, 0.35))) +
  coord_cartesian(clip = "off") +   # keep brackets/points visible
  theme_classic(base_size = 12) +
  theme(
    strip.text = element_text(size = 8),
    axis.title.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.border = element_rect(color = "black", fill = NA, size = 0.5),
    strip.background = element_rect(color = "black", fill = "white", size = 0.5)
  ) +
  labs(y = "Concentration (log2 pg/mL)", color = "Treatment")

print(p)

```

# 5 Save raw data
```{r}
write_two_tables_minimal <- function(res.final, file = "res_final_both.xlsx") {
  raw <- as.data.frame(res.final$raw, stringsAsFactors = FALSE)
  expr <- as.data.frame(res.final$expr_combat, stringsAsFactors = FALSE)
  meta <- as.data.frame(res.final$meta, stringsAsFactors = FALSE)

  ids <- rownames(raw)
  cond2_map <- setNames(as.character(meta$Condition2), as.character(meta$Record_ID))
  cond2_col <- cond2_map[ids]

  expr <- expr[ids, , drop = FALSE]

  left_tbl <- cbind(Record_ID = ids, Condition2 = cond2_col, raw, stringsAsFactors = FALSE)
  rownames(left_tbl) <- NULL
  rownames(expr) <- NULL

  # R does not allow a zero-length column name (attempting `` = ... fails).
  # Use a single space " " as a blank-looking header instead.
  empty_col <- data.frame(blank = rep(NA, nrow(left_tbl)), stringsAsFactors = FALSE)
  names(empty_col) <- " "

  combined <- cbind(left_tbl, empty_col, expr, stringsAsFactors = FALSE)

  writexl::write_xlsx(list(Sheet1 = combined), path = file)
  invisible(file)
}
write_two_tables_minimal(res.final, here("data","cytokine","res_final_ETI.xlsx"))
```

# 6 Save figures
```{r}
# out <- here("data","plots","cytokine")
# if(!dir.exists(out)) {
#   dir.create(out,recursive = TRUE)
# }
# 
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 1600)
# 
# # pca
# p1 +
#   theme(plot.tag = element_text(size = 14, face = "bold"))
# 
# ggsave(plot=last_plot(),
#        device="jpeg",
#        dpi=1600,
#        filename = paste0(out,"/","ETI_PCA_all.jpeg"),
#        height=4,
#        width=5)
# 
# ggsave(plot=p_sample + p2 +
#   theme(plot.tag = element_text(size = 14, face = "bold")),
#        device="jpeg",
#        dpi=1600,
#        filename = paste0(out,"/","ETI_PCA_fibro.jpeg"),
#        height=4,
#        width=9)
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 100)
# 
# # combined cytokine
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 1600)
# 
# ggsave(plot=p_sample + p2 +
#   theme(plot.tag = element_text(size = 14, face = "bold", family="Arial")),
#        device="jpeg",
#        dpi=1600,
#        filename = here("data","plots","ETI","bal_ETI.combined.cytoLvl.MDSplot.jpeg"),
#        height=4,
#        width=9)
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 100)
# 
# # profibrotic cytokine
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 1600)
# 
# ggsave(plot=p_sample + p2 +
#   theme(plot.tag = element_text(size = 14, face = "bold", family="Arial")),
#        device="jpeg",
#        dpi=1600,
#        filename = here("data","plots","ETI","bal_ETI.combined.cytoLvl.fibro.MDSplot.jpeg"),
#        height=4,
#        width=9)
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 100)
```

# 7 Save tables
```{r}
scores_final %>% write_csv(here("output","Fig6B_cytokine.csv"))

```